---
title: "Moderation analysis using the PROCESS macro"
author: "Mauricio Garnier-Villarreal, Joris M. Schröder & Joseph Charles Van Matre"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 5
editor_options:
  chunk_output_type: inline
---

# What is moderation analysis?

With moderation analysis, we are trying to find out whether the effect or association between two variables depends on another variable. Let's say that we are interested in the association between *Lack of confidence in the government* and *Perception of corruption*, but we specifically want to know whether the association depends on the *Sex* or the *Secular Values* of individuals. The latter two variables are called *moderators* of the association between *Lack of confidence in the government* and *Perception of corruption*.

# Setup the R session

When we start working in R, we always need to setup our session. For this we need to set our working directory, in this case I am doing that for the folder that holds the downloaded [World Values Survey (WVS)](https://www.worldvaluessurvey.org/) ```SPSS``` data set

```{r, eval=F, error=F}
setwd("~path_to_your_file")
```

The next step for setting up our session will be to load the packages that we will be using

```{r, eval=T, message=F}
library(rio)
```


# Installing the PROCESS macro

In this tutorial, we will be showing you how to do moderation analysis using the **PROCESS** macro. To use the process macro, you first need to download it from this website:  [https://www.processmacro.org/download.html](https://www.processmacro.org/download.html). You will download a zip archive that you need to extract (using for example [7-Zip](https://www.7-zip.org/)). The extracted folder contains a folder *PROCESS v4.1 for R* (the latest version by the time we created this tutorial), which contains the *process.R* script. 
The process macro is a script rather than an r package, and its usage is therefore a bit different from what you are used to. Andrew Hayes, the author of the PROCESS macro writes:

*"PROCESS for R is a program file or “script” (process.R) that when executed defined a function called process in the existing workspace. Once the process script is executed (without changing the file whatsoever), then the process function is available for use and process.R can be closed. Have patience, as process is a large script file and it will take a few minutes for it to execute. When it is complete, a message in the console will appear stating that PROCESS is ready for use. PROCESS was written in base R v3.6. PROCESS for R does not rely on any packages and does not require the installation of any packages prior to use. Because R is a script that defines a function rather than a package, the process script must be executed each time you begin an R session and intend to use the PROCESS function. To circumvent this, save the workspace after running process.R, either using the existing workspace name or a different one. With the workspace saved, PROCESS will be available for use and will function much like a package whenever using that workspace."* (Hayes, 2022, p. 611)

In short, you need to run the full script once every time you start a new R session (or use the approach described by Hayes above). One way to do that is to store the *process.R* script in the same directory as your script for analysis, and then to use ```source("process.R")```, which will run the code in the *process.R* script.

```{r}
source("process.R")
```


# Import the data set

Here we will be importing the ```.sav``` WVS data set

```{r, eval=T}
dat <- import("WVS_Cross-National_Wave_7_sav_v2_0.sav")
dim(dat)
```

We are calling our data set **dat** and asking to see the dimension of it. We see that the data set has 76897 subjects, and 548 columns. 

## Prepare the data set

In cases with large data sets like this we might want to select a subset of variables that we want to work with, since it is not easy to keep an overview of 548 variables. We first identify the variables that we want to work with. The variables we will use here are:

* Q260: sex, 1 = Male, 2 = Female
* Q262: age in years
* Y001: post-materialism index
* SACSECVAL: secular values
* Q112-Q120: Corruption Perception Index
* Q65-Q73: Lack of Confidence in the government

We first create the vector **vars** containing the variable names of the variables that we want to keep. Next, we create a new data set **dat2** with only these 17 variables, and make sure we did it correctly by looking at the the dimension of the data ```dim(dat2)```. We also look at the first 6 rows: ```head(dat2)```. These are quick checks that we have created the new data correctly. 

```{r, eval=T}
vars <- c("Q260","Q262", "Y001", "SACSECVAL", "Q112", "Q113", "Q114", "Q115", "Q116", "Q117", "Q118", "Q119", "Q120", "Q65", "Q69", "Q71", "Q72", "Q73")
dat2 <- dat[,vars]
dim(dat2)
head(dat2)
```


### Create composite scores

We will be using the composite scores for *Corruption Perception Index* and *Lack of Confidence in the government* instead of their single items. So, we first need to compute them, we will use the mean across all items for each of these indices.

```{r, eval=T}
dat2$Corrup <- rowMeans(dat2[,c("Q112", "Q113", "Q114", "Q115", "Q116", "Q117", "Q118", "Q119", "Q120")], na.rm=T)
dat2$LCGov <- rowMeans(dat2[,c("Q65", "Q69", "Q71", "Q72", "Q73")], na.rm=T)
head(dat2)
```

We compute the mean across the specified variables for each subject with ```rowmeans()```. Remember to include the ```na.rm=T``` argument, so the missing values are ignored. In this way, the mean score represents the mean for all of the items that the respondents did answer. Otherwise, you get a missing value for every individual that has a missing value on at least *one* of the items.   

Note that the PROCESS macro does not work with variables that are designated as factors. Categorical variables to be used in moderator analysis should therefore not be designated as factor variables when you are working with the PROCESS macro. In our data set this is not the case, so we can proceed.

### Recoding sex variable

We recode the variable "Sex" (Q260) with values 1 and 2 to having the values 0 and 1. The value 0 now means *male*, and 1 means *female*. When creating this type of variable, is recommended to name the variable by the category with the value of 1, in this case we will named of variable *Female*

```{r}
dat2$Female <- dat2$Q260-1
head(dat2)
```

### Select variables for analysis

Now, we will select only the variables of interest in a separate data set, excluding the individual items of the mean scores. 

```{r}
dat2 <- na.omit(dat2[,c("Female", "Q262", "Y001", "SACSECVAL", "Corrup", "LCGov")])
head(dat2)
dim(dat2)
```

The new **dat2** data set only include the 5 continuous variables of interest, and 1 binary variable. With the ```na.omit()``` function we are excluding all cases with some missing values.

# Moderation analysis steps

Moderation is split into multiple steps, (a) 

* Estimate the *main effects* model that includes only the predictors as a *normal* multiple regression.
* Estimate the *interaction* model that now also includes the interactions between predictors. 
* Compare the models: the *p-value* test the Null hypothesis of the two predictors being independent, and the change in $R^2$ represents the effect size magnitude of the interaction
* Probe the interaction: estimate the simple slopes, the slope for the focal predictor at fixed values of the moderator predictor.
* Plot the simple slopes

The first three steps test the Null Hypothesis of the interaction (with the respective effect size), while the last two steps seek to explain *how does the moderator affects the focal relation?*, and help interpret these conditional relations. 

We will see how to implement these steps for the two common scenarios with a categorical or a continuous moderator.

# Categorical moderator

For the categorical predictor model, we will have *Lack of confidence in the government* as the focal predictor of *Perception of corruption*, and *Sex* as the categorical moderator.

## Main effects 

The main effects model presents both the effects of each predictor on the outcome.

```{r}
main_cat <- lm(Corrup ~ LCGov + Female, data=dat2)
summary(main_cat)
```


### Interpretation

* We reject the null hypothesis of both predictors being equally good predictors as the mean model, $F(2, 71630) = 4341, p < .001$. 
* Both predictors explained 11% of the variance in the outcome ($R^2 = 0.108$).
* The average outcome score for *Male* with 0 *Lack of confidence in the government* is 2.55 ($b_0 = 2.55, SE = 0.01, p < .001$)
* As *Lack of confidence in the government* increase by 1 unit, *Perception of corruption* increases by 0.36 ($b_1 = 0.36, SE = 0.004, p < .001$), holding *Sex* constant.
* As *Sex* changes, *Perception of corruption* decreases by 0.023 for *Female* ($b_2 = -0.023, SE = 0.006, p < .001$), holding *Lack of confidence in the government* constant.

## Interaction model

The code for a simple moderation model using the PROCESS macro is ```process (data = my_data_frame, y = "dependent_variable", x = "independent_variable", w ="moderator_variable", model = 1)```.

*my_data_frame* is a place holder for the name of the data frame with the data that we want to use, and we have to replace the placeholder variable names with the actual variable names from our data set. We additionally add the option ```intprobe=1``` to already get the results for probing the interaction.

```{r}
process(data = dat2, y = "Corrup", x = "LCGov", w = "Female", model = 1, intprobe=1)
```

The PROCESS macro named the interaction term ```Int_1```. The *p-value* for this term tests the Null Hypothesis that the 2 predictors are independent. In our example, the coefficient for the interaction term is -0.008 and the *p-value* is 0.313.

### Effect size

To assess the effect size of the interaction, we can look at this as the change in $R^2$ when the interaction is added. The output is shown under the header *Test(s) of highest order unconditional interaction(s):*, and we can see that the change in the explained variance is 0 ($\Delta R^2 = 0.000$), given by *R2-chng*. 

### Probing

Probing the interaction means to estimate the dependent focal regressions at specific levels of the moderator. When using the ```intprobe``` option, the PROCESS macro also shows the output for probing the interactions. When we use the option ```intprobe = x```, x defines the alpha level of interactions that you want to see probed. With ```intprobe = .05```, all interactions are probed that meet the standard .05 alpha level of statistical significance. With ```intprobe = 1``` we set the alpha level to 1 so that all interactions are probed regardless of their level of statistical significance.

The output for probing the interactions is shown under the heading *Conditional effects of the focal predictor at values of the moderator(s):*. Here we see that for the *Male* participants, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.373 ($b_{1M} = 0.373, SE = 0.006, p < .001$), and for the *Female* participants it is 0.365 ($b_{1F} = 0.365, SE = 0.006, p < .001$).

The difference in the slopes for women rather than men is -0.008 (0.3733 - 0.3653), which is exactly the slope for the interaction term ```Int_1```.

### Interpretation

* We fail fail to reject the null hypothesis of the 2 predictors being independent ($b_{*} = -0.008, SE = 0.008, p = 0.313$).
* The slope between *lack of confidence in the government* and *Perception of corruption* for female participants is 0.008 points lower than the slope for male participants ($b_{*} = -0.008, SE = 0.008, p = 0.313$).
* For *Male* participants, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.373 ($b_{1M} = 0.373, SE = 0.006, p < .001$).
* For *Female* participants, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.365 ($b_{1F} = 0.365, SE = 0.006, p < .001$).
* The addition of the interaction does not improve the model's predictive accuracy ($\Delta R^2 = 0.000$).

# Continuous moderator

For the continuous predictor model, we have *Lack of confidence in the government* as the focal predictor of *Perception of corruption*, and *Secular values* as the continuous moderator.


## Main effects 

The main effects model presents both the effects of each predictor on the outcome. With a continuous predictor, the main effects model is just a multiple regression model, nothing special yet

```{r}
main_cont <- lm(Corrup ~ LCGov + SACSECVAL, data=dat2)
summary(main_cont)
```

### Interpretation

* We reject the null hypothesis of both predictors being equally good predictors as the mean model (a model including only the intercept but no predictors), $F(2, 71630) = 6691, p < .001$. 
* Both predictors explained 16% of the variance in the outcome ($R^2 = 0.157$).
* The average outcome score for *Perception of corruption*, when both predictors are equal to 0 is 2.73 ($b_0 = 2.73, SE = 0.01, p < .001$)
* As *Lack of confidence in the government* increase by 1 unit, *Perception of corruption* increases by 0.46 points ($b_1 = 0.46, SE = 0.004, p < .001$), holding *Secular Values* constant.
* As *Secular Value* increases by 1 unit, *Perception of corruption* decreases by -1.11 points ($b_2 = -1.11, SE = 0.017, p < .001$), holding *Lack of confidence in the government* constant.


## Interaction model

The interaction model is an extension of the main effects model, that includes an interaction term between *Lack of confidence in the government* and *Secular Values*. 

In comparison to the model with *Sex* as a categorical moderator, we simply replace *Female* with the variable name for *Secular values*, namely *SACSECVAL*. Again, we want to get the output for probing the interactions, so we specify ```intprobe=1```. As we have seen in the example with a categorical moderator above, probing the interaction means to estimate the dependent focal regressions at specific levels of the moderator. With a categorical moderator, there are only two levels of the moderator that we can use. With a continuous moderator, however, it is not immediately clear for which values of the moderator we would like to see the slopes. With a continuous moderator we therefore need to choose values of interest of the moderator to estimate the simple slopes. Ideally, there are meaningful values for you to choose. Another common option is to use the mean, the value at 1 standard deviation (SD) below the mean, and the value one SD above the mean, to represents average, low, and high values of the moderator. With the PROCESS macro, we simply include the ```moments=1``` option, to probe the interaction for these values of the moderator. 

```{r}
process(data = dat2, y = "Corrup", x = "LCGov", w = "SACSECVAL", model = 1, intprobe=1, moments=1)
```

In the output, we see that the interaction term ```Int_1```, and the *p-value* for this term testing the Null Hypothesis that the 2 predictors are independent.

### Effect size

To assess the effect size of the interaction, we can look at this as the change in $R^2$ when the interaction is added. The output is shown under the header *Test(s) of highest order unconditional interaction(s):*, and we can see that the change in the explained variance is 
$\Delta R^2 = 0.0015$, as given by *R2-chng*. 

### Probing

If we go back to the coefficient for the interaction term, you can see that it is -0.243. This coefficient tells the change in the slope is when the moderator increases by 1 unit. Unless you have very clear metric of the moderator, this parameter is hard to interpret. It is therefore useful to probe the interaction by looking at the slope for specific values of the moderator.

The output shows that that for the low *Secular Values*, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.495 ($b_{LSV} = 0.495, SE = 0.005, p < .001$), for average *Secular Values* it is 0.453 ($b_{MSV} = 0.453, SE = 0.004, p < .001$), and for high *Secular Values* it is 0.410 ($b_{HSV} = 0.410, SE = 0.006, p < .001$). Keep in mind that high and low values of the moderator are represented by values one SD above or below the mean.

This way we see a a general trend, as *Secular Values* increases, the focal regression decreases strength. 

Another option with a continuous moderator is to estimate simple slopes for a long array of moderator values. A recommendation would be to test across a large number of possible values between the minimum and maximum of values for the moderator. With the PROCESS macro, we can do that using the ```wmodval``` option.

We first create a vector with the values at which we want to probe the interactions. In this case, we use steps of 0.1 between the minimum and the maximum of the SACSECVAL variable.

```{r}
vals <- seq(from=min(dat2$SACSECVAL, na.rm=T), to=max(dat2$SACSECVAL, na.rm=T), by=.1)
vals
```

Then we provide the ```vals``` vector to the ```wmodval``` option:

```{r}
process(data = dat2, y = "Corrup", x = "LCGov", w = "SACSECVAL", model = 1, intprobe=1, wmodval = vals)
```

We can now see that the slope of interest changes from 0.54 to 0.32 from the lowest and highest values. And for every one of them we reject the null hypothesis of each simple slope being equal to 0 


### Interpretation

* We reject the null hypothesis of the 2 predictors being independent ($b_{*} = -0.243, SE = 0.021, p < .001$).
* As the moderator *Secular Values* increase by 1 unit, the focal regression (```Corrup ~ LCGov```) decreases by 0.243 points ($b_{*} = -0.243, SE = 0.021, p < .001$)
* For low *Secular Values*, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.495 ($b_{LSV} = 0.495, SE = 0.005, p < .001$)
* For average *Secular Values* it is 0.453 ($b_{MSV} = 0.453, SE = 0.004, p < .001$)
* For high *Secular Values* it is 0.410 ($b_{HSV} = 0.410, SE = 0.006, p < .001$).
* In the general trend we see that the focal slope goes from 0.54 at the minimum observed *Secular Values* to 0.32 at the maximum observed *Secular values* score. 
* The addition of the interaction increases the model's predictive accuracy by 0.15% ($\Delta R^2 = 0.0015$).

## Visualizing interactions 

Unfortunately, the PROCESS macro is not very useful for plotting an interaction. To generate plots of interactions, you can instead use the method described in the *Moderation analysis with lm()* tutorial. If you want to stick with the process macro, you can request the data necessary to generate a plot from the PROCESS macro using the ```plot = 2``` option, but you then have to create a graph manually.

# PROCESS models

When working with the PROCESS macro, the argument ```model``` specifies the type of model. Such as moderation, mediation, serial, parallel, conditional processes. In this tutorial we showed the simple moderation example. The other numbered models in PROCESS are available in Appendix A of *Introduction to Mediation, Moderation, and Conditional Process Analysis* (Hayes, 2022)

# References

Hayes, Andrew F. (2022). Introduction to mediation, moderation, and conditional process analysis: A regression-based approach (Third edition, Vol. 1–1 online resource (xx, 732 pages) : illustrations.). The Guilford Press.


