---
title: "Moderation with bruceR"
author: "Mauricio Garnier-Villarreal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 5
editor_options:
  chunk_output_type: inline
---

# What is moderation analysis?

With moderation analysis, we are trying to find out whether the effect or association between two variables depends on another variable. Let's say that we are interested in the association between *Lack of confidence in the government* and *Perception of corruption*, but we specifically want to know whether the association depends on the *Sex* or the *Secular Values* of individuals. The latter two variables are called *moderators* of the association between *Lack of confidence in the government* and *Perception of corruption*.

# Setup the R session

When we start working in R, we always need to setup our session. For this we need to set our working directory, in this case I am doing that for the folder that holds the downloaded [World Values Survey (WVS)](https://www.worldvaluessurvey.org/) ```SPSS``` data set

```{r, eval=F, error=F}
setwd("~path_to_your_file")
```


The next step for setting up our session will be to load the packages that we will be using

```{r, eval=T, message=F}
library(rio)
library(effectsize)
library(marginaleffects)
library(ggplot2)
library(parameters)
library(car)
library(bruceR)
library(tidyr)
```


# Import the data set

Here we will be importing the ```.sav``` WVS data set

```{r, eval=T}
dat <- import("WVS_Cross-National_Wave_7_sav_v2_0.sav")
dim(dat)
```

Here we are calling our data set **dat** and asking to see the dimension of it. We see that the data set has 76897 subjects, and 548 columns. 

## Prepare the data set

In cases with large data sets like this we might want to select a subset of variables that we want to work with. Since it is not easy to see 548 variables. 

```{r, eval=T}
vars <- c("Q260","Q262", "Y001", "SACSECVAL", "Q112", "Q113", "Q114", "Q115", "Q116", "Q117", "Q118", "Q119", "Q120", "Q65", "Q69", "Q71", "Q72", "Q73")
dat2 <- dat[,vars]
dim(dat2)
head(dat2)
```

Here we are first creating a vector with the variable names for the ones I want to keep. You can see all variable names for the full data set as well:

```{r, eval=F}
colnames(dat)
```

After identifying which variables we will work with, we create a new data set **dat2** with only these 17 variables, and make sure we did it correctly by looking at the the dimension of the data **dim(dat2)**. We also look at the first 6 rows: **head(dat2)**. These are quick checks that we have created the new data correctly. 

The variables we will use here are:

* Q260: sex, 1 = Male, 2 = Female
* Q262: age in years
* Y001: post-materialism index
* SACSECVAL: secular values
* Q112-Q120: Corruption Perception Index
* Q65-Q73: Lack of Confidence in the government

### Create composite scores

We will be using the composite scores for *Corruption Perception Index* and *Lack of Confidence in the government* instead of their single items. So, we first need to compute them, we will use the mean across all items for each composite

```{r, eval=T}
dat2$Corrup <- rowMeans(dat2[,c("Q112", "Q113", "Q114", "Q115", "Q116", "Q117", "Q118", "Q119", "Q120")], na.rm=T)
dat2$LCGov <- rowMeans(dat2[,c("Q65", "Q69", "Q71", "Q72", "Q73")], na.rm=T)
head(dat2)
```
With the ```rowmeans()``` we compute the mean across the specified variables, for each subject. Remember to include the ```na.rm=T``` argument, so the missing values are properly ignored.   

### Transform categorical variables to ```factor()```

when we want to use categorical variables as predictors, it is recommended to transformed them as ```factor``` type in ```R```. In this case we will use the *Sex* variable ```Q260```, which is coded as 1 and 2 right now. 

With the ```factor()``` function we an transform it, and change the numbers to meaningful labels. 

```{r}
dat2$Sex <- factor(dat2$Q260, 
                       levels = 1:2,
                       labels = c("Male","Female") )
```


### Set variables for analysis

Now, we will select only the variables of interest in a separate data set. 

```{r}
dat2 <- drop_na(dat2[,c("Sex", "Q262", "Y001", 
                        "SACSECVAL", "Corrup", "LCGov")])
head(dat2)
dim(dat2)
```
The new ```dat2``` data set only include the 6 continuous variables of interest, and 1 binary variable. With the ```drop_na()``` function we are excluding all cases with some missing values.


# Moderation analysis steps

Moderation is split into multiple steps, (a) 

* Estimate the *main effects* model that includes only the predictors as a *normal* multiple regression.
* Estimate the *interaction* model that now also includes the interactions between predictors. 
* Compare the models: the *p-value* test the Null hypothesis of the two predictors being independent, and the change in $R^2$ represents the effect size magnitude of the interaction
* Probe: estimate the simple slopes, the slope for the focal predictor at fixed values of the moderator predictor.
* Plot the simple slopes

The first three steps test the Null Hypothesis of the interaction (with the respective effect size), while the last two steps seek to explain *how does the moderator affects the focal relation?*, and help interpret these conditional relations. 

We will see how to implement these steps for two common interaction scenarios, with categorical and continuous moderators

# Categorical moderator

For the categorical predictor model, we will have *Lack of confidence in the government* as the focal predictor, and *Sex* as the categorical moderator. Having them predict the *Perception of corruption*

## PROCESS function

Here we will show the use of the function ```PROCESS()``` from the ```bruceR``` package. This function follows the [process](https://www.processmacro.org/) approach by Andrew Hayes. 

For basic moderation we need to specify the arguments ```data```, which is the outcome variable ```y```, which is the focal predictor ```x```, and which is the moderator predictor ```mods```, and if we wish to center the variables ```center```. 

```{r}

mp1 <- PROCESS(data = dat2, 
               y = "Corrup", 
               x = "LCGov", 
               mods ="Sex", 
               center = FALSE)

```

The first section of the output presents the two models estimated, first the model with only the focal predictor (notice this is not the full main effects model), and then the model with the interaction. 

In the second part, the specific elements of the interaction are shown. First we see the *F* test comparing the main effects model and the interaction model. This is the Null Hypothesis Significance test (NHST) of the moderation. Here we fail to reject the null hypothesis of no interaction $F(1, 71629) = 1.02, p = .313$ 



### Probing

Probing the interaction means to estimate the dependent focal regressions at specific levels of the moderator. To see and test the respective regressions. Before here we can talk about the 2 predictors being independent without stating which is the moderator and which is the focal predictor. When we probing and plotting the interactions, we need to choose a moderator and focal predictor. Notice that the model does not know which predictor is which, you have to decide this. 

When including a categorical variable, most likely this one will be the moderator, so in our case *Sex* will be treated as the moderator.

This is shown at the end of the output, the slopes for all groups in the categorical moderator, in this case ```Sex```. We can also extract these simple slopes

```{r}
print(mp1$results[[1]]$simple.slopes, digits=3)
```

Here we see that for the *Male* participants, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.373 ($b_{1M} = 0.373, SE = 0.006, p < .001$), and for the *Female* participants it is 0.365 ($b_{1F} = 0.365, SE = 0.006, p < .001$)

If we look back at the interaction regression output, you will see that the slope for the interaction term ```LCGov:SexFemale``` is -0.008, which is the slope difference between mean and women (0.3733 - 0.3653), this is an ease of interpretation when the moderator is categorical. 


### Plotting 

A last part is to plot the interactions, here we will show how to plot the interactions with the ```marginaleffects``` package

In the ```marginaleffects``` package, we have to first specify on ```lm()``` model that includes the interaction, we are plotting the *conditional* relation, and the same scale as the outcome variable in the *y-axis*. Then we need to specify our focal and moderator predictor in the ```by``` argument, the function will see the first variable as the focal one and the second one as the moderator. We will need to provide the interaction model from the output of the ```POCESS```function as ```mp1$model.y```

```{r}
plot_predictions(mp1$model.y,
                 by = c("LCGov", "Sex") )
```


Notice that when the moderator is a ```factor()``` type variable, ```plot_predictions``` will automatically plot the focal regression for all categories, and set the respective labels. 


### Effect size

The next way to compare the models relates to the effect size of the interaction, we can look at this as the change in $R^2$ when the interaction is added, or the proportion of variance explained uniquely by the interaction

Which can be estimated the $\eta^2$ of the interaction term, as this is the proportion of explained variance uniquely by the interaction. We can get this with ```eta_squared()``` function, make sure to set the argument ```partial = F``` as this will ask for the full $\eta^2$ instead of the partial. And we use the ```Anova()``` function to use the type 2 sum of of squares 

```{r}
print(eta_squared(Anova(mp1$model.y, type=2), partial = F), digits=6)
```

### Interpretation


* We fail fail to reject the null hypothesis of the interaction and main effects model being equally good at predicting the outcome. Or, fail to reject the null hypothesis of the 2 predictors being independent ($F(1, 71629)=1.019, p = .313$).
* Female participants have in average a focal slope (```Corrup ~ LCGov```) 0.008 points lower than the male participants ($b_{*} = -0.008, SE = 0.008, p = 0.313$).
* For *Male* participants, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.373 ($b_{1M} = 0.373, SE = 0.006, p < .001$).
* For *Female* participants, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.365 ($b_{1F} = 0.365, SE = 0.006, p < .001$).
* The addition of the interaction increases the model's predictive accuracy by 0.0013% ($\eta^2 = 0.000013$).
* In the plots, we see that both groups slopes are close to each other. 


# Continuous moderator

For the continuous predictor model, we will have *Lack of confidence in the government* as the focal predictor, and *Secular Values* as the continuous moderator. Having them predict the *Perception of corruption*

## PROCESS function

When the moderator is continuous the ```PROCESS()``` function works the same way, you just provide the continuous predictor in the ```mods```argument. For basic moderation we need to specify the arguments ```data```, which is the outcome variable ```y```, which is the focal predictor ```x```, and which is the moderator predictor ```mods```, and if we wish to center the variables ```center```. 

```{r}
mp2 <- PROCESS(data = dat2, 
               y = "Corrup", 
               x = "LCGov", 
               mods ="SACSECVAL", 
               center = FALSE)

```

The first section of the output presents the two models estimated, first the model with only the focal predictor (notice this is not the full main effects model), and then the model with the interaction. 

In the second part, the specific elements of the interaction are shown. First we see the *F* test comparing the main effects model and the interaction model. This is the Null Hypothesis Significance test (NHST) of the moderation. Here we reject the null hypothesis of no interaction $F(1, 71629) = 129.58, p < .001$ 


### Probing

Probing the interaction means to estimate the dependent focal regressions at specific levels of the moderator. To see and test the respective regressions. Before here we can talk about the 2 predictors being independent without stating which is the moderator and which is the focal predictor. When we probing and plotting the interactions, we need to choose a moderator and focal predictor. Notice that the model does not know which predictor is which, you have to decide this. 

Here we are choosing to treat *Secular Values* as the moderator variable

The last section of the output presents the simple slopes. If the moderator is continuous and no test values are given, by default the function will estimate the simple slopes for the mean - 1SD, mean and mean + 1SD (low, medium, high). 


Here we see that for the low *Secular Values*, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.495 ($b_{LSV} = 0.495, SE = 0.005, p < .001$), for average *Secular Values* it is 0.453 ($b_{MSV} = 0.453, SE = 0.004, p < .001$), and for high *Secular Values* it is 0.410 ($b_{HSV} = 0.410, SE = 0.006, p < .001$). 

This way we see a a general trend, as *Secular Values* increases, the focal regression decreases strength. 

If we look back at the regression table for the interaction model, you will see that the slope for the interaction term ```LCGov:SACSECVAL``` is -0.243, which is the change in the slope when the moderator increases by 1 unit. This is a harder parameter to interpret, unless you have very clear metric of the moderator, otherwise it is easier to interpret that change with simple slopes

Another option with a continuous moderator is to estimate simple slopes for a long array of moderator values. A recommendation would be to test across a large number of possible values, get the minimum and maximum of values for the moderator

```{r}
c(min(dat2$SACSECVAL, na.rm=T), max(dat2$SACSECVAL, na.rm=T))
```

Then create a sequence of values going from the minimum, to the maximum, increasing by a small value (choose something that make sense for the metric of the moderator). In this case an increase every 0.1 points makes sense. 

```{r}
vals2 <- seq(from=min(dat2$SACSECVAL, na.rm=T), to=max(dat2$SACSECVAL, na.rm=T), by=.1)
vals2
```

Now we can add the new set of testing values to the ```mod1.val``` argument. This way we can see the change in the slope of interest at smaller steps of change of the moderator. Everything esle stays the same. 

```{r}
mp3 <- PROCESS(data = dat2, 
               y = "Corrup", 
               x = "LCGov", 
               mods ="SACSECVAL", 
               center = FALSE, 
               mod1.val = vals2)
```

Here we see that the slope of interest changes from 0.54 to 0.32 from the lowest and highest values. And for every one of them we reject the null hypothesis of each simple slope being equal to 0. 

### Plotting 

A last part is to plot the interactions, here we will show how to plot the interactions with the ```marginaleffects``` package

In the ```marginaleffects``` package, we have to first specify on ```lm()``` model that includes the interaction, we are plotting the *conditional* relation, and the same scale as the outcome variable in the *y-axis*. Then we need to specify our focal and moderator predictor in the ```by``` argument, the function will see the first variable as the focal one and the second one as the moderator. We will need to provide the interaction model from the output of the ```POCESS```function as ```mp3$model.y```

The only new argument we need to add is ```newdata```, where we define which values of the moderator do we wish to plot, and all observed values for the focal predictor. We already have same these from probing simple slopes

```{r}
plot_predictions(mp3$model.y,
                 newdata = datagrid(LCGov=unique(dat2$LCGov),
                                    SACSECVAL = c(.186, .361, .535) ),
                 by = c("LCGov", "SACSECVAL") )
```

We can also plot the simple slopes for a larger number of of test values, like the larger sequence set we created. For that we simply use the larger values we created before. This plot allow us to see the changes in greater detail. 

```{r}
plot_predictions(mp3$model.y, 
                 newdata = datagrid(LCGov=unique(dat2$LCGov),
                                    SACSECVAL = vals2),
                 by = c("LCGov", "SACSECVAL"))
```

### Effect size

The next way to compare the models relates to the effect size of the interaction, we can look at this as the change in $R^2$ when the interaction is added, or the proportion of variance explained uniquely by the interaction

Which can be estimated the $\eta^2$ of the interaction term, as this is the proportion of explained variance uniquely by the interaction. We can get this with ```eta_squared()``` function, make sure to set the argument ```partial = F``` as this will ask for the full $\eta^2$ instead of the partial. And we use the ```Anova()``` function to use the type 2 sum of of squares 

```{r}
print(eta_squared(Anova(mp3$model.y, type=2), partial = F), digits=4)
```

### Interpretation

* We reject the null hypothesis of the interaction and main effects model being equally good at predicting the outcome. Or, reject the null hypothesis of the 2 predictors being independent ($F(1, 71629)= 74.107, p < .001$).
* As the moderator *Secular Values* increase by 1 unit, the focal regression (```Corrup ~ LCGov```) decreases by 0.243 points ($b_{*} = -0.243, SE = 0.021, p < .001$)
* For low *Secular Values*, the slope between *lack of confidence in the government* and *Perception of corruption* is 0.495 ($b_{LSV} = 0.495, SE = 0.005, p < .001$)
* For average *Secular Values* it is 0.453 ($b_{MSV} = 0.453, SE = 0.004, p < .001$)
* For high *Secular Values* it is 0.410 ($b_{HSV} = 0.410, SE = 0.006, p < .001$).
* In the general trend we see that the focal slope goes from 0.54 at the minimum observed *Secular Values* to 0.32 at the maximum observed *Secular values* score. 
* The addition of the interaction increases the model's predictive accuracy by 0.15% ($\eta^2 = 0.0015$).
* In the plots, we see the decrease of the slopes across a range of possible moderator values.  


# References

Hayes, A.F. (2022). Introduction to mediation, moderation, and conditional process analysis: A regression-based approach (Third edition, Vol. 1–1 online resource (xx, 732 pages) : illustrations.). The Guilford Press.
