---
title: "Logistic Regression"
author: "Mauricio Garnier-Villarreal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: gfm
toc: true
toc-depth: 5
number-sections: true
code-copy: true
---

# Introduction

Using linear regression for a binary outcome violates several key assumptions of the ordinary least squares (OLS) model. First, linear regression assumes that the dependent variable is continuous and can take any real value, but a binary outcome is discrete and restricted to only two values (0 and 1). This leads to predicted values that can fall outside the meaningful $[0,1]$ probability range, producing impossible predictions. Second, the assumption of normally distributed residuals is violated because with a binary outcome, the errors are inherently heteroscedastic and follow a binomial distribution rather than a normal one. Moreover, the linear functional form implies that the effect of predictors is constant across all values, whereas for a binary outcome we typically expect a nonlinear S‑shaped relationship where changes in probability are smaller near the extremes. These violations make OLS regression inappropriate for binary data, necessitating alternatives such as logistic regression, which is specifically designed to model probabilities within the $[0,1]$ interval and accounts for the binomial error structure.

In many cases we will be interested in dichotomous dependent variable with \textcolor{blue}{Yes} vs. \textcolor{red}{No} responses, like Employed, Voted in the last election, Passed an exam, or Migration. 

When working with categorical response data, it is important to exercise caution if you are considering categorizing a continuous variable, as this practice can introduce several problems. Categorization often leads to bias in estimates, magnifies differences between groups that may not actually be distinct, and reduces the available variance in the data, which in turn diminishes statistical power and can obscure meaningful relationships. If categorization is necessary, it should only be done when there is a clear and justifiable criterion—such as when groups are significantly, qualitatively, or theoretically different from one another. Otherwise, it is generally advisable to use variables in their natural metric whenever possible, preserving the full information contained in the data. An agnostic approach, where the variable metric is chosen based on what best aligns with the research question rather than convenience or convention, helps ensure that the analysis remains faithful to the underlying phenomena being studied.

# LM vs GLM

The **General Linear Model** refers to the classical linear regression framework in which the outcome variable $Y$ is assumed to follow a normal distribution and the residuals are normally distributed. This model takes the form $Y = \beta_0 + \beta_1 X_1 + \dots + \beta_k X_k + \epsilon$, with $\epsilon \sim N(0, \sigma^2)$. In contrast, the **Generalized Linear Model (GLM)** generalizes this idea to response variables that are not normally distributed, such as binary, count, or categorical outcomes. GLM accomplishes this by retaining a linear predictor while allowing the outcome to follow any distribution from the exponential family (e.g., binomial, Poisson, negative binomial) and connecting the linear predictor to the expected value of the outcome through a link function.

Every GLM consists of three essential components:

1. **Random component**: This specifies the probability distribution of the outcome variable $Y$. For example, in ordinary linear regression we have $Y \sim N(\mu, \sigma^2)$; for binary outcomes we use $Y \sim \text{Binomial}(n, p)$; for count data we might use $Y \sim \text{Poisson}(\lambda)$.
2. **Systematic component**: This is the linear predictor $\eta = \beta_0 + \beta_1 X_1 + \dots + \beta_k X_k$, which combines the predictors linearly. No distributional assumptions are made about the $X$ variables.
3. **Link function** $g(\cdot)$: This function connects the expected value of $Y$ (the random component) to the systematic component. Formally, $g(E(Y)) = \eta$, or equivalently $E(Y) = g^{-1}(\eta)$. For ordinary linear regression, the link function is the identity: $g(\mu) = \mu$. For logistic regression (binary outcome), the link is the logit function: $g(\mu) = \ln\left(\frac{\mu}{1-\mu}\right)$. Other common link functions include the probit, complementary log-log, and log links, each chosen to respect the measurement scale of the outcome.

Thus, the GLM framework provides a flexible and unified way to model diverse data types while retaining the interpretability of linear predictors.

# Logistic Regression

Logistic regression is used when the outcome variable $Y$ is binary (e.g., success/failure, yes/no). In the Generalized Linear Model (GLM) framework, logistic regression is characterized by:

- **Random component**: $Y$ follows a binomial distribution. For a single trial, this is the Bernoulli distribution: $Y \sim \text{Binomial}(n=1, p)$, where $p$ is the probability of a "success" (e.g., voting for Wilders).
- **Systematic component**: The linear predictor $\eta = \beta_0 + \beta_1 X_1 + \dots + \beta_k X_k$ combines the predictors linearly.
- **Link function**: Because $p$ is bounded between 0 and 1, we cannot model $p$ directly as a linear function of the predictors. Instead, we use the **logit link** function, which transforms the probability $p$ to the log-odds scale (also called the logit of $p$):

$$
\text{logit}(p) = \ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X_1 + \dots + \beta_k X_k
$$

The log-odds are unbounded (range $-\infty$ to $+\infty$), so they can be modeled linearly. The odds themselves are $\frac{p}{1-p}$ and range from $0$ to $+\infty$.

Alternatively, we can express the probability $p$ directly by applying the inverse of the logit function (the logistic function):

$$
p = \frac{e^{\beta_0 + \beta_1 X_1 + \dots + \beta_k X_k}}{1 + e^{\beta_0 + \beta_1 X_1 + \dots + \beta_k X_k}}
$$

This creates an S‑shaped curve (the logistic curve) that ensures predicted probabilities always lie between 0 and 1. The logistic function is nonlinear in $X$, but the logit transformation linearizes the relationship.

Below is a visualization of the logistic function (blue) compared to a linear model (red). The linear model can produce probabilities outside [0,1], while the logistic curve asymptotes at 0 and 1.

```{r, echo=FALSE, fig.width=8, fig.height=4}
library(ggplot2)
x <- seq(-6, 6, 0.01)
y <- exp(x)/(1 + exp(x))
ggplot(data.frame(x = x, y = y), aes(x, y)) +
  geom_line(color = "navy", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  geom_hline(yintercept = c(0, 1), color = "black", alpha = 0.5) +
  labs(x = "Linear predictor (η)", y = "Pr(Y=1 | X)") +
  theme_minimal()
```

The GLM framework unifies many models through the choice of random component and link function, as summarized in the table below:

| **Model**  | **Random** | **Systematic** |  **Link**  |
|------------|:-----------|---------------:|:----------:|
| Regression | Normal     |          Mixed |  Identity  |
| ANOVA      | Normal     |    Categorical |  Identity  |
| Logistic   | Binomial   |          Mixed |   Logit    |
| Poisson    | Poisson    |          Mixed | Log-linear |


Logistic regression thus provides a principled way to model binary outcomes while respecting the probabilistic nature of the response.

## Interpreting Coefficients in Logistic Regression

In logistic regression, the coefficients $\hat{\beta}$ represent the change in the **log-odds** of the outcome for a one-unit increase in the predictor, holding all other variables constant. For example, if $\hat{\beta} = 0.5$, then a one-unit increase in $X$ is associated with a 0.5 increase in the log-odds of $Y=1$. While mathematically precise, the log-odds scale is not intuitively meaningful for most researchers or readers. How much is a 0.5 increase in log-odds? It does not correspond directly to a change in probability, and its interpretation depends on where on the logistic curve the change occurs.

A more common and interpretable transformation is the **odds ratio (OR)** , obtained by exponentiating the coefficient: $OR = e^{\hat{\beta}}$. The odds ratio indicates the multiplicative change in the odds of the outcome for a one-unit increase in the predictor. Odds ratios range from 0 to $\infty$, with $OR = 1$ meaning no effect (i.e., $\hat{\beta} = 0$). An $OR > 1$ indicates that as $X$ increases, the odds of $Y=1$ increase; an $OR < 1$ indicates decreasing odds. For a change of $c$ units in $X$, the odds are multiplied by $e^{c\hat{\beta}}$. For instance, if $\hat{\beta} = 0.5$, then $OR \approx 1.65$, meaning a one-unit increase in $X$ raises the odds by about 65%. If $\hat{\beta} = -0.5$, then $OR \approx 0.61$, meaning the odds decrease by about 39%.

Despite their popularity, odds ratios can still be misleading if interpreted as risk ratios, especially when the outcome is common. A more directly interpretable approach is to work with **predicted probabilities** $\hat{p}$. Using the inverse logit transformation, we compute the probability of $Y=1$ for any combination of predictors:

$$
\hat{p} = \frac{e^{\hat{\beta}_0 + \hat{\beta}_1 X_1 + \dots + \hat{\beta}_k X_k}}{1 + e^{\hat{\beta}_0 + \hat{\beta}_1 X_1 + \dots + \hat{\beta}_k X_k}}
$$

Predicted probabilities are bounded between 0 and 1 and are expressed on the probability scale, making them easy to communicate. For example, we can say "a 30‑year‑old woman with high education and religiosity score of 4 has a 4% predicted probability of voting for Wilders."

To summarize the effect of a predictor across the sample, we often compute **average marginal effects**, which measure the average change in predicted probability for a one‑unit increase in a predictor, holding other variables at their observed values. For a continuous predictor like religiosity, the average marginal effect might be 0.015, meaning that, on average, a one‑unit increase in religiosity raises the probability of voting for Wilders by 1.5 percentage points. For a categorical predictor like sex, we can compute the average difference in predicted probabilities between groups (e.g., women vs. men). These probability‑based interpretations are usually the most accessible for substantive conclusions.


# R Setup

First, load the necessary libraries and set your working directory (adjust the path to where your data file is located).

```{r setup, message=FALSE, warning=FALSE}
# Set working directory (modify this to your own path)
# setwd("~/path_to_your_data")

# Load required packages
library(summarytools)
library(marginaleffects)
library(effectsize)
library(caret)
library(DescTools)
library(ggplot2)
library(arm)
library(nonnest2)
library(MASS)
library(parameters)
library(patchwork)
library(rio)
library(car)
library(tidyr)
```

# Data Import and Preparation

We import the data file `logisticregression.sav` (SPSS format). The data contain information on voters: whether they voted for Wilders (`wilders`), age (`AGE`), gender (`female`), education level (`education`), and religiosity (`religious`). We remove rows with missing values and convert the outcome to a factor.

```{r data-import}
# Import data
dat <- import("logisticregression.sav")

# Remove missing values
dat <- drop_na(dat)

# Convert outcome to factor
dat$wilders <- as.factor(dat$wilders)

# View first rows
head(dat)

# Check dimensions
dim(dat)
```

## Recode Variables

We recode `female` into a factor with meaningful labels and also recode `education` with meaningful labels. Here I add a number at the begining of the label to control the order of the categorial levels.

```{r recode}
dat$education <- recode(dat$education, "0 = '0lower than high school';
                        1 = '1high school'; 
                        2 = '2more than high school' ")
dat$Sex <- recode(dat$female, "0='0male'; 1='1female'")
```

# Descriptive statistics

Let's explore the distribution of the outcome and the predictors.

## Outcome Frequencies

```{r freq-outcome}
freq(dat$wilders)   # 11% voted for Wilders
```
Here we see that about 11% of the sample voted for Wilders

## Scatter Plot of Outcome vs. Age

Since the outcome is binary, a scatter plot with jittered points can give a rough idea of how many subjects at each level of Age

```{r scatter-plot}
ggplot(dat, aes(x = AGE, y = wilders)) +
  geom_point(position = position_jitter(height = 0.05, width = 0), alpha = 0.5) +
  labs(y = "Voted for Wilders (0 = No, 1 = Yes)", 
       x = "Age")
```

## Descriptive Statistics for Predictors

First we use the `freq()` function for frequency tables for the categorical predictors. And the `decr()` function for descriptive information for the continuous predictors 

```{r descriptives}
freq(dat[, c("education", "Sex")])
descr(dat[, c("AGE", "religious")])
```

# Logistic Regression Model

We fit a model predicting `wilders` from age, sex, education, and religiosity. We use the `glm()` function and specify `family = binomial(link = "logit")`. Where `glm()` stands for *generalized linear models*, and with `family = binomial(link = "logit")` we specify the random component as *binomial* distribution, and the link function as *logit* link. 

```{r model-fit}
regf <- glm(wilders ~ AGE + Sex + education + religious,
            data = dat, family = binomial(link = "logit"))
```

We can see the default `summary()`, as it displays the coefficients in log-odds units.

```{r summary}
summary(regf)
```

## Extracting Coefficients

- **Log-odds coefficients**:

```{r coef-logodds}
coef(regf)
```

- **Odds ratios** (exponentiated coefficients):

```{r coef-or}
exp(coef(regf))
```

## Nicely Formatted Tables with `parameters`

```{r parameters-logodds}

parameters(regf, digits = 3)
```

Interpretation (log-odds):
- **Age**: For a one‑year increase in age, the log‑odds of voting for Wilders decrease by 0.013, holding other variables constant.
- **Sex (female vs. male)**: Being female decreases the log‑odds by 0.532, holding other variables constant.
- **Education**: Compared to low education, middle education decreases log‑odds by 0.516; high education decreases by 1.26, holding other variables constant.
- **Religious**: A one‑unit increase in religiosity increases log‑odds by 0.16, holding other variables constant.

Now with odds ratios (exponentiate = TRUE):

```{r parameters-or}
parameters(regf, exponentiate = TRUE, digits = 3)
```

Interpretation (odds ratios):
- **Age**: Each additional year multiplies the odds by 0.987 (i.e., reduces odds by about 1.3%), holding other variables constant.
- **Sex (female)**: The odds for women are 0.588 times the odds for men (41.2% lower), holding other variables constant.
- **Education**: Middle education odds are 0.597 times than low education; high education odds are 0.283 times than low education, holding other variables constant.
- **Religious**: A one‑unit increase multiplies the odds by 1.17, holding other variables constant.

## Overall Effect of Categorical Predictors

Use `Anova()` from the `car` package to test the NHST effect of each predictor. When we have categorical predictors this is helpful to do the test for the overall variable, instead of a single group comparison

```{r anova}
Anova(regf, type = 2)
```

We reject the null hypothesis of no relation with *voting for Wilders* ($p < .05$).

We can also use the `parameters()` function around `Anova()` to get a cleaner table

```{r anova-par}
parameters(Anova(regf, type = 2))
```

## Predicted Probabilities

### Predicted Probability for a Specific Profile

We can compute the predicted probability for a 30‑year‑old woman with high education (education = 2) and religiosity = 4.

**Manual calculation using log‑odds:**

```{r prob-manual}
cc <- coef(regf)
lo <- cc["(Intercept)"] + 30*cc["AGE"] + 1*cc["Sex1female"] + 
  0*cc["education1high school"] + 
  1*cc["education2more than high school"] + 4*cc["religious"]
lo   # log-odds
invlogit(lo)   # probability
```
Here we first build the prediction from the model parameters (`cc`) in the log-odds metric (`lo`). This is the same as in linear regression, the difference is that this prediction is in a metric hard to interpret. Then we put the log-odds prediction into the inverse logit function (`invlogit()`), this function transforms the logit link function back to the probability metric. Just like in this equation


$$
\hat{p} = \frac{e^{\hat{\beta}_0 + \hat{\beta}_1 X_1 + \dots + \hat{\beta}_k X_k}}{1 + e^{\hat{\beta}_0 + \hat{\beta}_1 X_1 + \dots + \hat{\beta}_k X_k}}
$$

This is the same as 

```{r}
exp(lo)/(1+exp(lo))
```
Here we see that according to our model, a 30‑year‑old woman with high education (education = 2) and religiosity = 4 has a 4% chance of voting for Widlers. 

**Using `avg_predictions()` from `marginaleffects`:** we need to give the logistic regression model first, then in the `newdata` argument we specify the set of conditions of the predictions for which we want the predicted probability. And with `type = "response"` we ask for the prediction to be in the metric of the response variable, in this case it would be a probability

```{r prob-avgpredictions}
avg_predictions(regf, type = "response",
                newdata = datagrid(religious = 4, AGE = 30, Sex = "1female", education = "2more than high school"))
```

The probability is about 4%.

We can also ak for Predicted Probabilities Across a Range of Religiosity, at the average values of other predictors:

```{r prob-relig-range}
avg_predictions(regf, variables = list(religious = 0:10))
```



## Conditional Plots

We can use the `plot_predictions()` to visualize the effect of each predictor while holding others constant. And because we loaded the `patchwork` package we can set all the plots in a grid 

```{r cond-plots, fig.width=8, fig.height=6}
# Religiosity
p1 <- plot_predictions(regf, type = "response", 
                       condition = "religious") + 
  labs(title = "Religiosity")

# Education
p2 <- plot_predictions(regf, type = "response", 
                       condition = "education") + 
  labs(title = "Education")

# Sex
p3 <- plot_predictions(regf, type = "response",
                       condition = "Sex") + 
  labs(title = "Sex")

# Age
p4 <- plot_predictions(regf, type = "response", 
                       condition = "AGE") + 
  labs(title = "Age")

# Combine with patchwork
(p1 + p2) / (p3 + p4)
```

## Marginal Effects (Average Slope in Probability)

The average marginal effect of religiosity (change in probability per unit increase in religiosity):

```{r avg-slope}
avg_slopes(regf, variables = "religious")
```

Interpretation: A one‑unit increase in religiosity raises the probability of voting for Wilders by about 1.5 percentage points, on average.

## Differences in Probabilities Between Groups

For categorical predictors, we can estimate the group comparisons in the metric of difference in *predicted probabilities*. For education (pairwise comparisons):

```{r avg-comparisons-edu}
avg_comparisons(regf, variables = list(education = "pairwise"))
```

For sex, as it is a binary variable it is not necessary to specify `"pairwise"`, as there is only 1 possible comparison to make/ So, if we specify `variables = "Sex"`, we will get the average difference in *predicted probabilities* between men and women

```{r avg-comparisons-sex}
avg_comparisons(regf, type = "response", 
                variables = "Sex")
```


## Effect Size Transformations

Transform odds ratios to Cohen's d (for categorical predictors) or Pearson's r (for continuous predictors). We can transform the Odds Ratio into other measures of effect size, two that I recommend are to transform it into `Pearson r` for continuous predictors, and into `Cohen's d` for categorical predictors. This way they would be in commonly used metrics of effect size. 

We can do this with the `oddsratio_to_d()` and `oddsratio_to_r()` functions, where we first provide the respective OR, and it it then transformed into the desired metric `_to_d()` of `_to_r()`

```{r effectsize}
# Odds ratios (without intercept)
or_values <- exp(coef(regf)[-1])
or_values

# To Cohen's d
oddsratio_to_d(or_values[c("Sex1female", 
                           "education1high school", 
                           "education2more than high school")])

# To Pearson's r
oddsratio_to_r(or_values[c("AGE",
                           "religious")])
```
Here we see that between men and women, the OR equals a Cohen $d = -0.29$, between low education and mid education would be $d = -0.28$, between low education and high education would be $d = -0.69$. We can extrapolate the Cohen d between mid and high education as $d = -0.69 - -0.28 = -0.41$. 

For the continuous predictors, we see that the OR for Age are equivalent to $r = -0.0035$, and for Religiousness wit would be $r = 0.043$.


## Pseudo R‑Squared Measures

Several pseudo R² measures have been developed to provide a sense of model fit in logistic regression, though none directly correspond to the R² from ordinary least squares regression. Each of them try to approximate the idea of overall model effect size 

**McFadden's R²** is based on the log-likelihood of the fitted model relative to a null model (intercept only). It reflects the improvement in model fit. Values tend to be much lower than OLS R², with values between 0.2 and 0.4 often considered excellent fit.

$$
McFadden = \frac{-2LL_{null} - (-2LL_{k}) }{-2LL_{null}}
$$

**Cox and Snell R²** adjusts the likelihood ratio to have a maximum value less than 1. Because its upper bound is less than 1, it can be difficult to interpret, especially when the outcome is highly imbalanced.

$$
CoxSnell = 1-exp(-(-2LL_{null}-(-2LL_{k})))/n
$$

**Nagelkerke R²** rescales the Cox and Snell R² to have a maximum value of 1, making it more comparable to OLS R². It divides the Cox and Snell value by its theoretical maximum, providing a more intuitive measure of fit. In our model, Nagelkerke R² = 0.07, suggesting a modest improvement over the null model.

$$
Nagelkerke = \frac{CoxSnell \ R^2}{1-exp(-(-2LL_{null})/n)}
$$

**Tjur's R²** (also called the coefficient of discrimination) is calculated as the average predicted probability for observations with $Y=1$ minus the average predicted probability for observations with $Y=0$. It directly measures how well the model separates the two outcome groups. In our model, Tjur's R² = 0.04, meaning the predicted probabilities for those who voted for Wilders are, on average, 4 percentage points higher than for those who did not.

$$
Tjur = \left| E(\hat{Y} = 1| Y = 1) - E(\hat{Y} = 1| Y = 0)\right|
$$


All of these can be obtained using the `PseudoR2()` function from the `DescTools` package.

```{r pseudo-r2}
PseudoR2(regf, 
         which = c("McFadden", "CoxSnell", "Nagelkerke", "Tjur"))
```

From these measures, we recommend to use Nagelkerke R² and Tjur's R². 


## Model Comparison

We might also be interested in comparing logistic regression models, for this we will present two ways of model comparison.

### Likelihood Ratio Test (LRT)

First, we will go over the Likelihood Ratio Test (LRT), this compares the models log-likelihood between models. This follows a $\Delta \chi^2 = \chi^2_1 - \chi^2_2 $ distribution with $\Delta df = df_1 - df_2$. We can do this comparison with `anova()` function

Compare the full model against a null model (intercept only).

```{r lrt-null}
reg0 <- glm(wilders ~ 1, data = dat, family = binomial(link = "logit"))
anova(reg0, regf, test = "Chisq")
```

Compare against a model without education:

```{r lrt-noedu}
reg1 <- glm(wilders ~ AGE + female + religious,
            data = dat, family = binomial(link = "logit"))
anova(reg1, regf, test = "Chisq")
```

If we reject the null hypothesis between models, we generally choose the "larger" model (the one with more parameters). If we fail to reject the null hypothesis between models, we generally choose the "smaller" model (the one with less parameters). Both tests show that the full model fits significantly better ($p < .05$).



### Information Criteria (AIC, BIC)

We can also use relative measures of fit based on how much information we loose by using a model.

The most commonly used information criteria are: 
    -   Akaike Information Criteria (**AIC**): $-2\hat{l}_{m}+2k$. Approximation of the out-of-sample predictive accuracy 
    -   Bayesian Information Criteria (**BIC**): $-2\hat{l}_{m}+k\ln(n)$. Approximation of the Bayes factor comparison between 2 models
    
The lower the AIC and the BIC the better our model is. By themselves they do not say a thing $\to$ We use them to *compare* models!


Compare models using AIC and BIC (lower is better). Here we set the ICs in a table to check which model has the lowest

```{r ic-table}
ic <- matrix(c(AIC(reg0), BIC(reg0),
               AIC(reg1), BIC(reg1),
               AIC(regf), BIC(regf)),
             ncol = 2, byrow = TRUE)
colnames(ic) <- c("AIC", "BIC")
rownames(ic) <- c("Null Model", "Model without education", "Full Model")
ic
```

The full model has the lowest AIC and BIC, so it is the preferred model

#### Confidence Intervals for IC Differences

Use `icci()` from the `nonnest2` package to test whether the difference in IC between models is significant. By testing the Confidence Interval of the IC difference between models

```{r icci}
# Null vs. Model without education
icci(reg0, reg1)

# Model without education vs. Full model
icci(reg1, regf)
```

If the confidence interval for the difference does not include zero, the models differ significantly.

## Confusion Matrix

As the logistic regression is a model type that predicts a categorical variable, we can also evaluate the classification accuracy. This is how accurate are the models predicted category, compared to the observed data. We need to choose a cutoff (here we use 0.3, but 0.5 is common).

First we will save the model predicted probability of 1 (Voting for Wilders in this example). Then we choose a cut off point, in this case ($P(Y = 1) = 0.3$), so that someone that had a predicted probability of voting for Wilders equal or higher than $0.3$ we will classified them as $Vote_{pred} = 1$, while if the predicted probability of voting for Wilders is lower than $0.3$ we will classified them as $Vote_{pred} = 0$. 

Then we will use the observed and predicted variables win the `confusionMatrix()` function to estimate the classification accuracy measures

```{r confusion}
# Predicted probabilities
dat$pred_prob <- predict(regf, type = "response")

# Convert to predicted class (using cutoff 0.3)
cut_off <- 0.3
dat$wilders_pred <- as.factor(ifelse(dat$pred_prob > cut_off, 1, 0))

# Confusion matrix
confusionMatrix(dat$wilders_pred, dat$wilders)
```
This function provides several measures, here we will focus on three
- **Overall accuracy**: 88%
- **Sensitivity** (true positive rate for 0s): 99%
- **Specificity** (true positive rate for 1s): 2% (the model rarely predicts a "1")

In this example we see that the model mostly classify "0s" correctly, but it is bad at properly classifying "1s" correctly. 



## Conclusion

This tutorial covered the essential steps for conducting logistic regression in R, from data preparation to model interpretation and evaluation. Logistic regression is a powerful tool for binary outcomes, and understanding coefficients in both log‑odds and probability scales is key to meaningful interpretation.

# References

Agresti, A. (2019). An introduction to categorical data analysis (Third edition). John Wiley & Sons.


