---
title: "Multilevel Regression: centering"
author: "Mauricio Garnier-Villarreal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: gfm
toc: true
toc-depth: 5
code-copy: true
execute:
  echo: true
  warning: false
  eval: true
---


## Introduction to Centering in Multilevel Models

Centering is the process of subtracting a constant from a variable to create a new, shifted version of that variable. In single-level regression, centering is primarily used to give the intercept a meaningful interpretation: the expected value of the outcome when all predictors are zero (Aiken & West, 1991). If a predictor does not have a natural zero point (e.g., a Likert scale from 1 to 7), the intercept can be uninterpretable. Centering at the grand mean solves this by making zero represent the average score.

In multilevel models (MLMs), centering takes on additional importance because it helps us separate the effects of a predictor into its **within-cluster** and **between-cluster** components (Enders & Tofighi, 2007; Raudenbush & Bryk, 2002). Failing to separate these effects results in a "conflated" or "smushed" coefficient that is an uninterpretable blend of the two (Hoffman & Walters, 2022; Preacher et al., 2010).

### Why Do We Need to Separate These Effects?

Consider a study of students nested within schools. We want to examine the relationship between student Socio-Economic Status (SES) and math achievement. The relationship can be broken down into two distinct questions:

1.  **Within-School Effect**: For students in the *same* school, is higher SES associated with higher math scores?
2.  **Between-School Effect**: Do schools with a higher *average* SES tend to have higher average math scores?

These two effects can be, and often are, different in magnitude. The difference between them is known as the **contextual effect**: the effect of being in a higher-SES school, over and above the effect of one's own SES (Marsh's "Big-Fish-Little-Pond Effect"). A standard MLM with uncentered SES conflates these effects, producing a single coefficient that does not accurately represent either one.

This tutorial will guide you through the three main centering strategies for level-1 predictors, showing how to estimate and interpret these distinct effects using R.

### Packages and Data

We will use the `SB.csv` dataset (Snijders & Bosker, 1999) containing 2,287 pupils in 131 schools. Our outcome is a language post-test score (`langpost`), and our predictor is verbal IQ (`iq_verb`). We will also use several R packages for data manipulation, model fitting, and effect size calculation.

```{r setup, message=FALSE, warning=FALSE}
# Load required packages
library(lme4)
library(lmerTest)
library(rio)
library(parameters)
library(performance)
library(r2mlm)
library(dplyr)
library(ggplot2)
library(tidyr)

# Read data
SB <- import("SB.csv")
SB <- drop_na(SB)  # remove missing for simplicity
head(SB)
```


### Preliminary Step: Creating the Cluster Mean

To separate within- and between-cluster effects, we first need the cluster mean of our predictor. This will be used both for centering and as a level-2 predictor.

```{r}
# Calculate the mean iq_verb for each school
school_means <- SB %>%
  group_by(schoolnr) %>%
  summarise(iq_mean = mean(iq_verb, na.rm = TRUE))

# Merge the school mean back into the main data frame
SB <- left_join(SB, school_means, by = "schoolnr")

head(SB[, c("schoolnr", "iq_verb", "iq_mean")])

```


### The Uncentered Model (Baseline)

We start with a baseline model using the raw, uncentered predictor. This model will serve as a reference point. It includes a random intercept for schools but keeps the slope of iq_verb fixed for simplicity.

```{r}
model_uncentered <- lmer(langpost ~ 1 + iq_verb + (1 | schoolnr),
                         data = SB, REML = FALSE)

summary(model_uncentered)
parameters(model_uncentered)

```

**Interpretation of the Uncentered Model:**

  - The coefficient for `iq_verb`iq_verb is 2.49. This is a conflated, uninterpretable blend of the within-school and between-school effects. It does not tell us whether the effect is due to a student's own IQ relative to their classmates or due to the average IQ level of the school they attend.
  - The intercept is the predicted language score when `iq_verb = 0`, which is not meaningful in this context.

This model highlights why centering is necessary. We cannot answer our substantive questions about within- and between-school effects from these results.

### Grand Mean Centering

Grand Mean Centering (CGM) involves subtracting the overall grand mean of the predictor from every observation. The centered variable is created as:


$$ \text{iq\_cgm}_{ij} = \text{iq\_verb}_{ij} - \bar{x}_{..} $$

where $\bar{x}_{..}$ is the grand mean of `iq_verb`.

```{r}
# Calculate the grand mean
grand_mean_iq <- mean(SB$iq_verb)

# Create the grand-mean centered variable
SB$iq_cgm <- SB$iq_verb - grand_mean_iq

# Verify that the mean of the new variable is zero
mean(SB$iq_cgm)
```

#### Model 2a: Grand Mean Centering Only

This model includes only the grand-mean centered predictor. The coefficient will still be conflated because the between-school variance is still present in the predictor.

```{r}
model_cgm_only <- lmer(langpost ~ 1 + iq_cgm + (1 | schoolnr),
                       data = SB, REML = FALSE)

summary(model_cgm_only)
parameters(model_cgm_only)
```

**Interpretation:**

  - The coefficient for `iq_cgm` is 2.49—identical to the uncentered model. Grand mean centering alone does not disaggregate the effects; it only shifts the location of the intercept.
  - The intercept is now the predicted language score for a student with average IQ (since `iq_cgm = 0`), which is more interpretable: 4.61
  
#### Model 2b: Grand Mean Centering + Group Mean

To actually separate the within and contextual effects, we add the school mean (`iq_mean`) as a level-2 predictor.

```{r}
model_cgm_full <- lmer(langpost ~ 1 + iq_cgm + iq_mean + (1 | schoolnr),
                       data = SB, REML = FALSE)

summary(model_cgm_full)
parameters(model_cgm_full)
```


**Explanation of Syntax:**

  - `langpost ~ 1 + iq_cgm + iq_mean`: This specifies the fixed part of the model. We include an intercept (1), the grand-mean centered predictor, and the school-level mean.
  - `(1 | schoolnr)`: This is the random part, specifying a random intercept for each school. The slopes are fixed in this example for simplicity.
  - `REML = FALSE`: We use Maximum Likelihood (ML) estimation to allow for model comparisons later. For final inference, Restricted Maximum Likelihood (REML) is generally preferred.
  
**Interpretation of Fixed Effects:**

  - Intercept ($\gamma_{00}$): The estimated math achievement for a student with an average IQ (since `iq_cgm = 0`) in a school with an average IQ (`iq_mean` is not centered yet). This intercept is not as clean as it could be because `iq_mean` is not centered.
  - iq_cgm ($\gamma_{10}$): This is the within-school effect. For students in the same school, a one-unit increase in a student's IQ relative to the grand mean is associated with a `r round(fixef(model_cgm_full)["iq_cgm"], 2)`-point increase in their language score. Because we have included `iq_mean`, this coefficient is a "pure" estimate of the within-school effect (Enders & Tofighi, 2007).
  - iq_mean ($\gamma_{01}$): This is the contextual effect. For two students with the same IQ (`iq_cgm` is held constant), a one-unit increase in their school's average IQ is associated with a `r round(fixef(model_cgm_full)["iq_mean"], 2)`-point increase in their language score. This represents the added benefit of being in a higher-IQ school, above and beyond one's own IQ.

To get the between-school effect (the effect of a school's average IQ on school-average achievement), we would add the within and contextual effects: $\gamma_{10} + \gamma_{01}$. For our model, this is `r round(fixef(model_cgm_full)["iq_cgm"] + fixef(model_cgm_full)["iq_mean"], 2)`.


### Group Mean Centering

Group Mean Centering (CWC) – also known as Centering Within Cluster – involves subtracting the cluster mean from each observation. The centered variable is created as:

$$ \text{iq\_cwc}_{ij} = \text{iq\_verb}_{ij} - \bar{x}_{.j} $$


where $\bar{x}_{.j}$ is the mean of `iq_verb` for school $j$.

```{r}
# Create the group-mean centered variable using the pre-calculated school means
SB$iq_cwc <- SB$iq_verb - SB$iq_mean

# Verify that the mean of the new variable is zero for each school
SB %>%
  group_by(schoolnr) %>%
  summarise(mean_cwc = mean(iq_cwc)) %>%
  pull(mean_cwc) %>%
  summary()
```

#### Model 3a: Group Mean Centering Only

This model includes only the group-mean centered predictor. Because all between-school variance has been removed from `iq_cwc`, its coefficient is a pure within-school effect, even without adding the group mean back in.

```{r}
model_cwc_only <- lmer(langpost ~ 1 + iq_cwc + (1 | schoolnr),
                       data = SB, REML = FALSE)

summary(model_cwc_only)
parameters(model_cwc_only)

```

**Interpretation:**

  - iq_cwc ($\gamma_{10}$): This is a pure estimate of the within-school effect. For students in the same school, a one-unit increase in a student's IQ relative to their school's average is associated with a `r round(fixef(model_cwc_only)["iq_cwc"], 2)`-point increase in their language score. This is the same within-effect we got from the full CGM model.
  - The intercept is the predicted language score for a student at their school's average IQ.
  

#### Model 3b: Group Mean Centering + Group Mean

To also obtain the between-school effect, we add the school mean (`iq_mean`) as a level-2 predictor.

```{r}
model_cwc_full <- lmer(langpost ~ 1 + iq_cwc + iq_mean + (1 | schoolnr),
                       data = SB, REML = FALSE)

summary(model_cwc_full)
parameters(model_cwc_full)
```


**Interpretation of Fixed Effects:**

- Intercept ($\gamma_{00}$): The estimated math achievement for a student with an IQ equal to their school's average (since `iq_cwc = 0`), in a school with an average IQ (`iq_mean = 0`? Again, not centered). This is the unadjusted mean for an average school.
  - iq_cwc ($\gamma_{10}$): This remains the pure within-school effect, unchanged from Model 3a.
  - iq_mean ($\gamma_{01}$): This is the between-school effect. A one-unit increase in a school's average IQ is associated with a `r round(fixef(model_cwc_full)["iq_mean"], 2)`-point increase in a student's predicted language score (for a student at their school's average IQ, where `iq_cwc = 0`).

To get the contextual effect from this model, we subtract the within effect from the between effect: $\gamma_{01} - \gamma_{10}$. For our model, this is `r round(fixef(model_cwc_full)["iq_mean"] - fixef(model_cwc_full)["iq_cwc"], 2)`, which should match the contextual effect from the full CGM model.

### The Fully Centered Model (CWC + Centered Group Mean)

In the previous models, the intercept's interpretation is still slightly awkward because the level-2 predictor (`iq_mean`) is not centered. To make the intercept maximally interpretable (i.e., the expected outcome for a student with average IQ in a school of average IQ), we can also center the group means at their grand mean.

**Centering the Group Means**

This is a form of centering at level-2. We create a new variable, iq_mean_c, by subtracting the grand mean of the school means from each school's mean.

```{r}
# Calculate the grand mean of the school means
grand_mean_schools <- mean(SB$iq_mean)

# Create the centered school mean variable
SB$iq_mean_c <- SB$iq_mean - grand_mean_schools

# Verify its mean is zero
mean(SB$iq_mean_c)
```

#### Model 4: The Final Model

We now fit the "fully centered" model with the group-mean centered predictor (`iq_cwc`) at level-1 and the centered school mean (`iq_mean_c`) at level-2.

```{r}
model_final <- lmer(langpost ~ 1 + iq_cwc + iq_mean_c + (1 | schoolnr),
                    data = SB, REML = FALSE)

summary(model_final)
parameters(model_final)
```

##### Detailed Interpretation of Final Model Results

The `parameters(model_final)` output provides the parameter estimates for our fully centered multilevel model. This model includes:

- **Level-1**: Group-mean centered IQ (`iq_cwc`), representing each student's deviation from their school's average IQ
- **Level-2**: Centered school mean IQ (`iq_mean_c`), representing each school's deviation from the overall average school IQ
- **Random intercept** for schools (`(1 | schoolnr)`), allowing each school to have its own baseline language score

###### Fixed Effects

The fixed effects describe the average relationships in the population.

| Parameter | Coefficient | SE | 95% CI | t-value | p-value |
|:---|:---:|:---:|:---:|:---:|:---:|
| **(Intercept)** | 40.74 | 0.28 | [40.18, 41.30] | 143.25 | < .001 |
| **iq cwc** | 2.41 | 0.07 | [2.27, 2.56] | 33.70 | < .001 |
| **iq mean c** | 4.00 | 0.30 | [3.41, 4.60] | 13.15 | < .001 |

###### Intercept ($\gamma_{00} = 40.74$)

The intercept represents the predicted language score for a student with:

- An IQ equal to their school's average (`iq_cwc = 0`)
- Attending a school with the overall average IQ (`iq_mean_c = 0`)

In substantive terms, a typical student (average IQ relative to their school) in a typical school (average school IQ) is predicted to have a language score of **40.74**. This is a highly interpretable baseline value.

###### iq cwc ($\gamma_{10} = 2.41$)

This is the **within-school effect** of IQ. For students in the same school, a one-point increase in a student's IQ relative to their school's average is associated with a **2.41-point increase** in their predicted language score.

- **Interpretation**: Within any given school, students with higher IQ than their schoolmates tend to have higher language scores. For every point a student's IQ exceeds their school's average, their language score is expected to be 2.41 points higher.
- **Statistical significance**: The effect is  significant ($p < .001$), and the 95% confidence interval [2.27, 2.56] does not include zero.

###### iq mean c ($\gamma_{01} = 4.00$)

This is the **between-school effect** of IQ. A one-point increase in a school's average IQ (compared to the overall average) is associated with a **4.00-point increase** in the predicted language score for a student at that school's average IQ.

- **Interpretation**: Schools with higher average IQ tend to have higher average language scores. For a student whose IQ is at their school's average, attending a school with an average IQ one point higher than the typical school is associated with a 4.00-point higher language score.
- **Statistical significance**: The effect is significant ($p < .001$), with a 95% confidence interval [3.41, 4.60].
- **Comparison with within effect**: The between-school effect (4.00) is substantially larger than the within-school effect (2.41), suggesting that school-level factors associated with average IQ have a stronger impact than individual-level IQ differences within schools.

###### Contextual Effect

The contextual effect (the effect of school average IQ above and beyond individual IQ) can be calculated as:

$$ 
\text{Contextual Effect} = \gamma_{01} - \gamma_{10} = 4.00 - 2.41 = 1.59 
$$

This means that for two students with the same individual IQ, the one attending a school with an average IQ one point higher is predicted to score **1.59 points higher** on the language test. This represents the pure contextual benefit of being in a higher-IQ school environment.

###### Random Effects

The random effects describe the variance components – how much schools and students vary around the fixed effects.

| Parameter | SD | Variance (SD²) |
|:---|:---:|:---:|
| **SD (Intercept: schoolnr)** | 2.78 | 7.73 |
| **SD (Residual)** | 6.49 | 42.12 |

###### SD (Intercept: schoolnr) = 2.78

This is the standard deviation of school intercepts around the fixed intercept. After controlling for school-average IQ, schools still vary in their adjusted mean language scores.

- **Interpretation**: The intercepts for different schools (their baseline language scores after accounting for IQ) are normally distributed with a standard deviation of 2.78.
- **Practical meaning**: About 95% of schools have adjusted mean language scores within approximately ±1.96 × 2.78 = ±5.45 points of the fixed intercept (40.74). This means school means range from about 35.29 to 46.19 after controlling for IQ.
- **Substantive implication**: Even after accounting for IQ differences, substantial between-school variation remains – other school-level factors (teaching quality, resources, etc.) likely contribute to these differences.

###### SD (Residual) = 6.49

This is the standard deviation of the level-1 residuals – the within-school variation among students after accounting for IQ.

- **Interpretation**: Students' observed language scores deviate from their school's predicted line with a standard deviation of 6.49 points.
- **Practical meaning**: About 95% of students have residuals within ±1.96 × 6.49 = ±12.72 points of their predicted score. This represents the unexplained within-school variation.
- **Substantive implication**: Considerable individual differences remain unexplained – other student-level factors (motivation, prior achievement, family background) likely contribute to these differences.

### Variance Partitioning

The total variance in language scores can be estimated from the random effects:

- **Between-school variance** = (2.78)² = 7.73
- **Within-school variance** = (6.49)² = 42.12
- **Total variance** = 7.73 + 42.12 = 49.85

The intraclass correlation (ICC) – the proportion of total variance that is between schools – is:

$$ 
\text{ICC} = \frac{7.73}{49.85} = 0.155 
$$

About **15.5%** of the variance in language scores is between schools, with the remaining 84.5% within schools. This confirms that most of the variation is at the student level, but school differences are still meaningful.


### Summary Table for Interpretation

| Parameter | Estimate | Interpretation |
|:---|:---:|:---|
| **Intercept (γ₀₀)** | 40.74 | Predicted language score for an average student (at school mean IQ) in an average school (at overall mean school IQ). |
| **Within-school IQ (γ₁₀)** | 2.41 | For students in the same school, a 1-point higher IQ (relative to school mean) predicts 2.41-point higher language score. |
| **Between-school IQ (γ₀₁)** | 4.00 | For a student at their school's mean IQ, attending a school with 1-point higher average IQ predicts 4.00-point higher language score. |
| **Contextual effect** | 1.59 | Net benefit of being in a higher-IQ school, holding individual IQ constant. |
| **School SD (τ₀₀)** | 2.78 | Schools vary around the intercept with SD of 2.78 (95% of schools within ±5.45 points). |
| **Residual SD (σ)** | 6.49 | Students vary around their school's predicted line with SD of 6.49 (95% within ±12.72 points). |
| **ICC** | 0.155 | 15.5% of total variance is between schools; 84.5% is within schools. |

### Conclusions from the Final Model

1. **IQ matters both within and between schools**: Both individual IQ relative to schoolmates (γ₁₀ = 2.41) and school-average IQ (γ₀₁ = 4.00) are strong, significant predictors of language achievement.

2. **Context matters**: The contextual effect of 1.59 confirms that school-level IQ has an effect above and beyond individual IQ – consistent with "peer effects" or school environment influences.

3. **Most variance is within schools**: With an ICC of 0.155, the majority of variation in language scores is at the student level, but school differences are still substantial and meaningful.

4. **Substantial unexplained variance remains**: Both within-school (SD = 6.49) and between-school (SD = 2.78) residual variance indicate that additional predictors at both levels could improve the model.

5. **The model successfully disaggregates effects**: By using group-mean centering and including the centered school mean, we have cleanly separated within-school and between-school effects, providing clear, interpretable estimates for each.


### Summary of Model Results

| Model | Intercept | Level-1 Slope | Level-2 Slope | Interpretation of Level-1 Slope |
| :--- | :--- | :--- | :--- | :--- |
| **Uncentered** | 11.17 | 2.49 | - | Conflated (uninterpretable) |
| **CGM Only** | 40.61 | 2.49 | - | Still conflated (no level-2 control) |
| **CGM Full** | 21.94 | 2.41 | 1.59 | Within effect (with `iq_mean` control) |
| **CWC Only** | 40.29 | 2.41 | - | Pure within effect |
| **CWC Full** | -6.63 | 2.41 | 4.00 | Within + between effects |
| **Final (Centered Means)** | 40.74 | 2.41 | 4.00 | Within + between, with clean intercept |

Notice that the within-effect from the CWC models (2.41) is different from the within-effect derived from the CGM models (2.49). This is because the CGM within-effect is conditional on controlling for the school mean, while the CWC within-effect is unconditional. They represent the same underlying relationship but in slightly different parameterizations. The between-effect (4.00) and the contextual effect (calculated as 4.00 - 2.41 = 1.59) are consistent across models.

### Effect Size Measures

We can now compare the effect sizes across these different models.

#### Using the `performance` package

The `r2()` function provides marginal (fixed effects only) and conditional (fixed + random effects) $R^2$.

```{r}
r2(model_uncentered)
r2(model_cgm_only)
r2(model_cgm_full)
r2(model_cwc_only)
r2(model_cwc_full)
r2(model_final)
```

You will notice that `model_uncentered` and `model_cgm_only` have the same marginal and conditional $R^2$. `model_cwc_only` has a different $R^2$ because it is a different model (it excludes the between-school variance from the predictor). `model_cgm_full`, `model_cwc_full`, and `model_final` all have the same $R^2$, as they are statistically equivalent; centering simply reparameterizes them without changing the overall model fit.

#### Using the `r2mlm` package

The `r2mlm()` function provides a much more detailed decomposition. We'll run it on the final model.

```{r}
r2mlm(model_final)
```

##### Detailed Explanation of `r2mlm()` Output

The `r2mlm()` function from the `r2mlm` package provides a comprehensive decomposition of variance explained in multilevel models, following the framework developed by Rights & Sterba (2019). The output is divided into two main sections: `$Decompositions` and `$R2s`.

###### 1. `$Decompositions` – Variance Components

This table shows the proportion of variance in the outcome variable (`langpost`) attributable to different sources. The columns represent:

- **total**: Proportion of the *total* outcome variance (sum of within-cluster and between-cluster variance).
- **within**: Proportion of the *within-cluster* variance only (variance among students within the same school).
- **between**: Proportion of the *between-cluster* variance only (variance among school means).

The rows represent different sources of variance:

- **fixed, within**: Variance explained by the fixed effect of the level-1 predictor (`iq_cwc`). This is the variance attributable to students' deviations from their school's mean IQ.
- **fixed, between**: Variance explained by the fixed effect of the level-2 predictor (`iq_mean_c`). This is the variance attributable to differences in average IQ between schools.
- **slope variation**: Variance explained by random slopes. In this model, the slope is fixed, so this is zero.
- **mean variation**: Variance explained by random intercepts (between-school differences in mean achievement after accounting for predictors).
- **sigma2**: Unexplained residual variance at level-1 (within schools).

###### Interpreting the Numbers:

| Source | Total | Within | Between | Interpretation |
|:---|:---:|:---:|:---:|:---|
| **fixed, within** | 0.256 | 0.332 | NA | 25.6% of total variance is explained by within-school IQ differences. Within schools, 33.2% of the within-school variance is explained by IQ. |
| **fixed, between** | 0.135 | NA | 0.588 | 13.5% of total variance is explained by between-school IQ differences. Between schools, 58.8% of the between-school variance is explained by average IQ. |
| **mean variation** | 0.094 | NA | 0.412 | 9.4% of total variance is due to remaining between-school differences (random intercepts) after controlling for IQ. Between schools, 41.2% of the between-school variance remains unexplained. |
| **sigma2** | 0.515 | 0.668 | NA | 51.5% of total variance is unexplained within-school variance. Within schools, 66.8% of the within-school variance remains unexplained. |

**Check**: The total column sums to 1.00 (0.256 + 0.135 + 0.094 + 0.515 = 1.00), confirming that all variance is accounted for.

###### 2. `$R2s` – R-Squared Measures

This table provides various R² measures, each representing the proportion of variance explained by different combinations of predictors. The notation follows Rights & Sterba (2019):

- **f1**: Variance explained by level-1 predictors via *fixed slopes* (within effect).
- **f2**: Variance explained by level-2 predictors via *fixed slopes* (between effect).
- **v**: Variance explained by *random slope variation* (here zero because slopes are fixed).
- **m**: Variance explained by *random intercept variation* (between-school differences after controlling for predictors).
- **f**: Variance explained by *all fixed effects* (f1 + f2).
- **fv**: Variance explained by *fixed effects + random slope variation* (f + v).
- **fvm**: Variance explained by *fixed effects + random slope variation + random intercept variation* (f + v + m) – this is the *conditional* R².

###### Interpreting the Numbers:

| Measure | Total | Within | Between | Interpretation |
|:---|:---:|:---:|:---:|:---|
| **f1** | 0.256 | 0.332 | NA | The within-school IQ effect explains 25.6% of total variance and 33.2% of within-school variance. |
| **f2** | 0.135 | NA | 0.588 | The between-school IQ effect explains 13.5% of total variance and 58.8% of between-school variance. |
| **v** | 0.000 | 0.000 | NA | No variance is explained by random slope variation (slopes are fixed). |
| **m** | 0.094 | NA | 0.412 | Remaining between-school differences explain 9.4% of total variance and 41.2% of between-school variance. |
| **f** | 0.391 | NA | NA | All fixed effects together explain 39.1% of total variance. |
| **fv** | 0.391 | 0.332 | NA | Fixed effects + random slopes (same as f because v=0) explain 39.1% of total variance and 33.2% of within variance. |
| **fvm** | 0.485 | NA | NA | The full model (fixed effects + random intercepts) explains 48.5% of total variance. This is the *conditional R²*. |

##### Key Takeaways from This Output:

1. **Within-school IQ is a strong predictor**: It explains 25.6% of total variance and 33.2% of within-school variance.
2. **Between-school IQ is even stronger at the school level**: It explains 58.8% of between-school variance, but only 13.5% of total variance because between-school variance is a smaller portion of the total.
3. **Substantial unexplained variance remains**: 51.5% of total variance (and 66.8% of within-school variance) is unexplained – there are other student-level factors not in the model.
4. **School differences persist after controlling for IQ**: Random intercepts explain 9.4% of total variance and 41.2% of between-school variance, meaning schools differ in ways not captured by average IQ.
5. **Total model performance**: The conditional R² (fvm) of 0.485 indicates that the full model explains 48.5% of the total variance in language scores.


### Summary of Interpretations and Recommendations

| Centering Strategy | Model | $\gamma_{10}$ (Level-1 Slope) | $\gamma_{01}$ (Level-2 Slope) | Intercept |
| :--- | :--- | :--- | :--- | :--- |
| **Uncentered** | Baseline | Conflated | N/A | Predicted outcome when `iq_verb` = 0. (Often meaningless) |
| **CGM** | Only | Conflated | N/A | Predicted outcome for student with avg. IQ. |
| | Full | Within Effect | Contextual Effect | Predicted outcome for avg. IQ student in school with `iq_mean`=0. |
| **CWC** | Only | Within Effect | N/A | Predicted outcome for student at school's avg. IQ. |
| | Full | Within Effect | Between Effect | Predicted outcome for student at school's avg. IQ in school with `iq_mean`=0. |
| **CWC + Centered Means** | Final | Within Effect | Between Effect | Predicted outcome for student at school's avg. IQ in school with avg. IQ. |

Recommendations:

  - **If your primary interest is a Level-1 predictor**, use Group Mean Centering (CWC). It provides an unbiased estimate of the within-cluster effect and more accurate variance components (Raudenbush & Bryk, 2002). You must include the cluster mean as a level-2 predictor to avoid conflating the effect and to obtain the between effect.
  - **If your primary interest is a Level-2 predictor**, and you have Level-1 covariates you need to control for, Grand Mean Centering (CGM) is often preferred. It effectively partials out the Level-1 covariate, providing a clean estimate of the level-2 effect (Enders & Tofighi, 2007).
  - **If you are interested in both within and between effects**, or testing contextual hypotheses, use either CGM or CWC with the cluster mean included. Both models will allow you to derive the same three effects (within, between, contextual). For the most interpretable intercept, also center your level-2 predictors (cluster means) at their grand mean.
  - **For a clean, fully disaggregated model**, always include the cluster mean of your Level-1 predictor in the Level-2 model. This separates the variance and prevents your Level-1 coefficient from being "smeared" with Level-2 variance.
  
  
### References

  - Aiken, L. S., & West, S. G. (1991). Multiple regression: Testing and interpreting interactions. Sage.
  - Enders, C. K., & Tofighi, D. (2007). Centering predictor variables in cross-sectional multilevel models: A new look at an old issue. Psychological Methods, 12(2), 121–138.
  - Hoffman, L., & Walters, R. W. (2022). Catching up on multilevel modeling. Annual Review of Psychology, 73, 659-689.
  - Hox, J. J., Moerbeek, M., & van de Schoot, R. (2018). Multilevel analysis: Techniques and applications (3rd ed.). Routledge.
  - Preacher, K. J., Zyphur, M. J., & Zhang, Z. (2010). A general multilevel SEM framework for assessing multilevel mediation. Psychological Methods, 15(3), 209–233.
  - Raudenbush, S. W., & Bryk, A. S. (2002). Hierarchical linear models: Applications and data analysis methods (2nd ed.). Sage.
  - Rights, J. D., & Sterba, S. K. (2019). Quantifying explained variance in multilevel models: An integrative framework for defining R-squared measures. Psychological Methods, 24(3), 309–338.
  - Snijders, T. A. B., & Bosker, R. J. (2012). Multilevel analysis: An introduction to basic and advanced multilevel modeling (2nd ed.). Sage.
  - Wu, Y.-W. B., & Wooldridge, P. J. (2005). The impact of centering first-level predictors on individual and contextual effects in multilevel data analysis. Nursing Research, 54(3), 212-216.
