---
title: "Multilevel Regression: interactions"
author: "Mauricio Garnier-Villarreal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: gfm
toc: true
toc-depth: 5
number-sections: true
code-copy: true
execute:
  echo: true
  warning: false
  eval: true
---


# What is Moderation Analysis?

Moderation occurs when the relationship between two variables depends on a third variable. In traditional single-level regression, a moderation (interaction) hypothesis is tested by including a product term in the model:

$$ Y = b_0 + b_1X + b_2Z + b_3XZ + e $$

In this model, the effect of $X$ on $Y$ is no longer constant; it becomes a function of $Z$:

$$ \text{Effect of } X = b_1 + b_3Z $$

If $b_3$ is statistically significant, we conclude that $Z$ moderates the $X \rightarrow Y$ relationship (or that $X$ moderates the $Z \rightarrow Y$).

In multilevel modeling (MLM), moderation can occur at different levels of analysis:
- **Level-1 interactions**: Both predictors are at level-1 (e.g., student-level variables interacting to predict a student outcome)
- **Level-2 interactions**: Both predictors are at level-2 (e.g., school-level variables interacting to predict school outcomes)
- **Cross-level interactions**: A level-2 variable moderates a level-1 relationship (e.g., school characteristics changing how student SES affects achievement)

This tutorial will guide you through estimating, probing, and interpreting each type of interaction using R.


## Packages and Data

We will use the `SB.csv` dataset (Snijders & Bosker, 1999) containing 2,287 pupils in 131 schools. Our outcome is a language post-test score (`langpost`). We will use several predictors: verbal IQ (`iq_verb`, level-1), group size (`groupsiz`, level-2), and percent minority (`percmino`, level-2). We'll also use the `popular.csv` dataset (Hox, 2010) for a binary moderator example.

```{r setup, message=FALSE, warning=FALSE}
# Load required packages
library(lme4)
library(lmerTest)
library(rio)
library(parameters)
library(performance)
library(r2mlm)
library(dplyr)
library(ggplot2)
library(tidyr)
library(marginaleffects)
library(sjPlot)


# Read data
SB <- import("SB.csv")
SB <- drop_na(SB)  # remove missing for simplicity
head(SB)

# For binary moderator example
pop <- import("popular2.sav")
pop <- drop_na(pop)
pop$girl <- factor(pop$sex, labels = c("boy", "girl"))
head(pop)
```

# Theoretical and Analytical Models

In MLM, interactions are specified by including product terms in the fixed part of the model. However, the model also includes random effects that account for clustering. The general form of a two-level model with interactions can be written as:

**Level 1 (within clusters):**
$$ Y_{ij} = \beta_{0j} + \beta_{1j}X_{1ij} + \beta_{2j}X_{2ij} + \beta_{3j}X_{1ij}X_{2ij} + e_{ij} $$

**Level 2 (between clusters):**
$$ \beta_{0j} = \gamma_{00} + \gamma_{01}W_{1j} + u_{0j} $$
$$ \beta_{1j} = \gamma_{10} + \gamma_{11}W_{1j} + u_{1j} $$
$$ \beta_{2j} = \gamma_{20} + u_{2j} $$
$$ \beta_{3j} = \gamma_{30} + u_{3j} $$

Where:
- $\gamma_{00}$ is the overall intercept
- $\gamma_{10}$, $\gamma_{20}$, $\gamma_{30}$ are average slopes
- $u$ terms are random effects (variances $\tau_{00}$, $\tau_{11}$, $\tau_{22}$, $\tau_{33}$ and covariances)
- $e_{ij}$ is the level-1 residual (variance $\sigma^2$)

The specific form depends on which effects are allowed to vary randomly. For testing interactions, the fixed effects ($\gamma$ coefficients) are our primary interest, but correctly specifying the random part is crucial for accurate standard errors.

# Level-1 Interactions

Level-1 interactions involve two predictors measured at the individual (lowest) level. For example, we might hypothesize that the effect of IQ on language scores depends on homework time. Both variables are at the student level.

## Model Specification

For a level-1 interaction, we include the product term in the level-1 model. We also need to decide whether to allow the slopes to vary randomly. A basic model with a level-1 interaction (between `iq_verb` and `groupsiz` – though note `groupsiz` is actually a level-2 variable in our data; for a pure level-1 example we would need two level-1 predictors). For demonstration, let's create a level-1 predictor `homework` (simulated) to interact with IQ.

```{r create-homework}
# Create a simulated homework variable (level-1)
set.seed(123)
SB$homework <- rnorm(nrow(SB), mean = 5, sd = 2)
SB$homework <- ifelse(SB$homework < 0, 0, SB$homework)  # no negative hours
```

We should always start with **main effects** model, in which we estimate the multilevel regression **without** the interaction between predictors, so a multiple regression with random intercepts. For the CI in MLM we ask for `profile` likelihood intervals, these are more accurate in MLM models 

```{r}
m0_l1 <- lmer(langpost ~ 1 + iq_verb + homework + (1 | schoolnr),
              data = SB, REML = FALSE)

summary(m0_l1)
parameters(m0_l1, ci_method = "profile")
```

**Explanation of syntax:**

- `langpost ~ 1 + iq_verb + homework` specifies the fixed part: intercept (1) and main effects of IQ and homework.
- `(1 | schoolnr)` adds a random intercept for schools, allowing the mean language score to vary across schools.
- `REML = FALSE` uses maximum likelihood (ML) instead of restricted maximum likelihood (REML), which is necessary for likelihood‑ratio tests.
- The `parameters()` function with `ci_method = "profile"` computes confidence intervals based on profile likelihood, which are more accurate for mixed models than Wald intervals.

**Interpretation:**
The fixed effects show that IQ is a strong positive predictor of language scores ($b = 2.49$, $p < .001$), while homework has a negligible and non‑significant effect ($b = -0.02$, $p = 0.727$). The random intercept variance ($\tau_{00} = 9.50$) indicates moderate variation between schools after accounting for the predictors. The residual variance ($\sigma^2 = 42.22$) represents within‑school variation.

Now we fit a model with an interaction between `iq_verb` and `homework`:

```{r}
m1_l1 <- lmer(langpost ~ 1 + iq_verb * homework + (1 | schoolnr),
              data = SB, REML = FALSE)

summary(m1_l1)
parameters(m1_l1, ci_method = "profile")
```

**Explanation of syntax:**

- `iq_verb * homework` is shorthand for `iq_verb + homework + iq_verb:homework`. The colon `:` denotes the interaction term.
- The rest of the formula is the same as the main‑effects model.

**Interpretation:**

- The interaction term `iq_verb:homework` has a coefficient of 0.029 ($p = 0.377$), which is not statistically significant. This means that the effect of IQ does not change significantly with different levels of homework.
- The main effect of IQ (2.34) now represents the effect of IQ when homework = 0. Because homework was not centered, this may not be a meaningful value. Similarly, the homework main effect (-0.37) is the effect when IQ = 0.
- The random intercept variance (9.49) and residual variance (42.21) remain similar to the main‑effects model.

We can also tests the significance of the interaction with the LRT model comparison, by comparing the main effects and interaction models. In this example the $p$-value will be the same as the $p$-value of the regression table as both predictors are continuous. If you have categorical predictors with more than 2 categories the results would not be equivalent and we recommend you to use the LRT comparison in those cases

```{r}
anova(m0_l1, m1_l1)
```

In this example, we fail to reject the null hypothesis of `iq_verb` slopes being equal across `homework` levels as $p = 0.377$. 


## Allowing Random Slopes

If we believe the IQ effect (or the interaction itself) varies across schools, we can add random slopes:

```{r l1-interaction-random}
m1_l1_random <- lmer(langpost ~ 1 + iq_verb * homework + 
                       (1 + iq_verb | schoolnr),
                     data = SB, REML = FALSE)

summary(m1_l1_random)
parameters(m1_l1_random, ci_method = "profile")
```

**Explanation:** The random part `(1 + iq_verb | schoolnr)` allows both the intercept and the slope of IQ to vary across schools. The `|` separates the random effects from the grouping variable. This means we estimate a random intercept variance, a random slope variance for IQ, and their covariance.

**Interpretation:** The random slope variance for IQ is 0.105, indicating that the effect of IQ on language scores varies across schools. The correlation between the random intercept and slope is -0.97, suggesting that schools with higher average language scores tend to have weaker IQ effects. The interaction term remains non‑significant ($p = 0.421$). Adding random slopes improved the model fit (AIC decreased from 15263 to 15248), but the interaction still does not reach significance.


## Effect size difference

We can also evaluate the overall model fit improvement when we add the interaction, by evaluating the change in $R^2$. We can first check this by the change in conditional and marginal $R^2$ from the `performance` package

```{r}
r2(m0_l1)
r2(m1_l1)
```
In this case we see that the conditional and marginal $R^2$ are basically equal between the main effects and interaction models. Indicating that the model fit doesn't improve by adding the interaction


```{r}
r2mlm(m0_l1, bargraph = F)
r2mlm(m1_l1, bargraph = F)
```
When comparing the detail particioning of $R^2$ by the `r2mlm()` function, we see that the improvement when adding the interaction is negligible. 




## Probing and Plotting Level-1 Interactions

As in single-level regression, we probe interactions by examining simple slopes of the focal predictor at different values of the moderator. We can use the marginaleffects package for this.

First, choose representative values of the moderator. Common choices are the mean and ±1 standard deviation.

```{r}
# Get mean and SD of homework
mean_hw <- mean(SB$homework, na.rm = TRUE)
sd_hw <- sd(SB$homework, na.rm = TRUE)

# Values for probing
hw_vals <- c(mean_hw - sd_hw, mean_hw, mean_hw + sd_hw)
names(hw_vals) <- c("-1 SD", "Mean", "+1 SD")
hw_vals <- round(hw_vals, 3)
hw_vals
```

Now compute simple slopes of IQ at these homework values:

```{r}
avg_slopes(m1_l1, 
           variables = "iq_verb",
           by = "homework",
           newdata = datagrid(homework = hw_vals))
```

**Explanation of syntax:**  
- `avg_slopes()` computes average marginal effects (or simple slopes) from a fitted model.  
- `variables = "iq_verb"` specifies the predictor whose slope we want.  
- `by = "homework"` indicates that we want separate slopes for different values of homework.  
- `newdata = datagrid(homework = hw_vals)` tells the function to evaluate the slopes at the three chosen homework levels.

**Interpretation:** The effect of IQ on language scores is 2.43 when homework is low (3.05 hours), 2.48 at average homework (5.02 hours), and 2.54 when homework is high (6.99 hours). Although the slopes increase slightly with homework, the interaction term was not significant, so these differences are not reliable.

We can also plot the interaction:

```{r}
plot_predictions(m1_l1, 
                 by = c("iq_verb", "homework"),
                 newdata = datagrid(iq_verb = unique(SB$iq_verb),
                                    homework = hw_vals ))
```

**Explanation:**  
- `plot_predictions()` generates a plot of predicted values.  
- `by = c("iq_verb", "homework")` places `iq_verb` on the x‑axis and creates separate lines for each value of `homework`.  
- The `newdata` argument defines the grid of predictor values for which predictions are made.

This plot shows the predicted language score across IQ values for different levels of homework, allowing visual assessment of whether the slopes differ. The lines appear nearly parallel, consistent with the non‑significant interaction.

## Model tables

We can use the `tab_model()` function to present the results from both models next to each other

```{r}
tab_model(m0_l1, m1_l1)
```


# Level-2 Interactions

Level-2 interactions involve two predictors measured at the cluster level. For example, we might hypothesize that the effect of school size on average achievement depends on school type (public vs. private). In our data, `groupsiz` (group size) and `percmino` (percent minority) are level-2 variables.

## Model Specification

For a level-2 interaction, we include the product term in the level-2 equation for the intercept. First, we need to create school-mean aggregated variables.

```{r l2-prep}
# Aggregate to school level
school_data <- SB %>%
  group_by(schoolnr) %>%
  summarise(langpost_mean = mean(langpost),
            groupsiz = first(groupsiz),
            percmino = first(percmino))
```



**Note:** This could be a single-level model at the school level, aggregated regressions. In a true multilevel context, we would fit this as a level-2 model with random intercepts, but with only 131 schools, we might treat schools as fixed or use MLM with random intercepts but level-2 predictors only. Here's how to specify it in MLM form, but including the random intercept:

As always we should start with the main effects model

```{r}
m0_l2 <- lmer(langpost ~ 1 + groupsiz + percmino + (1 | schoolnr),
                  data = SB, REML = FALSE)

summary(m0_l2)
parameters(m0_l2, ci_method = "profile")
```

**Interpretation:** Both group size (0.12, $p = 0.009$) and percent minority (-0.14, $p < .001$) are significant predictors of language scores. The random intercept variance (14.21) is larger than in the level‑1 models, because we are not yet including the strong level‑1 predictor IQ.

Then we can add the interaction

```{r}
m1_l2 <- lmer(langpost ~ 1 + groupsiz * percmino + (1 | schoolnr),
                  data = SB, REML = FALSE)

summary(m1_l2)
parameters(m1_l2, ci_method = "profile")
```

**Interpretation:**

- The interaction term `groupsiz:percmino` has a coefficient of 0.0033 ($p = 0.242$), which is not significant. This indicates that the effect of group size does not depend on percent minority (or vice versa).
- Because we have not centered the predictors, the main effects are conditional on the other predictor being zero, which may not be meaningful (e.g., `percmino = 0` is possible but may be outside the typical range).


## Centering Level-2 Predictors

For better interpretation, we should center level-2 predictors, especially when including interactions.

```{r}
# Center level-2 predictors
SB <- SB %>%
  mutate(groupsiz_c = groupsiz - mean(groupsiz, na.rm = T),
         percmino_c = percmino - mean(percmino, na.rm = T))

m1_l2_centered <- lmer(langpost ~ 1 + groupsiz_c * percmino_c + (1 | schoolnr),
                       data = SB, REML = FALSE)

summary(m1_l2_centered)
parameters(m1_l2_centered, ci_method = "profile")

```

**Explanation:** Centering subtracts the mean, so the new variables have a mean of zero. This makes the main effects interpretable as the effect at the mean of the other predictor.

**Interpretation:**  
- Now the main effect of `groupsiz_c` (0.117, $p = 0.017$) is the effect of group size on language scores at the average percent minority.  
- The main effect of `percmino_c` (-0.128, $p < .001$) is the effect of percent minority at the average group size.  
- The interaction term remains unchanged (0.0033, $p = 0.242$).

As before, we can test the significance of the interaction with a model comparison between the main effects and interaction model. 

```{r}
anova(m0_l2, m1_l2)
```
In this case we fail to reject the null hypothesis of a level-2 interaction, as $p = 0.242$. 


## Effect size difference

We can evaluate the improvement in model fit by the change in $R^2$. In tis case we see that the main effect model even present higher conditional and marginal $R^2$ 

```{r}
r2(m0_l2)
r2(m1_l2)
```

Similarly, with the $R^2$ decmpisition, we see that the interaction present neglible improvent in model fit. 

```{r}
r2mlm(m0_l2, bargraph = F)
r2mlm(m1_l2, bargraph = F)
```



## Probing and Plotting Level-2 Interactions

Probing a level-2 interaction follows the same logic as level-1: examine simple slopes of one predictor at conditional values of the other.

```{r}
# Values for percent minority
perc_vals <- c(mean(SB$percmino) - sd(SB$percmino),
               mean(SB$percmino),
               mean(SB$percmino) + sd(SB$percmino))
perc_vals <- round(perc_vals, 1)
names(perc_vals) <- c("-1 SD", "Mean", "+1 SD")
perc_vals <- perc_vals - mean(SB$percmino)
perc_vals <- round(perc_vals, 3)
perc_vals

# Simple slopes of groupsiz at these values
avg_slopes(m1_l2_centered,
           variables = "groupsiz_c",
           by = "percmino_c",
           newdata = datagrid(percmino_c = perc_vals))
```

**Interpretation:** The effect of group size on language scores is 0.074 when percent minority is low (-12.98% from mean), 0.117 at the mean, and 0.161 when percent minority is high (+13.02% from mean). Although the slope increases with percent minority, the interaction is not significant, so these differences are not reliable.

We need to adjust the values because the predictor is centered. Alternatively, we can use the original metric with appropriate centering.

Plotting:

```{r}
plot_predictions(m1_l2_centered, 
                 by = c("groupsiz_c", "percmino_c"),
                 newdata = datagrid(groupsiz_c = unique(SB$groupsiz_c),
                                    percmino_c = perc_vals))
```

This shows how the relationship between group size and language scores changes across levels of percent minority. The lines are slightly diverging, but the interaction is not significant.

## Model tables

We can present the model parameters in a table next to each other

```{r}
tab_model(m0_l2, m1_l2)
```

# Cross-Level Interactions

Cross-level interactions are the most distinctive type of moderation in multilevel models. They occur when a level-2 variable moderates a level-1 relationship. For example, we might hypothesize that the effect of student IQ on language scores depends on school-level characteristics like group size or percent minority.

## Model Specification

A cross-level interaction is specified by including the level-2 predictor in the level-2 equation for the slope of the level-1 predictor. In the combined model, this creates a product term between the level-1 and level-2 variables.

As always we start with the main effects model (without interactions), in this case we have random intercept and slope

```{r}
m0_cross <- lmer(langpost ~ 1 + iq_verb + groupsiz_c + (1 + iq_verb | schoolnr),
                data = SB, REML = FALSE)

summary(m0_cross)
parameters(m0_cross, ci_method = "profile")
```

**Interpretation:**  
- IQ has a strong positive effect (2.51, $p < .001$).  
- Group size has a small positive effect that is not significant (0.05, $p = 0.149$).  
- The random intercept variance is 61.89 and the random slope variance for IQ is 0.203, indicating substantial variation in the IQ effect across schools.  
- The correlation between intercept and slope is -0.97, meaning schools with higher average language scores tend to have weaker IQ effects.

Now, we can estimate the interaction between level-2 and level-1 predictors

```{r}
m_cross <- lmer(langpost ~ 1 + iq_verb * groupsiz_c + (1 + iq_verb | schoolnr),
                data = SB, REML = FALSE)

summary(m_cross)
parameters(m_cross, ci_method = "profile")
```

**Explanation:**

- `iq_verb * groupsiz_c` includes the main effects and the cross-level interaction.
- `(1 + iq_verb | schoolnr)` allows both the intercept and the IQ slope to vary across schools. This is important because if we believe a level-2 variable moderates the IQ slope, there must be slope variation to explain (Aguinis, Gottfredson, & Culpepper, 2013).


**Interpretation:**

- The interaction term `iq_verb:groupsiz_c` has a coefficient of -0.028 ($p = 0.008$). This means that for a one‑unit increase in group size (centered), the slope of IQ decreases by 0.028. In other words, the effect of IQ on language scores is weaker in larger schools.
- The main effect of IQ (2.50) is the effect of IQ at the mean of group size (since `groupsiz_c` is centered).
- The random slope variance for IQ (0.166) is smaller than in the main‑effects model (0.203), indicating that group size explains part of the between‑school variation in the IQ slope.
- The correlation between intercept and slope is still very strong (-0.97).


We can alsways test the signifficance of the interaction with a model comparison, in this case we reject the null hypothesis of no cross-level interaction as $p = 0.008$ 

```{r}
anova(m0_cross, m_cross)
```

  
  
## Testing the Need for Random Slopes

Before testing a cross-level interaction, we should verify that the slope of the level-1 predictor actually varies across clusters. We can compare models with and without the random slope.

```{r}
# Model with fixed slope
m_fixed <- lmer(langpost ~ 1 + iq_verb * groupsiz_c + (1 | schoolnr),
                data = SB, REML = FALSE)

# Model with random slope (already fit)
anova(m_fixed, m_cross)
```

A significant chi-square test indicates that allowing the slope to vary improves model fit, justifying the search for cross-level moderators. Here, the test is significant ($p < .001$), confirming that the IQ slope varies across schools.

## Effect size difference

In this case we see that the model with the interaction increases the marginal $R^2$ by 0.5%  

```{r}
r2(m0_cross)
r2(m_cross)
```

Similarly, we can see small model fit improvements in the $R^2$ decomposition

```{r}
r2mlm(m0_cross, bargraph = F)
r2mlm(m_cross, bargraph = F)
```


## Probing and Plotting Cross-Level Interactions

Probing a cross-level interaction involves examining the simple slope of the level-1 predictor (IQ) at different values of the level-2 moderator (group size).

```{r}
# Values of group size (centered) at which to probe
gs_vals <- c(mean(SB$groupsiz) - sd(SB$groupsiz),
             mean(SB$groupsiz),
             mean(SB$groupsiz) + sd(SB$groupsiz))
gs_vals_centered <- gs_vals - mean(SB$groupsiz)
names(gs_vals_centered) <- c("-1 SD", "Mean", "+1 SD")
gs_vals_centered <- round(gs_vals_centered, 3)
gs_vals_centered

# Simple slopes of IQ at these group size values
avg_slopes(m_cross,
           variables = "iq_verb",
           by = "groupsiz_c",
           newdata = datagrid(groupsiz_c = gs_vals_centered))

```


**Interpretation:** The effect of IQ on language scores is 2.95 in smaller schools (group size = -7.3, i.e., one SD below the mean), 2.75 in average‑sized schools, and 2.54 in larger schools (group size = +7.3). The decreasing pattern is consistent with the negative interaction term.

Plotting:

```{r}
plot_predictions(m_cross,
                 by = c("iq_verb", "groupsiz_c"),
                 newdata = datagrid(iq_verb = unique(SB$iq_verb),
                                    groupsiz_c = gs_vals_centered))
```

This plot shows three regression lines (IQ on language scores) for different levels of group size, allowing visual comparison of slopes. The lines clearly diverge, with the slope decreasing as group size increases.

## Model tables

We can present the model parameters in a table next to each other

```{r}
tab_model(m0_cross, m_cross)
```


# Example with a Binary Moderator

Cross-level interactions are particularly common with binary level-2 moderators, such as experimental condition or school type. Using the `popular` dataset (Hox, 2010), we can test whether the effect of extraversion on popularity depends on teacher experience (continuous) or sex (binary).

As always we start with the main effects model

```{r}
# Model with sex as a binary moderator of extraversion effect
m0_binary <- lmer(popular ~ 1 + extrav + girl + texp  + (1 + extrav | class),
                 data = pop, REML = FALSE)

summary(m0_binary)
parameters(m0_binary, ci_method = "profile")
```

**Interpretation:** Extraversion (0.453, $p < .001$), being a girl (1.252, $p < .001$), and teacher experience (0.091, $p < .001$) all positively predict popularity. The random intercept variance (1.281) and random slope variance for extraversion (0.034) indicate that both the intercept and the effect of extraversion vary across classes.

And then we can add the interaction term

```{r binary-moderator}
# Model with sex as a binary moderator of extraversion effect
m_binary <- lmer(popular ~ 1 + extrav * girl + texp  + (1 + extrav | class),
                 data = pop, REML = FALSE)

summary(m_binary)
parameters(m_binary, ci_method = "profile")
```

**Interpretation**:
- `extrav` is the effect of extraversion for boys (`girl = 0`): 0.427, $p < .001$.
- `extrav:girl` is the difference in the extraversion slope for girls compared to boys: 0.050, $p = 0.103$, which is not significant. Therefore, sex does not significantly moderate the effect of extraversion on popularity.

```{r}
anova(m0_binary, m_binary)
```


Probing a binary moderator is simpler: we only have two simple slopes (for each group).

```{r binary-probe}
avg_slopes(m_binary,
           variables = "extrav",
           by = "girl")
```

**Interpretation**: For boys, the effect of extraversion on popularity is 0.443; for girls, it's 0.462. The difference (0.019) is the interaction term (0.050 in the model output; the slight discrepancy is due to rounding). The p‑value of the interaction was 0.103, so this difference is not statistically significant.

Plot:

```{r binary-plot}
plot_predictions(m_binary, 
                 condition = c("extrav", "girl"))
```

The plot shows nearly parallel lines for boys and girls, consistent with the non‑significant interaction.

Model tables

We can present the model parameters in a table next to each other

```{r}
tab_model(m0_binary, m_binary)
```


# Three-Way Interactions

MLM can also accommodate higher-order interactions, such as a level-1 × level-1 × level-2 interaction. For example, the interaction between IQ and homework might itself be moderated by group size.

As always we should start with the main effects model

```{r}
m0_3way <- lmer(langpost ~ 1 + iq_verb + homework + groupsiz_c + 
                 (1 + iq_verb | schoolnr),
               data = SB, REML = FALSE)

summary(m0_3way)
parameters(m0_3way, ci_method = "profile")
```

Then we can add the 3-way interaction terms

```{r three-way}
m_3way <- lmer(langpost ~ 1 + iq_verb * homework * groupsiz_c + 
                 (1 + iq_verb | schoolnr),
               data = SB, REML = FALSE)

summary(m_3way)
parameters(m_3way, ci_method = "profile")
```

**Interpretation**: The three-way term `iq_verb:homework:groupsiz_c` has a coefficient of 0.002 ($p = 0.660$), which is not significant. This means that the IQ × homework interaction does not vary significantly across levels of group size.

Now, the regression terms test specific 2-way or 3-way interaction terms. While for an overall test is better to do the model comparison LRT test

```{r}
anova(m0_3way, m_3way)
```

The likelihood‑ratio test comparing the main‑effects model to the full three‑way model yields a p‑value of 0.0605, which is marginally non‑significant, suggesting that adding the three‑way interaction and its lower‑order terms does not significantly improve the model fit.

## Probing Three-Way Interactions

Probing a three-way interaction is more complex: we examine the two-way interaction at different levels of the third variable. For example, we could probe the IQ × homework interaction at low, medium, and high group size.

First, get values for homework:

```{r three-way-values}
# Values for homework
mean_hw <- mean(SB$homework, na.rm = TRUE)
sd_hw <- sd(SB$homework, na.rm = TRUE)
hw_vals <- c(mean_hw - sd_hw, mean_hw, mean_hw + sd_hw)
names(hw_vals) <- c("-1 SD", "Mean", "+1 SD")
hw_vals <- round(hw_vals, 3)
hw_vals
```

Now compute simple slopes of IQ at different homework and group size combinations:

```{r three-way-probe}
avg_slopes(m_3way,
           variables = "iq_verb",
           by = c("homework", "groupsiz_c"),
           newdata = datagrid(homework = hw_vals,
                              groupsiz_c = gs_vals_centered))
```

This gives the simple slope of IQ at each combination of homework (three levels) and group size (three levels) – nine slopes in total. We can then see if the pattern of IQ slopes across homework levels differs between group size conditions. For example, at low group size (-7.3), the IQ slopes are 2.93, 2.94, and 2.96 across low, mean, and high homework, showing a slight increase. At high group size (7.3), the slopes are 2.45, 2.53, and 2.61, also increasing but with a smaller range. However, the three‑way interaction was not significant, so these differences are not reliable.

## Plotting Three-Way Interactions

Plotting a three-way interaction can be done with separate panels:

```{r three-way-plot}
plot_predictions(m_3way,
                 by = c("iq_verb", "homework", "groupsiz_c"),
                 newdata = datagrid(groupsiz_c = gs_vals_centered,
                                    homework = hw_vals,
                                    iq_verb = unique(SB$iq_verb)))
```

This produces three plots (one per group size level) showing the IQ × homework interaction. The plots allow visual inspection of whether the pattern of lines changes across group size levels. In this case, the lines appear similar across panels, consistent with the non‑significant three‑way term.

## Model tables

We can present the model parameters in a table next to each other

```{r}
tab_model(m0_3way, m_3way)
```


# Effect Sizes for Interactions

Several approaches exist for quantifying the importance of interaction effects in MLM (Rights & Sterba, 2019).

## Pseudo-$R^2$ Change

We can compute the reduction in variance components when adding the interaction:

```{r pseudo-r2}
# Model without interaction
m_main <- lmer(langpost ~ 1 + iq_verb + groupsiz_c + (1 + iq_verb | schoolnr),
               data = SB, REML = FALSE)

# Model with interaction (already fit)
# Compare variance components
VarCorr(m_main)
VarCorr(m_cross)

# Proportion reduction in slope variance
tau11_main <- as.data.frame(VarCorr(m_main)) %>% 
  filter(grp == "schoolnr" & var1 == "iq_verb") %>% 
  pull(vcov)
tau11_cross <- as.data.frame(VarCorr(m_cross)) %>% 
  filter(grp == "schoolnr" & var1 == "iq_verb") %>% 
  pull(vcov)

(tau11_main - tau11_cross) / tau11_main
```

This tells us how much of the slope variance across schools is explained by group size. Here, about 18% of the variance in the IQ slope is accounted for by group size.

## Using the `performance` Package

```{r}
r2(m_main)
r2(m_cross)
```


## Using the `r2mlm` Package

The `r2mlm` package provides comprehensive $R^2$ measures for MLM, partitioning variance into fixed and random components at each level.

```{r r2mlm}
r2mlm(m_main)
r2mlm(m_cross)
```

The output shows total, within-cluster, and between-cluster variance explained, as well as decomposition into fixed (`f1`, `f2`), random slope (`v`), and random intercept (`m`) components. In the interaction model, the fixed‑effect variance explained (`f`) increases slightly, while the random slope variance (`v`) decreases, indicating that group size helps explain why the IQ slope varies across schools.

# Summary and Recommendations

| Interaction Type | Specification | Probing | Key Considerations |
|:---|:---|:---|:---|
| **Level-1 × Level-1** | Include product term in level-1 model | Simple slopes of one predictor at levels of the other | Center predictors for interpretability; consider random slopes for either predictor |
| **Level-2 × Level-2** | Include product term in level-2 intercept equation | Simple slopes at levels of the moderator | Center level-2 predictors; ensure sufficient level-2 sample size |
| **Cross-level (L1 × L2)** | Include L2 predictor in L2 slope equation | Simple slopes of L1 predictor at levels of L2 moderator | Must have random slope of L1 predictor to test; L2 sample size critical for power |

**General Recommendations** (Aguinis et al., 2013; Enders & Tofighi, 2007):

1. **Center predictors** appropriately – group-mean centering for level-1 predictors when testing cross-level interactions, grand-mean centering for level-2 predictors.
2. **Include random slopes** for level-1 predictors involved in cross-level interactions to avoid biased standard errors.
3. **Check statistical power** – cross-level interactions require adequate level-2 sample size (often >50). Use tools like the power calculator from Mathieu et al. (2012).
4. **Graph interactions** to aid interpretation, but don't rely on graphs alone for conclusions about effect importance.
5. **Report complete results** including fixed effects, random effects (variances and covariances), and effect size measures.

# References

- Aguinis, H., Gottfredson, R. K., & Culpepper, S. A. (2013). Best-practice recommendations for estimating cross-level interaction effects using multilevel modeling. *Journal of Management, 39*(6), 1490-1528.
- Enders, C. K., & Tofighi, D. (2007). Centering predictor variables in cross-sectional multilevel models: A new look at an old issue. *Psychological Methods, 12*(2), 121-138.
- Hox, J. J., Moerbeek, M., & van de Schoot, R. (2018). *Multilevel analysis: Techniques and applications* (3rd ed.). Routledge.
- Mathieu, J. E., Aguinis, H., Culpepper, S. A., & Chen, G. (2012). Understanding and estimating the power to detect cross-level interaction effects in multilevel modeling. *Journal of Applied Psychology, 97*(5), 951-966.
- Preacher, K. J., Curran, P. J., & Bauer, D. J. (2006). Computational tools for probing interaction effects in multiple linear regression, multilevel modeling, and latent curve analysis. *Journal of Educational and Behavioral Statistics, 31*(4), 437-448.
- Rights, J. D., & Sterba, S. K. (2019). Quantifying explained variance in multilevel models: An integrative framework for defining R-squared measures. *Psychological Methods, 24*(3), 309-338.
- Snijders, T. A. B., & Bosker, R. J. (2012). *Multilevel analysis: An introduction to basic and advanced multilevel modeling* (2nd ed.). Sage.

