---
title: "Multilevel Regression: centering"
author: "Mauricio Garnier-Villarreal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: gfm
toc: true
toc-depth: 5
number-sections: true
code-copy: true
execute:
  echo: true
  warning: false
  eval: true
---



## Introduction to Longitudinal Data Analysis

In the previous tutorials of this course we used multilevel models (MLM) to handle **clustered** data, where observations were nested within higher-level units such as schools or classrooms. We covered random intercept and random slope models (16_MLM1), how centering separates within- and between-cluster effects (16_1_MLM_centering), and how to estimate, probe, and plot interactions at different levels (16_2_MLM_interactions).

This tutorial introduces a second, very common application of MLM: **longitudinal data**, where repeated observations are nested within individuals. This is the same statistical machinery we have already learned, but the "cluster" is now a person and the key predictor is **time**. With MLM, when discussing longitudinal data we are primarily interested in data with observations nested within individuals, where the individual is the level-2 variable. When we include time as a level-1 predictor in these models, the model becomes a **multilevel growth model** (also known as a latent growth curve model, and identical to those fit with SEM) (Singer & Willett, 2003).

The central idea is that each person has their own trajectory over time, described by an intercept (where they start) and a slope (how fast they change). MLM lets us estimate the **average** trajectory in the population and, at the same time, quantify how much people **differ** from that average trajectory. As Little (2024) puts it, growth curve models characterize between-person differences in growth by representing each individual's cross-time observations as a parsimonious model of the growth process.

## Rationales for Longitudinal Research

Baltes and Nesselroade (1979) outlined five main rationales for conducting longitudinal research. These rationales continue to motivate longitudinal designs today and help us understand *why* we invest in collecting repeated measures.

### Rationale 1: Intraindividual change (and stability)

- Direct identification of intraindividual change (and stability).
- Measuring the same individual (or entity) repeatedly allows researchers to identify if and how specific attributes of the individual changed (or remained the same) over time.
- Developmental (and other) theories of change often conceptualize change as either an **incremental** or a **transformational** process:
  - *Incremental*: change in magnitude (e.g., vocabulary growing steadily).
  - *Transformational*: change between discrete states (e.g., moving from non-reader to reader).

### Rationale 2: Interindividual differences in change

- Direct identification of interindividual differences (or similarity) in intraindividual change.
- Do different individuals change different amounts, or in different directions?
- Baltes and Nesselroade (1979) suggested that heterogeneity in change is the norm, given the "existence of diversity, multidirectionality, and large interindividual differences in developmental outcomes" (p. 24).

### Rationale 3: Interrelationships in behavioral change

- Analysis of interrelationships in behavioral change.
- "The examination of interrelationships in change among distinct behavioral classes is particularly important if a structural, holistic approach to development is taken" (Baltes & Nesselroade, 1979, p. 25).
- This holistic approach centers on the idea that changes in multiple constructs are expected to occur simultaneously and/or sequentially.

### Rationale 4: Causes of intraindividual change

- Analysis of causes (determinants) of intraindividual change.
- Explaining or accounting for the observed within-person change process.
- Identify the time-varying factors and/or mechanisms that impact and/or drive the within-person changes identified in Rationale 1.

### Rationale 5: Causes of interindividual differences in change

- Analysis of causes (determinants) of interindividual differences in intraindividual change.
- Given that individuals differ in how they change over time (Rationale 2), researchers are often interested in identifying the factors and/or mechanisms that can account for those between-person differences.

## Questions about Growth

When we model longitudinal data, we are usually trying to answer three broad questions (Singer & Willett, 2003):

1. **What is the shape of the mean trend over time?**
   - What is the average intercept for the entire group?
   - What is the average slope for the entire group?
2. **Is there significant between-person variability in the shape of the trend?**
   - Do two or more groups differ in intercept? In slope?
3. **What variables are systematically associated with change over time?**

These questions map directly onto the fixed effects (average trajectory) and random effects (variability around that trajectory) of a multilevel growth model.

## Modeling Frameworks

- Both **SEM** and **MLM** can estimate growth models (Singer & Willett, 2003).
- A large variety of models can be estimated with both frameworks.
- Some specific cases can only be estimated in one framework (e.g., certain latent variable or measurement-error corrections are more natural in SEM; certain nested data structures are more natural in MLM).
- In this tutorial we use the MLM framework, because it builds directly on everything we have done so far.

## The Meaning of *Time*

- Time *means* nothing by itself.
- Time is a *proxy* for something else, such as (Singer & Willett, 2003):
  - Cognitive development
  - Application of a new policy
  - Learning/experience
- This has important consequences: the choice of the time variable (and its origin) shapes how we interpret the intercept and the slope, as we will see later.

## Packages and Data

We will use the `simchild.csv` dataset, which contains measures of closeness and conflict with MOM and DAD over 6 time points (Grades 1 through 6). This is a simulated dataset used for demonstration, and its structure mirrors the panel datasets commonly used in longitudinal teaching.

Key variables:

- `id`: person identifier (the level-2 unit)
- `GRADE`: grade in school (1 to 6)
- `GRADE_C`: grade centered so that Grade 1 = 0 (i.e., `GRADE - 1`)
- `CLOSEDAD`: closeness with dad (the outcome in our first models)
- `CLOSEMOM`: closeness with mom (used later as a time-varying covariate)
- `CONFLMOM`: conflict with mom (used for the gender example)
- `SEX`: sex of the child (0 = girls, 1 = boys)

The data are in **long (tall) format**, with one row per person per time point. As we learned in 16_MLM1, this is the format required by most software. Little (2024) contrasts the tall format (one row per occasion, with a `time` column) with the wide format (one column per occasion), and notes that in the tall format the same variable is measured repeatedly for each individual.

```{r}
# Load required packages
library(lme4)
library(lmerTest)
library(rio)
library(parameters)
library(marginaleffects)
library(ggplot2)
library(tidyr)
library(psych)
library(performance)
library(r2mlm)
library(car)

# Read data
simchild <- import("simchild.csv", na.strings = "-999999")

# Inspect the data
summary(simchild)
head(simchild)
dim(simchild)
```

We can check the structure of the long-format data by looking at a few individuals:

```{r}
# Look at the first observations for a few individuals
head(simchild[simchild$id %in% unique(simchild$id)[1:3],
              c("id", "GRADE", "GRADE_C", "CLOSEDAD", "CLOSEMOM", "SEX")], 12)
```

To get a first sense of the data, we can plot the trajectories of a random sample of 20 individuals:

```{r}
#| fig-width: 8
#| fig-height: 5
# Sample 20 individuals for a readable plot
ids <- sample(unique(simchild$id), 20)
simchild_short <- subset(simchild, id %in% ids)

ggplot(data = simchild_short,
       mapping = aes(x = GRADE_C, y = CLOSEDAD,
                     colour = factor(id))) +
  geom_line() +
  labs(colour = "id",
       x = "Grade (centered)",
       y = "Closeness with Dad",
       title = "Individual trajectories over time")
```

The plot shows that individuals start at different levels (different intercepts) and change at different rates (different slopes). This is exactly the kind of heterogeneity that the multilevel growth model is designed to capture.

## The Multilevel Growth Model

### The Two-Level Growth Model

The basic growth model with a single predictor, *time*, is:

**Level 1 (within-person):**

$$
y_{ij} = \beta_{0j} + \beta_{1j}\,time_{ij} + e_{ij}, \qquad e_{ij} \sim N(0, \sigma^2)
$$

**Level 2 (between-person):**

$$
\beta_{0j} = \gamma_{00} + u_{0j}
$$

$$
\beta_{1j} = \gamma_{10} + u_{1j}
$$

with the random effects covariance matrix

$$
T = \begin{bmatrix}
\tau_{00} & \\
\tau_{10} & \tau_{11}
\end{bmatrix}
$$

where $\mathbf{u}_j = (u_{0j}, u_{1j})' \sim MVN(\mathbf{0}, \mathbf{T})$ (Singer & Willett, 2003). Little (2024) presents the same model in SEM notation, where the level-1 equation is defined by how the loadings of the intercept and slope factors are specified (typically fixed rather than freely estimated, so that the factors can be interpreted as an intercept and a slope).

Substituting the level-2 equations into the level-1 equation gives the combined (reduced-form) model:

$$
y_{ij} = \gamma_{00} + \gamma_{10}\,time_{ij} + u_{0j} + u_{1j}\,time_{ij} + e_{ij}
$$

### Parameter Interpretation

The parameters of the growth model have direct and intuitive interpretations (Singer & Willett, 2003):

| Parameter | Meaning |
|---|---|
| $\gamma_{00}$ | Average score on $y$ at time = 0 (average intercept) |
| $\gamma_{10}$ | Average increase in $y$ for each 1-unit increase in time (average slope) |
| $\tau_{00}$ | Variability in scores on $y$ at time = 0 (intercept variance) |
| $\tau_{11}$ | Variability in the slope of time (variability in individuals' change over time) |
| $\tau_{10}$ | Covariance between intercept and slope |
| $\sigma^2$ | Within-person residual variance |

The fixed effects ($\gamma$) describe the **average** trajectory; the random effects ($\tau$) describe **how much people differ** from that average trajectory.

### Alternative Models

Depending on the research question, we may want to constrain some variance components to zero. The following are all special cases of the full model, defined by which elements of $\mathbf{T}$ are free (Singer & Willett, 2003):

- **Fixed intercept, fixed slope:** $T = \begin{bmatrix} 0 & \\ 0 & 0 \end{bmatrix}$ — no between-person variability at all; this is just an OLS regression.
- **Random intercept, fixed slope:** $T = \begin{bmatrix} \tau_{00} & \\ 0 & 0 \end{bmatrix}$ — people differ in where they start, but change at the same rate.
- **Fixed intercept, random slope:** $T = \begin{bmatrix} 0 & \\ 0 & \tau_{11} \end{bmatrix}$ — people start at the same place, but change at different rates.
- **Random intercept, random slope:** $T = \begin{bmatrix} \tau_{00} & \\ \tau_{10} & \tau_{11} \end{bmatrix}$ — people differ in both starting point and rate of change (the full model).

Any of these models may be sensible depending on the situation. We can test the significance of any variance component (e.g., $\tau_{11}$) using the profile likelihood method or the deviance test for nested models. It is also easy to specify other level-1 predictors (time-varying covariates) or level-2 predictors of intercepts and slopes (time-invariant covariates).

## Choosing the Time Origin

The intercept is "located" where the origin (zero-point) of the time variable is situated. Therefore, **time = 0 needs to be meaningful**. Most commonly, time is centered at the first observation (Singer & Willett, 2003).

For example, if the time metric is a calendar year, it is best to center the year so that the origin is the first assessment:

```{r}
year <- c(1974, 1982, 1985, 1991)
time <- year - 1974

rbind(year, time)
```

The time metric then becomes "years since initial assessment."

Other scenarios for a meaningful origin might be (Singer & Willett, 2003):

- time since diagnosis
- time since birth
- developmental state
- time until death
- time before and after some important event

### Why the Origin Matters

Choosing the origin of time is important because it affects the interpretation of (Singer & Willett, 2003):

- the mean intercept ($\gamma_{00}$)
- the intercept variance ($\tau_{00}$)
- the intercept/slope covariance ($\tau_{10}$)

Two important notes:

- **Recentering** time will affect any parameter associated with the intercept (but not the slope).
- **Rescaling** time (e.g., converting from months to years) will affect any parameter associated with the slope (but not the intercept).

Crucially, the **fit of the model does not change** based on centering or rescaling; only the interpretation of the model parameters changes. Little (2024) devotes specific attention to the importance of coding time properly, noting that the coding of observations at different intervals of time affects the interpretation of growth models.

### The Aperture

The value of time that corresponds to the smallest dispersion of predicted scores (the smallest $\tau_{00}$) is called the **aperture**. The aperture also happens to be the place where the covariance between intercepts and slopes is 0 ($\tau_{10}$ flips from negative to positive) (Singer & Willett, 2003).

We can use this fact to solve for the aperture:

$$
aperture = a^* - \frac{\tau_{10}}{\tau_{11}}
$$

where $a^*$ is usually time = 0. The aperture might not fall in the range of observed data.

An example of where the concept of the aperture might be useful: suppose a researcher wishes to plan a study to collect data at just the time at which children begin to diverge on some characteristic. We might use data collected sometime after the suspected aperture point to make a good estimate of when the aperture occurs (Singer & Willett, 2003).

### Rescaling Time

We can also rescale time to change the metric of the slope. For example, dividing time by 10 changes the slope from "per grade" to "per decade of grades":

```{r}
#| eval: true
# Rescale time: slope per decade
simchild$dec <- simchild$GRADE_C / 10

m1e <- lmer(CLOSEDAD ~ 1 + dec + (1 + dec | id),
            data = simchild, REML = FALSE)
parameters(m1e, ci_method = "profile")

# Rescale time: slope per 10 grades (slope becomes 10x larger)
simchild$mon <- simchild$GRADE_C * 10

m1e1 <- lmer(CLOSEDAD ~ 1 + mon + (1 + mon | id),
             data = simchild, REML = FALSE)
parameters(m1e1, ci_method = "profile")
```

Notice how the slope estimate scales with the time metric, while the model fit (AIC, log-likelihood) stays identical.

## Alternative Metrics of Time

Say you want to measure growth in language ability in elementary school children. There are several options for what to use as the variable *time* (Singer & Willett, 2003):

- **wave**: reflexively chosen by most researchers (data organization habit)
- **grade**: perhaps reflects change due to "nurture"
- **age**: perhaps reflects change due to "nature"
- perhaps more than one of these?

Sometimes these variables are completely confounded (they yield the same information), and sometimes they are not. It is possible for children of the same age to be in different grades, and for children in the same wave of measurement to have different ages. **What is chosen as the time metric can have profound consequences for interpretation**, so decide what theory suggests is the variable truly responsible for change (Biesanz et al., 2004).

### Cohort-Sequential Designs

It is sometimes possible to obtain information for trajectories across many occasions (ages) by measuring on fewer occasions (waves) (Singer & Willett, 2003).

A **cohort** is a group of individuals followed longitudinally, but starting at different times. For example, we might follow a cohort of 5-year-olds for 4 years (yielding data for ages 5–8) and a cohort of 7-year-olds for those same 4 years (yielding data for ages 7–10). In only 4 years, we have collected data spanning ages 5–10.

**Cohort-sequential designs** (or accelerated longitudinal designs) are data collection strategies that pursue this strategy to save costs in terms of time and money. "Gluing together" cohort trajectories has benefits, but it also involves assumptions — fortunately, some of those assumptions are testable.

Cohorts might differ due to **cohort effects** (or history effects): differences between groups of people born in different eras. For example, individuals born in the 1950s may perform worse on a reaction-time test than those born in the 1970s not because they are older, but because they did not grow up using computers. In this example, cohort is causing the effect, not age (Duncan & Duncan, 2004).

## Building Growth Models: A Sequential Approach

As in 16_MLM1, we follow a sequential approach (Peugh, 2010): start with the simplest model, then build up by adding random effects and predictors, comparing nested models along the way. For all model comparisons with likelihood ratio tests we use **ML** estimation (`REML = FALSE`).

### Fixed Intercept, Fixed Slope (OLS)

The simplest model is an OLS regression, which assumes no between-person variability in intercepts or slopes. We fit it with `lm()` because `lmer()` cannot estimate a model with no random effects:

```{r}
m1 <- lm(CLOSEDAD ~ 1 + GRADE_C, data = simchild)
parameters(m1)
```

This gives us the average trajectory, but it ignores the clustering of observations within persons, so the standard errors are too small and we cannot describe individual differences.

### Null Model (Random Intercept Only)

The null model includes a random intercept for persons but no predictor:

```{r}
m0a <- lmer(CLOSEDAD ~ 1 + (1 | id),
            data = simchild, REML = FALSE)
parameters(m0a, ci_method = "profile")
```

We can compute the intraclass correlation (ICC), which here represents the proportion of variance that is between persons:

```{r}
# ICC: proportion of variance between persons
3.18^2 / (3.18^2 + 2.8^2)
```

An ICC of about 0.56 means that roughly 56% of the variability in closeness with dad is **between** persons, and 44% is **within** persons across time. This substantial between-person variance justifies a multilevel model.

### Random Intercept, Fixed Slope

Now we add time as a fixed predictor, keeping the random intercept:

```{r}
m1a <- lmer(CLOSEDAD ~ 1 + GRADE_C + (1 | id),
            data = simchild, REML = FALSE)
parameters(m1a, ci_method = "profile")
```

The fixed slope for `GRADE_C` is the **average** change in closeness per grade. The random intercept variance ($\tau_{00}$) tells us how much persons differ in their starting level. Because the slope is fixed, this model assumes everyone changes at the same rate.

### Fixed Intercept, Random Slope

We can also allow the slope to vary while fixing the intercept. In `lmer()` syntax, `(0 + GRADE_C | id)` removes the random intercept and keeps only the random slope:

```{r}
m1b <- lmer(CLOSEDAD ~ 1 + GRADE_C + (0 + GRADE_C | id),
            data = simchild, REML = FALSE)
parameters(m1b, ci_method = "profile")
```

This model assumes everyone starts at the same level but changes at different rates. It is rarely the best substantive choice, but it is useful as a comparison point.

### Random Intercept, Random Slope (Full Model)

The full growth model allows both the intercept and the slope to vary across persons:

```{r}
m1c <- lmer(CLOSEDAD ~ 1 + GRADE_C + (1 + GRADE_C | id),
            data = simchild, REML = FALSE)
parameters(m1c, ci_method = "profile")
```

Interpretation:

- **Fixed intercept** ($\gamma_{00}$): the average closeness with dad at Grade 1 (when `GRADE_C = 0`).
- **Fixed slope** ($\gamma_{10}$): the average change in closeness per grade.
- **Random intercept variance** ($\tau_{00}$): variability in starting levels.
- **Random slope variance** ($\tau_{11}$): variability in rates of change.
- **Intercept–slope correlation** ($\rho_{01} = 0.47$): persons who start higher tend to increase more (or, with a negative correlation, those who start higher change more slowly).


### Expanded interpretation: `m1c` (full growth model)

**Fixed effects (the average trajectory).**

| Parameter | Estimate | SE | 95% CI | t(3695) | p |
|---|---|---|---|---|---|
| Intercept ($\gamma_{00}$) | 36.26 | 0.12 | [36.03, 36.49] | 313.72 | < .001 |
| GRADE_C ($\gamma_{10}$) | -0.52 | 0.03 | [-0.58, -0.47] | -18.30 | < .001 |

- **Intercept ($\gamma_{00} = 36.26$).** This is the predicted closeness with dad at Grade 1 (`GRADE_C = 0`), averaged over all persons. The very small standard error (0.12) and narrow 95% CI [36.03, 36.49] indicate that the average starting level is estimated precisely. On the observed range of `CLOSEDAD` (13–40), 36.26 sits toward the upper end, so on average children report fairly high closeness with their fathers at the start of the study.
- **Time slope ($\gamma_{10} = -0.52$).** This is the average change in closeness with dad per additional grade. The negative sign means closeness declines, on average, by about half a point per grade. Across the full 5-grade span (Grade 1 → Grade 6) the model predicts a total decline of $-0.52 \times 5 = -2.60$ points — roughly 7% of the starting level, or about 1 residual SD (2.60 / 2.53 = 1.03). The effect is highly significant ($t = -18.30$, $p < .001$) and the CI excludes zero comfortably.

**Random effects (between-person heterogeneity).**

| Parameter | Estimate (SD) | Implied variance |
|---|---|---|
| SD (Intercept: id) | 2.58 | 6.66 |
| SD (GRADE_C: id) | 0.39 | 0.15 |
| Cor (Intercept ~ GRADE_C: id) | 0.47 | cov $= 0.47 \times 2.58 \times 0.39 = 0.47$ |
| SD (Residual) | 2.53 | 6.40 |

- **Random intercept variance ($\tau_{00} = 2.58^2 = 6.66$).** Children differ substantially in where they start. Assuming approximate normality, about 95% of children have Grade 1 closeness within $\pm 1.96 \times 2.58 = \pm 5.06$ points of the average, i.e., roughly between 31.2 and 41.3.
- **Random slope variance ($\tau_{11} = 0.39^2 = 0.15$).** The rate of change also differs across children. The implied 95% range of slopes is $-0.52 \pm 1.96 \times 0.39 = [-1.28, 0.24]$. This is substantively important: although the *average* trajectory declines, some children decline far more steeply (slopes near −1.3) while others are essentially flat or even slightly increasing (slopes near +0.2). A single "average decline" does not describe every child.
- **Intercept–slope correlation ($\rho_{01} = 0.47$).** Because the average slope is negative, a positive correlation means that children who start *higher* decline *more slowly* (slopes closer to zero), while children who start *lower* decline *more steeply*. The trajectories therefore **fan out** over time: initial differences between children widen as they move through the grades.
- **Residual variance ($\sigma^2 = 2.53^2 = 6.40$).** After accounting for each child's own intercept and slope, individual observations still deviate from that child's line by about 2.53 points. This is the within-person, occasion-specific variation (measurement error plus short-term fluctuation).

**Variance partitioning at Grade 1.** Using only the intercept and residual variances, the between-person share at Grade 1 is $6.66 / (6.66 + 6.40) = 0.51$ — about 51% of the variance in Grade 1 closeness is between children and 49% is within children. This is close to the ICC of 0.56 from the null model and confirms that a substantial share of the variation is between persons, which is exactly what the random effects capture.

**Substantive summary.** The full model says three things at once: (1) on average, closeness with dad declines over the elementary grades; (2) children differ both in starting level and in how quickly they change, so the decline is not uniform; and (3) starting level and rate of change are positively related, so higher-starting children decline more slowly. This is a far richer picture than the single average slope from the OLS model.



### Model Comparison

We can test whether the random slope is needed using a likelihood ratio test. Nested models with the same fixed effects are compared with `anova()`:

```{r}
anova(m1a, m1c)
anova(m1b, m1c)
```

The comparison of `m1a` (fixed slope) and `m1c` (random slope) tests whether the slope varies significantly across persons. The comparison of `m1b` (fixed intercept) and `m1c` tests whether the intercept varies. A significant chi-square indicates that the additional random effect improves model fit.

For confidence intervals on the random effects, we use **profile likelihood** intervals, which are more accurate than Wald intervals in MLM (Hox et al., 2018; this may take a while to compute):

```{r}
#| eval: false
parameters(m1c, ci_method = "profile")
parameters(m1c, ci_method = "profile", ci = .90)
```


### Visualizing Random Effects

We can extract and plot the random effects (BLUPs) with `ranef()` and `ggplot2`, just as we did for the school random effects in 16_MLM1:

```{r}
#| fig-width: 9
#| fig-height: 5
randE <- ranef(m1c)
dd <- as.data.frame(randE)

ggplot(dd, aes(y = grp, x = condval)) +
  geom_point() +
  facet_wrap(~ term, scales = "free_x") +
  geom_errorbar(aes(xmin = condval - 2 * condsd,
                    xmax = condval + 2 * condsd),
                width = 0,
                orientation = "y")
```

This caterpillar plot shows each person's random intercept deviation (left panel) and random slope deviation (right panel). Points far from zero indicate persons whose trajectory differs substantially from the average. With many persons the plot becomes crowded, so we can subset to a random sample of 50:

```{r}
#| fig-width: 9
#| fig-height: 5
ids <- sample(dd$grp, 50)
dd2 <- subset(dd, grp %in% ids)

ggplot(dd2, aes(y = grp, x = condval)) +
  geom_point() +
  facet_wrap(~ term, scales = "free_x") +
  geom_errorbar(aes(xmin = condval - 2 * condsd,
                    xmax = condval + 2 * condsd),
                width = 0,
                orientation = "y")
```

#### Visualizing random effects with marginaleffects

We can also visualize the random effects directly as predicted trajectories, using the `marginaleffects` package. At the **population level**, `plot_predictions()` averages over the random effects and shows the average growth trajectory with its confidence band:

```{r}
#| fig-width: 8
#| fig-height: 5
plot_predictions(m1c, condition = "GRADE_C")
```

At the **unit level**, we generate predictions for each person (which include their random effects) and plot one line per person. This is the trajectory-based analogue of the caterpillar plot:

```{r}
#| fig-width: 8
#| fig-height: 5
# Unit-level predictions for a random sample of 20 persons
preds_re <- predictions(m1c,
                        newdata = datagrid(id = sample(m1c@frame$id, 20),
                                           GRADE_C = 0:5))

ggplot(preds_re, aes(GRADE_C, estimate, group = id)) +
  geom_line() +
  labs(y = "Predicted Closeness with Dad",
       x = "Grade",
       title = "Individual growth trajectories")
```

The caterpillar plot emphasizes the **distribution** of the random intercepts and slopes (how spread out persons are, and which persons are outliers), while the `marginaleffects` trajectory plot emphasizes the **implied growth lines** for individual persons. Together they give a complete picture of the between-person heterogeneity in the model.

### Using the Aperture to Recenter Time

Recall that the aperture is the time point where the intercept–slope covariance is zero. For `m1c` the estimated random effects give an intercept SD of 2.5775, a slope SD of 0.3904, and an intercept–slope correlation of 0.47. We convert the correlation to a covariance using `psych::cor2cov()`:

```{r}
# Convert the correlation matrix to a covariance matrix
tauCor <- matrix(c(1, .47, .47, 1), byrow = TRUE, nrow = 2)

tauCov <- cor2cov(tauCor, c(2.5775, 0.3904))

# Covariance of intercept and slope (check by hand)
.47 * 2.5775 * .3904

# Aperture: time point where covariance is zero
-tauCov[1, 2] / tauCov[2, 2]

# Equivalent by hand
-(0.4729403 / 0.1524)
```

The aperture is about $-3.1$ grades relative to the current origin (`GRADE_C = 0`), which falls **outside** the observed range (grades 0–5). We can nonetheless recenter time at the aperture and refit the model:

```{r}
#| eval: true
# Create a new time variable with the origin at the aperture
simchild$NEW_GRADE <- simchild$GRADE_C - (-tauCov[1, 2] / tauCov[2, 2])

m1cAP <- lmer(CLOSEDAD ~ 1 + NEW_GRADE + (1 + NEW_GRADE | id),
              data = simchild, REML = FALSE)
parameters(m1cAP, ci_method = "profile")
```

When the origin is placed at the aperture, the estimated intercept–slope correlation becomes (approximately) zero, and the intercept variance is at its minimum. This can be useful when planning a study, but remember that the aperture may not correspond to a meaningful or observed point in time.

## Adding Predictors

In growth models, the intercept or slope can be predicted by level-2 predictors, or the outcome can be predicted by level-1 predictors. This distinction is often described as **time-invariant** (level-2) and **time-varying** (level-1) predictors (Singer & Willett, 2003).

- A **level-2 predictor** of the intercept or slope tells you how intercepts or slopes differ as a function of the subject.
- A **level-1 predictor** of the outcome tells you how the outcome is changing over time, *controlling* for the level-1 predictor.

Remember that the trajectory (intercept and slope) is interpreted where all covariates (other than time) equal 0. Therefore, level-1 predictors need to be centered in a meaningful way, especially if interactions are involved (Singer & Willett, 2003).

### Time-Invariant Predictors (Predicting Intercepts and Slopes)

When a variable predicts the slope, the model includes an interaction term between time and that predictor. The general model is (Singer & Willett, 2003):

$$
\begin{array}{l}
y_{ij} = \beta_{0j} + \beta_{1j}\,time_{ij} + e_{ij} \\
\beta_{0j} = \gamma_{00} + \gamma_{01}w_{j} + u_{0j} \\
\beta_{1j} = \gamma_{10} + \gamma_{11}w_{j} + u_{1j} \\
\overline{y_{ij} = \gamma_{00} + \gamma_{10}time_{ij} + \gamma_{01}w_{j} + u_{0j} + u_{1j}time_{ij} + \gamma_{11}time_{ij}w_{j} + e_{ij}}
\end{array}
$$

Some of the most interesting effects occur when variables predict **slopes** (i.e., when $\gamma_{11} \neq 0$).

### Example: Does Sex Predict Growth in Conflict with Mom?

We use `CONFLMOM` (conflict with mom) as the outcome and `SEX` as a level-2 predictor. First, recode sex into a labeled factor (boys = 1, girls = 0):

```{r}
simchild$SEX_cat <- car::recode(simchild$SEX,
                                " 0 = 'girls'; 1 = 'boys' ")

# Baseline growth model (no predictor)
m1d0 <- lmer(CONFLMOM ~ 1 + GRADE_C + (1 + GRADE_C | id),
             data = simchild, REML = FALSE)
parameters(m1d0, ci_method = "profile")
```

Next, we add `SEX_cat` as a predictor of the **intercept only** (main effect, no interaction):

```{r}
m1d <- lmer(CONFLMOM ~ 1 + GRADE_C + SEX_cat + (1 + GRADE_C | id),
            data = simchild, REML = FALSE)
parameters(m1d, ci_method = "profile")
```

### Expanded interpretation: `m1d` (conflict with mom, sex as a predictor of the intercept)

This model uses `CONFLMOM` (conflict with mom) as the outcome and adds `SEX_cat` as a time-invariant (level-2) predictor of the intercept only, keeping the random intercept and random slope for time.

| Parameter | Estimate | SE | 95% CI | t(5094) | p |
|---|---|---|---|---|---|
| Intercept ($\gamma_{00}$) | 15.09 | 0.23 | [14.63, 15.54] | 65.03 | < .001 |
| GRADE_C ($\gamma_{10}$) | 0.29 | 0.03 | [0.22, 0.35] | 8.83 | < .001 |
| SEX_cat [girls] ($\gamma_{01}$) | 0.38 | 0.31 | [-0.23, 0.99] | 1.23 | 0.220 |

**Fixed effects.**
- **Intercept ($\gamma_{00} = 15.09$).** With `SEX_cat` in the model, the intercept is the predicted conflict at Grade 1 for the **reference group**. Because the factor levels are ordered alphabetically (`boys`, `girls`), the reference is **boys** — which is why the coefficient row is labelled `[girls]`. So boys are predicted to have a conflict score of about 15.1 at Grade 1.
- **Time slope ($\gamma_{10} = 0.29$).** This is the average change in conflict per grade, assumed equal for boys and girls in this main-effects model. Note the direction is **positive**: unlike closeness, conflict with mom *increases* over the grades, by about 0.29 points per grade, or $0.29 \times 5 = 1.45$ points across the study. The effect is highly significant.
- **Sex effect ($\gamma_{01} = 0.38$).** This is the difference in Grade 1 conflict between girls and boys (girls − boys). Girls are predicted to score about 0.38 points higher, but the effect is not significant ($p = 0.220$) and the 95% CI [-0.23, 0.99] is wide and includes zero. There is no evidence of a sex difference in *initial* conflict with mom.

**Random effects.**

| Parameter | Estimate (SD) |
|---|---|
| SD (Intercept: id) | 5.03 |
| SD (GRADE_C: id) | 0.60 |
| Cor (Intercept ~ GRADE_C: id) | -0.20 |
| SD (Residual) | 3.22 |

- **Random intercept ($5.03^2 = 25.3$).** Children differ enormously in Grade 1 conflict: 95% fall roughly $15.09 \pm 1.96 \times 5.03 = [5.2, 25.0]$. This between-person variance is much larger than for closeness with dad, reflecting that conflict is a more variable construct.
- **Random slope ($0.60^2 = 0.36$).** The rate of change in conflict also varies widely, with 95% of slopes roughly $0.29 \pm 1.96 \times 0.60 = [-0.89, 1.47]$. Some children's conflict is decreasing while others' is increasing steeply.
- **Intercept–slope correlation ($\rho_{01} = -0.20$).** A weak negative correlation: children who start with higher conflict tend to increase slightly less over time. Because the mean slope is positive, this implies mild convergence rather than fanning out.
- **Residual SD ($3.22^2 = 10.4$).** Considerable occasion-to-occasion variation remains within children.

**Substantive summary.** Conflict with mom behaves quite differently from closeness with dad: it *increases* over the elementary grades, and there are large between-person differences in both level and change. The main-effects model finds no overall sex difference in initial conflict; whether the *rate of change* differs by sex is tested in the interaction model (`m1f`), whose simple slopes are reported next.


Finally, we add the **time × sex interaction**, which lets sex predict the **slope** as well:

```{r}
m1f <- lmer(CONFLMOM ~ 1 + GRADE_C + SEX_cat + GRADE_C:SEX_cat +
              (1 + GRADE_C | id),
            data = simchild, REML = FALSE)
parameters(m1f, ci_method = "profile")
```

Interpretation:

- The `GRADE_C` main effect is now the slope for the reference group (girls).
- The `GRADE_C:SEX_cat` interaction tells us whether boys and girls differ in their rate of change over time.
- The `SEX_cat` main effect tells us whether boys and girls differ in conflict at Grade 1 (time = 0).

### Probing the Interaction

As we learned in 16_2_MLM_interactions, we probe an interaction by examining simple slopes and plotting them. We use the `marginaleffects` package for this, exactly as in the previous tutorial.

First, we compute the simple slopes of time (`GRADE_C`) at each level of the moderator (`SEX_cat`):

```{r}
avg_slopes(m1f, variables = "GRADE_C", by = "SEX_cat")
```


### Expanded interpretation: `avg_slopes()` for the time × sex interaction (`m1f`)

The interaction model (`m1f`) adds `GRADE_C:SEX_cat`, allowing the rate of change in conflict to differ for boys and girls. `avg_slopes()` translates the interaction back into the simple slope of time for each group:

| Sex | Simple slope of time | SE | 95% CI |
|---|---|---|---|
| boys | 0.224 | 0.046 | [0.134, 0.314] |
| girls | 0.351 | 0.046 | [0.261, 0.441] |

- **Both simple slopes are positive and significant.** Conflict with mom increases over the grades for boys ($0.224$, $p < .001$) and for girls ($0.351$, $p < .001$). For boys the increase is about 0.22 points per grade ($\approx 1.1$ points over five grades); for girls it is about 0.35 points per grade ($\approx 1.8$ points over five grades).
- **The difference between the slopes is the interaction term.** Girls' slope minus boys' slope is $0.351 - 0.224 = 0.127$, which matches the interaction coefficient in `m1f` (0.12). So the interaction says: relative to boys, girls' conflict increases by an *additional* 0.13 points per grade.
- **But the difference is only marginal.** In the parameter table of `m1f` the interaction had $p = 0.055$ (95% CI [0.00, 0.25]). The `avg_slopes()` output reports the two simple slopes but does not test their difference directly; the marginal $p$-value of the interaction term is the relevant test, and it sits just above the conventional .05 threshold. The honest conclusion is that there is a *trend* toward girls showing a steeper increase in conflict than boys, but the evidence is not conclusive.
- **Reporting tip.** Reporting simple slopes for both groups is good practice because it shows the *magnitude and direction* of the effect in each group, not just whether the interaction is significant. Even a non-significant interaction is better described as "the slope was X for one group and Y for the other" than left as a single interaction coefficient.
- **Why the main-effects and interaction models differ.** In `m1d` (no interaction) both groups were forced to share the single slope of 0.29. Once the interaction is allowed, that shared slope splits into 0.22 for boys and 0.35 for girls. This illustrates how adding an interaction reinterprets the "main effect" of time: it becomes the slope for the reference group rather than a single slope for everyone.


This gives the rate of change in conflict separately for girls and boys, along with standard errors and confidence intervals. We can then plot the two trajectories:

```{r}
#| fig-width: 8
#| fig-height: 5
plot_predictions(m1f, condition = c("GRADE_C", "SEX_cat"))
```

The plot shows the predicted conflict trajectory for girls and boys. If the lines are not parallel, the interaction is present. This is exactly the same probing logic as in cross-level interactions (16_2_MLM_interactions), only now the level-1 predictor is *time*.

### Time-Varying Covariates

A **time-varying covariate** is a level-1 predictor that changes within persons over time. Here we examine change in `CLOSEDAD` controlling for `CLOSEMOM`. As always, the time-varying covariate should be centered so the intercept is interpretable (Singer & Willett, 2003).

```{r}
# Model with CLOSEMOM as a time-varying covariate
m1h <- lmer(CLOSEDAD ~ 1 + GRADE_C + CLOSEMOM + (1 + GRADE_C | id),
            data = simchild, REML = FALSE)
parameters(m1h, ci_method = "profile")
```

We can also test whether the effect of `CLOSEMOM` (its slope) differs across individuals by adding it as a random slope:

```{r}
m1h1 <- lmer(CLOSEDAD ~ 1 + GRADE_C + CLOSEMOM +
               (1 + GRADE_C + CLOSEMOM | id),
             data = simchild, REML = FALSE)
parameters(m1h1, ci_method = "profile")
anova(m1h, m1h1)
```

Now we center `CLOSEMOM` at its grand mean (36.97) so that the intercept is interpretable:

```{r}
simchild$CLOSEMOM_C <- simchild$CLOSEMOM - 36.97

m1hc <- lmer(CLOSEDAD ~ 1 + GRADE_C + CLOSEMOM_C + (1 + GRADE_C | id),
             data = simchild, REML = FALSE)
parameters(m1hc, ci_method = "profile")
```

### Expanded interpretation: `m1hc` (closeness with dad controlling for closeness with mom)

This model adds a **time-varying covariate**, `CLOSEMOM_C` (closeness with mom, centered at its grand mean of 36.97), to the full growth model. Because it is centered, the intercept is now interpretable at the average level of mom-closeness.

| Parameter | Estimate | SE | 95% CI | t(3655) | p |
|---|---|---|---|---|---|
| Intercept ($\gamma_{00}$) | 36.09 | 0.12 | [35.86, 36.32] | 307.87 | < .001 |
| GRADE_C ($\gamma_{10}$) | -0.46 | 0.03 | [-0.52, -0.41] | -15.63 | < .001 |
| CLOSEMOM_C ($\gamma_{20}$) | 0.17 | 0.02 | [0.13, 0.22] | 7.60 | < .001 |

**Fixed effects.**
- **Intercept ($\gamma_{00} = 36.09$).** The predicted closeness with dad at Grade 1 for a child at the *average* level of closeness with mom. Centering `CLOSEMOM` makes this value meaningful; uncentered, the intercept would be the predicted dad-closeness when mom-closeness is zero, which lies outside the observed range.
- **Time slope ($\gamma_{10} = -0.46$).** After controlling for closeness with mom, the average decline in dad-closeness is 0.46 points per grade. Compare this with the unconditional estimate in `m1c` (−0.52): the decline is slightly shallower once mom-closeness is held constant, but the downward trend is unchanged.
- **CLOSEMOM_C effect ($\gamma_{20} = 0.17$).** This is a **within-person** effect: for a given child, when their closeness with mom is one point above their own average, their closeness with dad is expected to be 0.17 points higher, controlling for the passage of time. The effect is small but highly significant ($t = 7.60$, $p < .001$). Substantively, closeness with mom and dad move together *within* a child over time — on occasions when a child feels closer to mom, they also tend to feel closer to dad.

**Random effects.**

| Parameter | Estimate (SD) | Compare to `m1c` |
|---|---|---|
| SD (Intercept: id) | 2.50 | 2.58 |
| SD (GRADE_C: id) | 0.38 | 0.39 |
| Cor (Intercept ~ GRADE_C: id) | 0.47 | 0.47 |
| SD (Residual) | 2.53 | 2.53 |

- The random effects are almost identical to the unconditional model. In particular, the residual SD is unchanged at 2.53, meaning `CLOSEMOM_C` explains very little of the *occasion-level* residual variation in `CLOSEDAD`. This is an important (and sometimes counterintuitive) lesson: a significant within-person predictor does not necessarily reduce residual variance much when its effect is small relative to the total within-person variation. Note also that the number of observations drops (df from 3695 to 3655) because `CLOSEMOM` has some missing values, so the two models are fit to slightly different samples.
- **Practical implication.** If the goal were to explain within-person variation in dad-closeness, we would look for stronger time-varying predictors. The value of `CLOSEMOM_C` here is mainly as a statistical control that sharpens the estimate of the time trend and guards against confounding the trajectory with concurrent family dynamics.

**Substantive summary.** The time trend in dad-closeness is robust to controlling for mom-closeness, and there is a small but reliable within-person association between the two. This is a good example of a **time-varying covariate** doing two jobs at once: controlling for a competing process and testing a within-person hypothesis.


We can combine the time-varying covariate with a time-invariant predictor, and also test whether the time-varying covariate interacts with time:

```{r}
# Time-varying covariate plus sex
m1i <- lmer(CLOSEDAD ~ 1 + GRADE_C * SEX_cat + CLOSEMOM_C +
              (1 + GRADE_C | id),
            data = simchild, REML = FALSE)
parameters(m1i, ci_method = "profile")

# Interaction between time and the time-varying covariate
m1hcI <- lmer(CLOSEDAD ~ 1 + GRADE_C * CLOSEMOM_C + (1 + GRADE_C | id),
              data = simchild, REML = FALSE)
parameters(m1hcI, ci_method = "profile")
```

### Multiple Group Growth Models

When we add a categorical predictor at level 2 (person level), we can estimate a **multiple group growth curve**. This lets us ask: "Do both groups have the same growth pattern?" and "Do groups differ in only part of the growth pattern?" (Singer & Willett, 2003).

**Groups differ in both intercept and slope:**

$$
\begin{array}{l}
y_{ij} = \beta_{0j} + \beta_{1j}\,time_{ij} + e_{ij} \\
\beta_{0j} = \gamma_{00} + \gamma_{01}Group_{j} + u_{0j} \\
\beta_{1j} = \gamma_{10} + \gamma_{11}Group_{j} + u_{1j}
\end{array}
$$

**Groups differ only on the intercept:**

$$
\begin{array}{l}
\beta_{0j} = \gamma_{00} + \gamma_{01}Group_{j} + u_{0j} \\
\beta_{1j} = \gamma_{10} + u_{1j}
\end{array}
$$

**Groups differ only on the slope:**

$$
\begin{array}{l}
\beta_{0j} = \gamma_{00} + u_{0j} \\
\beta_{1j} = \gamma_{10} + \gamma_{11}Group_{j} + u_{1j}
\end{array}
$$

The same logic applies to **continuous** level-2 predictors (Singer & Willett, 2003):

- Predicting the intercept indicates differences at time 0.
- Predicting the slope indicates differences in the rate of change for a one-unit increase in the predictor.

### Interactions

When we have differences in the slope, it is an interaction between the *time* and *w* predictors, i.e., a **cross-level interaction** between a level-1 (time) and a level-2 (w) predictor. We can use simple slopes to probe and plot differences in the growth pattern — what changes relative to 16_2_MLM_interactions is the *interpretation*, as everything is now in function of the growth model (Singer & Willett, 2003).

## Centering in Growth Models

Centering predictors applies in the same way as in previous MLM models (Singer & Willett, 2003):

- Uncentered data
- Grand mean centering
- Conditional value centering
- **Group mean centering**
- **Group mean centering, + mean**

There is no universally correct choice for how (or whether) to center. As detailed in 16_1_MLM_centering, the decision depends on what we want to make predictions about (Enders & Tofighi, 2007):

| If we prefer to make predictions about $y_{ij}$ based on… | Then use… |
|---|---|
| The absolute level of $x_{ij}$ | raw $x_{ij}$ |
| Standing on $x$ relative to all other level-1 units | grand mean centered $x_{ij}$ |
| Standing on $x_{ij}$ relative to all other level-1 units within a group | group mean centered $x_{ij}$ |
| Disentangling level-1 and level-2 effects | group mean centered $x_{ij}$, plus the group mean as a level-2 predictor |

And, as always: does $x$ have a meaningful zero? If not, then centering may be advisable (Singer & Willett, 2003). In growth models, **time** is the most important variable to center, because the intercept is located at the time origin. Note that "group mean centering" time would center each person at their own mean time, which is unusual; typically we center time at a **common, meaningful value** (the first wave, or the aperture).

## Effect Size Measures

As we saw in 16_MLM1 and 16_1_MLM_centering, traditional $R^2$ is not directly applicable in MLM because variance is partitioned across levels. We use the same pseudo-$R^2$ measures as before.

### Using the `performance` Package

The `r2()` function provides marginal (fixed effects only) and conditional (fixed + random effects) $R^2$:

```{r}
r2(m1c)
```

The marginal $R^2$ represents the variance explained by the fixed effects (the average trajectory), and the conditional $R^2$ includes the random effects (individual variability around that trajectory). The difference between them is the variance attributable to between-person differences in intercepts and slopes.

### Expanded interpretation: `r2(m1c)`


- **Marginal $R^2 = 0.046$.** The *fixed* effects — here, only the average time slope — explain about 4.6% of the total variance in closeness with dad. This is small, and it is exactly what we should expect in a growth model whose only fixed predictor is time: the average downward trend accounts for very little of the total variation, because most of the variation is between children (different intercepts) and around each child's own line.
- **Conditional $R^2 = 0.645$.** Once we also include the *random* effects (random intercepts and slopes), the model explains 64.5% of the total variance. The difference between the two, $0.645 - 0.046 = 0.599$, is the share attributable to between-person differences in level and change.
- **How to report it.** Because the two measures answer different questions, report both: the marginal $R^2$ describes how much the *average trajectory* explains, and the conditional $R^2$ describes how much the *full multilevel model* explains. A common mistake is to report only the conditional $R^2$, which overstates the role of the predictors; here the honest reading is that the substantive action is in the random effects, not the fixed time effect.
- **A low marginal $R^2$ is not a problem here.** In cross-sectional MLM, a low marginal $R^2$ often signals weak predictors. In growth models it frequently just reflects that time is a weak *fixed* predictor of a highly variable outcome, while the individual trajectories are the real story. Adding predictors of the intercept and slope (e.g., `SEX_cat`, `CLOSEMOM_C`) would raise the marginal $R^2$.


### Level-Specific Pseudo-R²

We can also compute the level-specific pseudo-$R^2$ measures (Raudenbush & Bryk, 2002) by comparing the null model to the fitted model:

```{r}
#| eval: true
# Level-1 (within-person) R²: reduction in residual variance
sigma2_null <- as.data.frame(VarCorr(m0a)) |>
  subset(grp == "Residual") |> (`[[`)("vcov")
sigma2_m1c <- as.data.frame(VarCorr(m1c)) |>
  subset(grp == "Residual") |> (`[[`)("vcov")

(sigma2_null - sigma2_m1c) / sigma2_null
```

### Rights & Sterba (2019) Framework

The `r2mlm()` function provides a comprehensive decomposition of variance into components attributable to level-1 predictors via fixed slopes ($f_1$), level-2 predictors via fixed slopes ($f_2$), random slope variation ($v$), random intercept variation ($m$), and residual (Rights & Sterba, 2019):

```{r}
#| eval: true
r2mlm(m1c, bargraph = FALSE)
```

In a growth model, the $v$ component captures the variance due to **random slopes** (individual differences in change), and the $m$ component captures the variance due to **random intercepts** (individual differences in starting level). These are the two most substantively interesting components in longitudinal models.

### Expanded interpretation: `r2mlm(m1c)`


The `r2mlm()` decomposition (Rights & Sterba, 2019) splits the total variance into four non-overlapping sources. The four `$Decompositions` values sum to 1.00:

| Source | Proportion | Meaning |
|---|---|---|
| `fixed` | 0.046 | Variance explained by the fixed time trend |
| `slope variation` | 0.025 | Variance due to individual differences in *rate of change* (random slopes) |
| `mean variation` | 0.574 | Variance due to individual differences in *starting level* (random intercepts) |
| `sigma2` | 0.355 | Unexplained within-person residual variance |

Check: $0.046 + 0.025 + 0.574 + 0.355 = 1.00$.

**Reading each component.**
- **`fixed` (0.046).** Same as the marginal $R^2$: the average trajectory explains 4.6% of total variance.
- **`slope variation` (0.025).** Only about 2.5% of the total variance is attributable to children changing at different rates. This is the component unique to the random-slope part of the model, and it is small — the average rate of change plus individual intercepts do most of the work. Nonetheless it is non-zero and, as the likelihood-ratio test showed, statistically necessary.
- **`mean variation` (0.574).** By far the largest component. Nearly 58% of the total variance is explained simply by the fact that children differ in their *starting level*. This is the signature of a growth model: between-person differences in level dominate.
- **`sigma2` (0.355).** About 35.5% of the total variance is unexplained within-person variation — the occasion-to-occasion fluctuation around each child's own trajectory. This is the part of the variance that additional time-varying covariates (e.g., `CLOSEMOM_C`) would try to explain.

**Reading the `$R2s` table.**
- **`f` (0.046)** = variance explained by fixed effects = marginal $R^2$.
- **`v` (0.025)** = variance explained by random slope variation.
- **`m` (0.574)** = variance explained by random intercept variation.
- **`fv` (0.071)** = fixed + random slopes. Adding the random slopes to the fixed effects raises the explained variance from 4.6% to 7.1%.
- **`fvm` (0.645)** = fixed + random slopes + random intercepts = the **conditional $R^2$**, and it matches `r2()` exactly (0.645). This is a useful consistency check: the two packages agree.

**Substantive summary.** The decomposition makes the structure of the growth model explicit. The overwhelming majority of the explainable variance is between-person variation in *level* (`m` = 0.574); a small amount comes from the fixed time trend (`f` = 0.046) and from individual differences in *change* (`v` = 0.025); and the rest is within-person residual noise (`sigma2` = 0.355). When reporting a growth model, this decomposition is more informative than a single $R^2$ because it shows *where* the explained variance comes from — and it highlights that, in these data, level differences between children matter far more than differences in their rates of change.

## Visualizing Trajectories

The `marginaleffects` package is particularly useful for growth models because it can produce both **unit-level** (person-specific) and **population-level** (average) predicted trajectories.

### Unit-Level Predictions

Unit-level predictions use the random effects, so each person gets their own fitted line:

```{r}
#| fig-width: 8
#| fig-height: 5
# Build a prediction grid for 20 sampled persons
nd <- datagrid(model = m1c,
               id = sample(m1c@frame$id, 20),
               GRADE_C = 0:5)

pred1 <- predictions(m1c, newdata = nd)

ggplot(pred1, aes(GRADE_C, estimate, group = id)) +
  geom_line() +
  labs(y = "Predicted Closeness with Dad",
       x = "Grade",
       title = "Linear growth model (unit-level predictions)")
```

### Unit-Level Predictions with a Predictor

We can also facet the person-level trajectories by a predictor such as sex:

```{r}
#| fig-width: 9
#| fig-height: 5
nd <- datagrid(model = m1f,
               id = sample(m1f@frame$id, 100),
               GRADE_C = 0:5,
               SEX_cat = levels(m1f@frame$SEX_cat))

pred2 <- predictions(m1f, newdata = nd)

ggplot(pred2, aes(GRADE_C, estimate, group = id)) +
  geom_line() +
  ylab("Predicted Closeness with Dad") +
  facet_wrap(~ SEX_cat, labeller = label_both)
```

### Population-Level Predictions

For population-level inference we average over the random effects by setting `re.form = NA` (or `id = NA` in the grid). This gives the average trajectory and its confidence band:

```{r}
#| fig-width: 8
#| fig-height: 5
# Population-level predictions (no predictors)
nd <- datagrid(model = m1c, id = NA, GRADE_C = 0:5)

pred3 <- predictions(m1c, newdata = nd, re.form = NA)

ggplot(pred3, aes(x = GRADE_C, y = estimate,
                  ymin = conf.low, ymax = conf.high)) +
  geom_ribbon(alpha = .1) +
  geom_line() +
  labs(title = "Population-level trajectories")
```

And with a predictor, we can compare population-level trajectories across groups:

```{r}
#| fig-width: 9
#| fig-height: 5
# Population-level predictions by sex
nd <- datagrid(model = m1f, id = NA, GRADE_C = 0:5,
               SEX_cat = levels(m1f@frame$SEX_cat))

pred4 <- predictions(m1f, newdata = nd, re.form = NA)

ggplot(pred4, aes(x = GRADE_C, colour = SEX_cat, y = estimate,
                  ymin = conf.low, ymax = conf.high)) +
  geom_ribbon(alpha = .1) +
  geom_line() +
  labs(title = "Population-level trajectories by sex",
       colour = "Sex")
```

The unit-level plots show the heterogeneity in individual trajectories, while the population-level plots show the average pattern (and its uncertainty). Reporting both is often the clearest way to communicate a growth model.

## Summary and Recommendations

Growth models in MLM are just another MLM regression. We need to adjust the interpretation as the structure and predictors change in function of time; all the other MLM characteristics apply the same way (Singer & Willett, 2003).

Key points to remember:

1. **Longitudinal data are nested data.** Repeated measures (level 1) are nested within persons (level 2). Everything we learned about random intercepts and slopes in 16_MLM1 applies directly (Hox et al., 2018).

2. **The time origin matters.** Recentering time changes the interpretation of the intercept, the intercept variance, and the intercept–slope covariance — but not the model fit. Center time at a meaningful value (usually the first wave). Rescaling time changes the slope metric (Singer & Willett, 2003).

3. **The aperture** is the time point with minimum intercept variance and zero intercept–slope covariance. It can be useful for study planning but may fall outside the observed range (Singer & Willett, 2003).

4. **Choose the time metric deliberately.** Wave, grade, and age can give different substantive answers. Let theory guide the choice (Biesanz et al., 2004).

5. **Build models sequentially.** Start with a null (random intercept) model, then add time as a fixed slope, then test whether the slope should be random, then add predictors. Use likelihood ratio tests (`anova()`) and information criteria (AIC, BIC) for model comparison.

6. **Use profile likelihood confidence intervals** for random effects — they are more accurate than Wald intervals (Hox et al., 2018). In this tutorial we obtain them with `parameters(model, ci_method = "profile")`.

7. **Interpret fixed and random effects together.** The fixed effects describe the average trajectory; the random effects describe how much people differ from it. The intercept–slope covariance describes whether starting level relates to rate of change (Singer & Willett, 2003).

8. **Center time-varying covariates** so the intercept is interpretable, and remember that level-2 predictors of the slope create time × predictor interactions (cross-level interactions) (Singer & Willett, 2003).

9. **Probe and plot interactions** with the `marginaleffects` package (`avg_slopes()` and `plot_predictions()`), exactly as in 16_2_MLM_interactions, but interpret them in terms of growth (differences in starting point and/or rate of change) (Aguinis et al., 2013).

10. **Report effect sizes** using marginal and conditional $R^2$ (`performance::r2()`) and the full decomposition (`r2mlm::r2mlm()`), and visualize both unit-level and population-level trajectories (`marginaleffects`) (Rights & Sterba, 2019).

## References

- Aguinis, H., Gottfredson, R. K., & Culpepper, S. A. (2013). Best-practice recommendations for estimating cross-level interaction effects using multilevel modeling. *Journal of Management, 39*(6), 1490–1528.
- Baltes, P. B., & Nesselroade, J. R. (1979). History and rationale of longitudinal research. In J. R. Nesselroade & P. B. Baltes (Eds.), *Longitudinal research in the study of behavior and development* (pp. 1–39). Academic Press.
- Biesanz, J. C., Deeb-Sossa, N., Papadakis, A. A., Bollen, K. A., & Curran, P. J. (2004). The role of coding time in estimating and interpreting growth curve models. *Psychological Methods, 9*(1), 30–52.
- Duncan, T. E., & Duncan, S. C. (2004). An introduction to latent growth curve modeling. *Behavior Therapy, 35*(2), 333–363.
- Enders, C. K., & Tofighi, D. (2007). Centering predictor variables in cross-sectional multilevel models: A new look at an old issue. *Psychological Methods, 12*(2), 121–138.
- Hox, J. J., Moerbeek, M., & van de Schoot, R. (2018). *Multilevel analysis: Techniques and applications* (3rd ed.). Routledge.
- Little, T. D. (2024). *Longitudinal structural equation modeling* (2nd ed.). Guilford Press.
- Peugh, J. L. (2010). A practical guide to multilevel modeling. *Journal of School Psychology, 48*(1), 85–112.
- Raudenbush, S. W., & Bryk, A. S. (2002). *Hierarchical linear models: Applications and data analysis methods* (2nd ed.). Sage.
- Rights, J. D., & Sterba, S. K. (2019). Quantifying explained variance in multilevel models: An integrative framework for defining R-squared measures. *Psychological Methods, 24*(3), 309–338.
- Singer, J. D., & Willett, J. B. (2003). *Applied longitudinal data analysis: Modeling change and event occurrence*. Oxford University Press.
- Snijders, T. A. B., & Bosker, R. J. (2012). *Multilevel analysis: An introduction to basic and advanced multilevel modeling* (2nd ed.). Sage.