---
title: "Importing data sets -- short version"
author: "Mauricio Garnier-Villarreal, Joris M. Schröder & Joseph Charles Van Matre"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 5
editor_options:
  chunk_output_type: inline
---

# Introduction

In this short version of the *importing data sets tutorial*, we focus on the the `import()` function from the `rio` package. A major advantage of the `import()` function is that it works with many file types. Because of this simplicity, we will use `import()` throughout the remaining tutorials. 

# Intoduction to data formats

A common characteristic across the different methods is that all the data sets are located in the folder that is **set as working directory**. But if you have them in another folder we will also show how to set the file path. 

Not every data set is the same. Depending on the format, a data set can be a *simpler* or more *complex* format to import and work with. Data can be shared in many format. The simpler ones are text types (like `.csv`, `.txt`), and more complex formats are the ones that are software specific (like `.sav` from `SPSS` and `.dta` from `Stata`). The text types only have the data set in a simple structure, like values separated by commas (`.csv` = comma-separated values) or values separated by blank spaces or tabs (`.txt`). The software data types, in contrast, include the data, and a series of code that defines the specific software structure behind it. 

In general terms, we suggest to export and share data in a text format, as these ones are easier to open and share across platforms and software programs. 

Different data formats require specific methods to import them into **R**. Therefore, it is important to be aware of the format that your data is saved in. The easiest way to figure that out is to look at the file extension of your data file. For example, The `.csv` extension of the `data_set.csv` file shows that this file is saved in the comma-separated values (`csv`) format. A problem is that most computers will hide the file extensions by default, but you easily look up it up online how to show the file extensions in your operating system. 

Below, we show you how to import data sets in different formats into **R**. We will use the different data formats in which the [World Value Survey (WVS)](https://www.worldvaluessurvey.org/) can be downloaded. These include the most common formats you will encounter, namely `text (.csv)`, `SPSS (.sav)`, `SAS (.sas7bdat)`, and `Stata (.dta)`. The data sharing rules of the WVS forbid us to upload the data, but you can simply download it from the [WVS website](https://www.worldvaluessurvey.org/). Tutorial 3.1 shows you how to do this. Here we will work with the *Wave 7 cross sectional* data set in text, SPSS, SAS, and Stata data formats. 


# The `rio` package
The `rio` package is designed to make data importing, exporting as user friendly as possible. It will automatically decide which functions is best to use for importing data, depending on the file type. The user only needs to know one function that works with many file types. As always, we need to start by loading the `rio` package (assuming that it is already installed).

```{r, eval=T}
library(rio)
```

As stated on its [support page](https://cran.r-project.org/web/packages/rio/vignettes/rio.html) the `rio` package supports a variety of file formats for import and export. To keep the package slim, non-essential formats are not installed (or loaded) by default. To ensure `rio` is fully functional, install these packages the first time you use `rio` via:

```{r, eval=F}
install_formats()
```

The main functionality of the `rio` package is provided by the `import()` function. This function does not actually import the file, but it identifies the file format and chooses the function from another package to import it. You can see the packages used to import each file format with  

```{r, eval=F}
?import
```


## Importing text files

To  import text files, we simply need to give the file name in quotes to the `import()` function, and specify possible missing value codes. In the case of the WVS, some missing values are coded as -999 and -9999 (because these are unlikely to occur in real data). To make sure that these are read as missing values, we provide `na.strings = c("-999", "-9999")` as the second argument to the `import()` function, as you can see below

```{r, eval=T}
dat <- import("WVS_Cross-National_Wave_7_csv_v2_0.csv", na.strings = c("-999", "-9999"))
dim(dat)
is(dat)
```

In the next line we are using the function `dim()` to see the size of each dimension of the data set. As this is a rectangular data set it has 2 dimensions, and the `dim()` function tells us that the data set has 76897 rows (subjects) and 548 columns (variables). Notice that the order in R will always be `rows/columns`. Lastly, we are using the function `is()` to ask R what type of object the imported data set now is in R. In this case, it is imported as a `data.frame`. 

The `attributes()` function shows potential additional information about the data you just imported. As the text file has no additional information, it keeps no information in the variable attributes. You can check this using the `attributes()` function

```{r, eval=T}
attributes(dat$Q60)
```

## Importing SPSS files
As the `import()` function does the decisions how to import data for us, we simply need to provide it with the new file name to import data in the `.sav` format from SPSS. By default, it is using the function from the `haven` package.

```{r, eval=T}
dat <- import("WVS_Cross-National_Wave_7_sav_v2_0.sav")
dim(dat)
is(dat)
```

Again, we use `dim()` to check the dimensions of the data set, and `is()` to check the type of object that the data is stored as. An advantage of using a data format other than text is that the file can include additional information. For example we see the details about each variable with the strcuture fuction `str()`

```{r, eval=F}
str(dat)
```

The output of this function is too long to show here but you can try for yourself. `str()` shows the data type of each each variable (e.g., integer or character, see tutorial 2), and the values it takes. Again, we can check for additional information using the `attributes()` function, in this case additional information read from the variable information in SPSS

```{r, eval=T}
attributes(dat$Q60)
```

We ask for the attributes of the variable `Q60`, which shows for example the variable label *Trust: People you know personally*, the values the variable takes, and corresponding value labels that are specified in the `.sav` file created in SPSS. 


## Importing SAS files
For SAS files in the `.sas7bdat` format, the `import()` function will also pick the right package for the import of files. 

```{r, eval=T}
dat <- import("WVS_CrossNat_W7_v2_0.sas7bdat")
dim(dat)
is(dat)
```

When looking at the saved attributes, we see that from the SAS file, we are only saving the item label, but not the value labels

```{r, eval=T}
attributes(dat$Q60)
```


## Importing Stata files
Finally, the same goes for the Stata files in the `.dta` format. Simply use the `import()` function which will choose the right package and function to import the data in the background.

```{r, eval=T}
dat <- import("WVS_Cross-National_Wave_7_stata_v2_0.dta")
dim(dat)
is(dat)
```

As it is using the `haven` functions, we keep the same information in the variable attributes

```{r, eval=T}
attributes(dat$Q60)
```


# Additional information
In the previous examples, we have use defaults that work well for general use cases. Here we present a few uses and cases you might ran into

## Need more detail about the functions
So far in our examples we used the defaults or minimal number of arguments for each function. You might run into a case that requires to manipulate the functions further. When this comes to be, go to the help page of a specific function to identify the detail on how it works, and other arguments on how to modify how it works

```{r,eval=F}
## for example
?import
```


## File paths
In the previous examples we assumed that the data file you want to import was in the working directory. If that is not the case, you can still import the file, but just requires to specify the full path for the file location, like this

```{r eval=FALSE}
dat <- import("C:/Users/Admin/Dropbox/Magic Briefcase/Vrije/R_tutorials/git/WVS_Cross-National_Wave_7_sav_v2_0.sav")
dim(dat)
is(dat)
```

Here is important to make sure the slash is this direction \/, some operating system tend to change the direction on the file manager. 

## `choose.file()`
So far we have used the import functions with the full file name (or path), but some people don't like this or are too use to browsing for a file. We can do this if we change the file name and ask R to search for a file in the browser

```{r eval=FALSE}
### WVS_Cross-National_Wave_7_sav_v2_0.sav
dat <- import(file.choose())
dim(dat)
is(dat)
```

When using the `file.choose()` function instead of the file name, R will open a browse window where you can look for the respective file in your computer. If you choose to use this approach, we strongly recommend to add the file name, or some identifier for the imported file. This is so we can keep track of which data set was imported for the future and reproducibility. 

## General recommendations
Some recommendations for general practice are:

* Use the `import()` function from the `rio` package, this works well for most common data types, and it is user friendly. 
* Save the data sets in the working directory folder, for facility to have all files for a respective analysis in a single place.
* Specify the full file name in the functions reading in the data. This would give us more tractability of the data analysis process and ease to reproduce it. 

