---
title: "Data management 1"
author: "Mauricio Garnier-Villarreal, Joris M. Schröder & Joseph Charles Van Matre"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 5
editor_options:
  chunk_output_type: inline
---

# Setup the R session

When we start up we always need to setup our session. For this wee need to set our working directory, in this case I am doing that for the folder that holds the downloaded [World Values Survey (WVS)](https://www.worldvaluessurvey.org/) ```SPSS``` data set

```{r, eval=F, error=F}
setwd("~path_to_your_file")
```


The next step for our session will be to load the packages that we will be using

```{r, eval=T}
library(rio)
library(car)
library(psych)
```


# Import the data set

Here we will be importing the ```.sav``` WVS data set

```{r, eval=T}
dat <- import("WVS_Cross-National_Wave_7_sav_v2_0.sav")
dim(dat)
```

Here we are calling our data set **dat** and asking to see the dimension of it. We see that the data set has 76897 subjects, and 548 columns. If you have downaloaded a different version of the WVS these number might differ. 

## Select variables of interest

In cases with large data sets like this we might want to select a subset of variables that we want to work with. Since it is not easy to see 548 variables. 

```{r, eval=T}
vars <- c("Q65", "Q69", "Q71", "Q72", "Q73", "Q275", "Q260", "Q262")
dat2 <- dat[,vars]
dim(dat2)
head(dat2)
```

Here we are first creating a vector with the variable names for the ones I want to keep. You can see all variable names for the full data set as well

```{r, eval=F}
colnames(dat)
```

After writing which variables we want to work with, we create a new data set **dat2** with only these 7 variables, and make sure we did it correctly by looking at the the dimension of the data **dim(dat2)** and looking at the first 6 rows of it **head(dat2)**. These are quick checks that we have done this correctly. 


# Recode items

## Reverse code

There are cases where you need to switch the direction of some items. For example, we want to create a scale score for **Confidence in the government**, which is evaluated in WVS with the items Q65, Q69, Q71, Q72, and Q73. 

Since we imported the ```SPSS``` file, which contains some attribute information for the items. We can see what information the have

```{r, eval=T}
attributes(dat2$Q65)
```

Here we can see that item **Q65** is about **Confidence in the Armed Forces**. We can also see that 1 means that they have **A great deal of confidence** and that 4 means that they have **None at all confidence**. So, we see that the higher the score the less confidence they have in the government. You can see the other items attributes

```{r,eval=F}
attributes(dat2$Q69) ## Confidence: The Police
attributes(dat2$Q71) ## Confidence: The Government
attributes(dat2$Q72) ## Confidence: The Political Parties
attributes(dat2$Q73) ## Confidence: Parliament
```

So, if we were to create a total score with the items like this, we would be create the scale score for **Lack of confidence in the government**. It just seems too many words, and I dont want to be interpreting negative statements. 

So, lets reverse code the items, so we can create the scale **Confidence in the Government**. First we want to see what are the values that we have in the data set. We can do this with the frequency tables.

```{r, eval=T}
table(dat2$Q65, useNA="always")
table(dat2$Q69, useNA="always")
table(dat2$Q71, useNA="always")
table(dat2$Q72, useNA="always")
table(dat2$Q73, useNA="always")
```

Here we see that for all the items, we have values from 1 to 4, and all other responses have been properly set as missing ```NA```. We are using 2 arguments from the ```table()``` function, first the variable, and second asking what to do with the missing values, so we can see them included in the frequencies. 

We can shorten this code, instead of doing it for one variable at the time. We can use the ```apply()``` function. First, add the data set with only the scale variables to which do want to look at the frequency tables, then the *MARGIN = 2* indicates that you want to execute the function *FUN* to each column, as the 2 inciates the second element of the dimensions. And finally, the function to apply *table*, and additional arguments for the function

```{r, eval=T}
apply(dat2[,c("Q65","Q69","Q71","Q72","Q73")], MARGIN=2, FUN=table, useNA="always")
```

At the end we get the same information, but it is good to see a couple of ways, so you can figure out which one you are more comfortable with. 

Now we know that we need to reverse code values from 1 to 4. We will use the ```recode()``` function from the ```car``` package. 

```{r, eval=T}
dat2$Q65_r <- recode(dat2$Q65, "1=4; 2=3; 3=2; 4=1")
dat2$Q69_r <- recode(dat2$Q69, "1=4; 2=3; 3=2; 4=1")
dat2$Q71_r <- recode(dat2$Q71, "1=4; 2=3; 3=2; 4=1")
dat2$Q72_r <- recode(dat2$Q72, "1=4; 2=3; 3=2; 4=1")
dat2$Q73_r <- recode(dat2$Q73, "1=4; 2=3; 3=2; 4=1")
```

First, we need to create a new variable in the data set, we can use t ```$``` to create a new item name. Here I am using the same name, ending with ```_r```, indicating that is the same item reversed. Then, we the ```recode()``` function takes 2 arguments, first the variable that you want to recode, second the way the values should be changed. Note that the recode values need to be between quotes ```""```, and that the values have to be separated by ```;```. 

Now we have reverse coded the 5 items, but we are not done yet. We need to make sure we did it correctly. We can do this by doing a cross-table between the original variable, and the new one. 

```{r, eval=T}
table(dat2$Q65, dat2$Q65_r, useNA="always")
```

Here we can see that the variable values follow the diagonal, indicating that values were recoded correctly. Only the ```NA``` is the same, because we want to keep the same missing values. You can see the cross-tables for the other variables with

```{r, eval=F}
table(dat2$Q69, dat2$Q69_r, useNA="always")
table(dat2$Q71, dat2$Q71_r, useNA="always")
table(dat2$Q72, dat2$Q72_r, useNA="always")
table(dat2$Q73, dat2$Q73_r, useNA="always")
```

Now we can see all the new variables in the data set

```{r, eval=T}
dim(dat2)
head(dat2)
colnames(dat2)
```

## Reverse code missing value codes

From the attributes we have seen that there were missing value codes, even when those had been correctly set as ```NA``` when importing the data set. But this might not be the case in all your data sets

```{r, eval=T}
attributes(dat2$Q65)
```

Here we see that values -5, -4, -2, and -1 should be considered missing. We can include these into the recode function, by adding these negative values and making them equal to ```=NA```. 

```{r, eval=F}
dat2$Q65_r <- recode(dat2$Q65, "1=4; 2=3; 3=2; 4=1; -5=NA; -4=NA; -2=NA; -1=NA")
```

## Dummy code

Another issue could be to have a categorical variable, and you need to create multiple dummy code variables. Changing a single categorical variable into multiple binary variables. 

For this example we will use the **Highest education level (Q275)** variable. In the attributes, we can see that the valid answers go from 0 to 9. But when looking at the frequency table, we see that we have only values from 0 (*Early childhood education*) to 8 (*Doctoral or equivalent*) 

```{r, eval=T}
attributes(dat2$Q275)
table(dat2$Q275, useNA = "always")
```

This would mean that we need to create 9 dummy code variables. In each dummy code variable having 1 indicating for each education level and 0 for all others. You could do this with a bunch of ```recode()``` functions. But we also have a nice function ```dummy.code()```  from the ```psych``` package. 

```{r}
dat2 <- data.frame(dat2, dummy.code(dat2$Q275))
head(dat2)
colnames(dat2)
```

This function give us a matrix with all the dummy code variables. We can combine it with the full data set with the ```data.frame()``` function. And we can see that the dummy code variables are manes ```X1```for Dummy code for value 1, we can see this with a cross table

```{r}
table(dat2$Q275, dat2$X1)
```


You could check that for all other values

```{r, eval=F}
table(dat2$Q275, dat2$X0)
table(dat2$Q275, dat2$X2)
table(dat2$Q275, dat2$X3)
table(dat2$Q275, dat2$X4)
table(dat2$Q275, dat2$X5)
table(dat2$Q275, dat2$X6)
table(dat2$Q275, dat2$X7)
table(dat2$Q275, dat2$X8)
```




# Create composite scores

We create composite scores when we need to combine multiple items into 1 *total* score. The most common case is when we need to create a scale score, like the **Confidence in the government** one. 

## Sum score 

We will see here to ways to create the composite scores. The first, and I think the most common one is to sum up all the items. We can do this creating a new variable with ```$```, and summing values for each row (subject) across a set of items

```{r, eval=T}
dat2$CG_sum <- rowSums(dat2[,c("Q65_r","Q69_r","Q71_r","Q72_r","Q73_r")], na.rm=T)
colnames(dat2)
```

For this we use the ```rowSums()``` function, which sums up values across rows. The first argument is the data set for which we want to sum up values, note this has to include only the items you want to sum. And the other argument we are using the ```na.rm=T```, this specifies that when missing values are present to ignore them and sum as many values as each subject has. Functionally treating the missing as a 0. 

## Mean score

The another method we will see is to get the average across items instead of the sum. This is very similar as before, we just switch to use the ```rowMeans()``` function.

```{r}
dat2$CG_mean <- rowMeans(dat2[,c("Q65_r","Q69_r","Q71_r","Q72_r","Q73_r")], na.rm=T)
colnames(dat2)
```

Now we see we have created 2 composite scores. With the sum and the mean across items. We recommend to use the mean instead of the sum. This for several reasons, but the main ones being that the mean keeps the scale of the original items (between 1 and 4 in this case) while the sum is in the scale between the minimum and maximum of the sum. And the second because with the sum the missing values either create a missing sum or counts the missing value as 0, while when the mean ignores it it will estimate the mean across all the answered items, not affecting the mean. 


# Variable calculations

Creating composite variables are a way of doing calculations with variables. But we can do any calculation that we consider necessary. 

## Equation calculations

For example, we can table the mean composite score and subtract any value of interest we want, could be 2 as it is the median range of possible scores

```{r, eval=T}
dat2$CG_mean_minus2 <- dat2$CG_mean - 2
head(dat2)
```
Now this variable would show negative values for subjects with a mean scors lower than 2. 

Another calculation we could do is to divide the mean composite score by the subjects age, making it relative to it. Not sure of this makes a lot of sense, but its something curious to show

```{r, eval=T}
dat2$CG_mean_age <- dat2$CG_mean/dat2$Q262
head(dat2)
```

## Scale

Another common variable calculation is to scale a variable. This means to change mean and/or standard deviation of the variable. You might up needing to mean center a variable, we can do this with the ```scale()```function

```{r}
dat2$CG_mean_center <- scale(dat2$CG_mean, center=T, scale = F)[,1]
head(dat2)
```

With this function, we can specify to mean center the variable ```center=T```, but not to modify the standard deviation ```scale=F```. If we want to create a standardized variable out of it we can include both arguments, lets remeber that a standardized variable is when the mean = 0, and sd = 1

```{r}
dat2$CG_std <- scale(dat2$CG_mean, center=T, scale = T)[,1]
head(dat2)
```

By setting both arguments ```T```, we have standardized the mean composite score of **Confidence in the government**. 


# Selecting subsets 

In many cases you would want to work a subset of subjects in the data set. So lets see a few ways to do these selections. We will use the ```subset()``` function. First, lets see if we wanted to select only subjects that have **Bachelor or equivalent** as their highest education level.

```{r, eval=F}
attributes(dat2$Q275)
table(dat2$Q275, useNA = "always")
```

We can see that this has the value **6** in this data set, so we can select these cases like this

```{r}
dat2_ed6 <- subset(dat2, Q275 == 6)
dim(dat2_ed6)
```

We first named our new data set, ```dat2_ed6``` in this case. Then use use the subset function, the first argument is the full data set. And the second argument is a (series) logical argument(s) to do the selection. In this case that **Q275 == 6** this indicating that only values equal to 6 will be selected. Note that these are 2 equal signs, not one. 

Here we present to logical operators you can use in R. 

Operator     | Description
----------   | --------------------
==           | exactly equal
\<           | lower than
\<=          | lower than or equal  
\>           | higher than
\>=          | higher than or equal
x & y        | x AND y
x \| y       | x OR y
!=           | not equal to
!x           | not x


Now that we have seen some operators, lets make a couple of other selection methods. Lets say that you want to select The subjects that have Bachelor or higher, this means values of 6 or higher

```{r}
dat2_ed678 <- subset(dat2, Q275 == 6 | Q275 == 7 | Q275 == 8)
dim(dat2_ed678)
table(dat2_ed678$Q275)
```


We can also do this with continuous variables, we have the age of the subjects (Q262), so lets say that we want to select only subjects that are older than 18 years old, meaning 19 or higher

```{r}
dat2_age19 <- subset(dat2, Q262 > 18)
dim(dat2_age19)
```

Or we could select 18 or higher, meaning to not exclude 18 years old

```{r}
dat2_age18 <- subset(dat2, Q262 >= 18)
dim(dat2_age18)
```

For a last one, lets combine this. Lets select subjects that have a bachelor degree as their highest education level, and they are 25 years old or older

```{r}
dat2_bach25 <- subset(dat2, Q262 >= 15 & Q275 == 6)
dim(dat2_bach25)
```

