---
title: "Basic plots"
author: "Mauricio Garnier-Villarreal, Joris M. Schröder & Joseph Charles Van Matre"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 5
editor_options:
  chunk_output_type: inline
---


# Setup the R session

When we start an R session we always need to set our working directory. In this case I am doing that for the folder that holds the downloaded [World Values Survey (WVS)](https://www.worldvaluessurvey.org/) ```SPSS``` data set.

```{r, eval=F, error=F}
setwd("~path_to_your_file")
```


The next step for our session will be to load the packages that we will be using.

```{r, eval=T}
library(rio)
library(ggplot2)
```


# Import the data set

Here we will be importing the ```.sav``` WVS data set

```{r, eval=T}
dat <- import("WVS_Cross-National_Wave_7_sav_v2_0.sav")
dim(dat)
```

Here we are calling our data set **dat** and asking to see the dimension of it. We see that the data set has 76897 subjects, and 548 columns. 

## Select variables of interest

In cases with large data sets like this we might want to select a subset of variables that we want to work with. Since it is not easy to see 548 variables. 

```{r, eval=T}
vars <- c("Q275", "Q260", "Q262", "Y001", "B_COUNTRY_ALPHA", "SACSECVAL",
          "Q112","Q113","Q114","Q115","Q116","Q117","Q118","Q119","Q120",
          "Q65","Q69","Q71","Q72","Q73")
dat2 <- dat[,vars]
dim(dat2)
head(dat2)
```

Here we are first creating a vector with the variable names for the ones we want to keep. You can see all variable names for the full data set as well:

```{r, eval=F}
colnames(dat)
```

After writing which variables we want to work with, we create a new data set **dat2** with only these 7 variables, and make sure we did it correctly by looking at the the dimension of the data **dim(dat2)** and looking at the first 6 rows of it **head(dat2)**. These are quick checks that we have done this correctly. 

# ```ggplot2``` basics

We will use the package ```ggplot2``` for most plots. Lets first go over some of the basics. 

```ggplot2``` creates packages in layers. Think about starting with a simple plot. Once we have a base, we can add layers on top. We will see in our examples how to start from simple plots and then add complexity. 

This makes ```ggplot2``` able to to manipulate the plot's features with great flexibility. We will start simple, but ```ggplot2``` can do much more than what we will see here. 

# Histograms

Lets start with a basic histogram for the *Secular values* variable in the data
```{r}
attributes(dat2$SACSECVAL)
```

The basic histogram starts with ```ggplot()``` function, where the first part ```ggplot(dat2, aes(x=SACSECVAL))``` creates an empty plot, which will not print anything until you add some *geom* to it. In this case we are adding ```geom_histogram()```. Think of the first part as telling with which variables we will work, and second part what to do with those variables

```{r}
ggplot(dat2, aes(x=SACSECVAL))+
  geom_histogram()
```

The function finds a number of ```bins``` for the plot, but you can modify it as you wish, to what you consider looks better in your plot

```{r}
ggplot(dat2, aes(x=SACSECVAL))+
  geom_histogram(bins=50)
```


Similar to a histogram, we can plot the density

```{r}
ggplot(dat2, aes(x=SACSECVAL))+
  geom_density()
```

Notice that the histogram shows the count on the *y* axis, while the density plot shows the *density*.

You can change the color of the columns in the histogram with the ```fill``` argument
```{r}
ggplot(dat2, aes(x=SACSECVAL))+
  geom_histogram(fill = "steelblue")
```

Maybe you want the histogram plot with the density metric. We can do this by specifying the density as the metric of the *y* axis this way
```{r}
ggplot(dat2, aes(x=SACSECVAL))+
  geom_histogram(aes(y = ..density..))
```

## Divide by color

We could be interested in seeing the histogram for different groups; this can be be done by specifying color differences according a one grouping variable like this:
```{r}
ggplot(dat2, aes(x = SACSECVAL, fill = as.factor(Q260) ))+
  geom_histogram()
```

Notice that for this to work, R has to know that it is a categorical variable that is why we are using the ```as.factor()``` function. Lets do this after we use the ```factor()``` function to modify the *sex* variable

```{r}
dat2$Sex <- factor(dat2$Q260, 
                       levels = 1:2,
                       labels = c("Male","Female") )
attributes(dat2$Sex)
```


```{r}
ggplot(dat2, aes(x = SACSECVAL, fill = Sex ))+
  geom_histogram()
```

Now that we have meaningful labels, we can see the histogram for both male and female subjects. Let's now look at the plots by the *edcuation* variable. We modify it with the ```factor()``` function the same way.

```{r}
dat2$Education <- factor(dat2$Q275, 
                       levels = 0:8,
                       labels = c("Early childhood",
                                  "Primary",
                                  "Lower secondary",
                                  "Upper secondary",
                                  "Post-secondary non-tertiary",
                                  "Short-cycle tertiary",
                                  "Bachelor or equivalent",
                                  "Master or equivalent",
                                  "Doctoral or equivalent") )
attributes(dat2$Education)
```

Let's do the histogram, with the density metric in the *y* axis, with different fill colors according to the education levels
```{r}
ggplot(dat2, aes(x = SACSECVAL, fill = Education ))+
  geom_histogram(aes(y=..density..))
```

## Grid of plots
And now we can separate the plots, creating one for male and one for female subjects, in a grid. The ```facet_grid()``` sets columns and rows of plots, by the specified grouping variables. With the ```Var1~Var2``` format you can specify that Var1 defines the columns, and Var2 defines the rows. If you only define one of this, you will only get rows or columns instead of a full grid. In this case we are looking at 1 row columns following the *Sex* variable
```{r}
ggplot(dat2, aes(x = SACSECVAL, fill = Education ))+
  geom_histogram(aes(y=..density..))+
  facet_grid(~Sex)
```

But here we have a problem, we get a full new plot for the ```NA``` in the *Sex* variable. Let's exclude that. We can do that with a little data manioulation in the data object when we called it into the function. We can use ```subset(dat2, !is.na(Sex))```. We are asking the function to first get a subset of the data where the variable *Sex* is not ```NA```.

```{r}
ggplot(subset(dat2, !is.na(Sex)), aes(x = SACSECVAL, fill = Education ))+
  geom_histogram(aes(y=..density..))+
  facet_grid(~Sex)
```

## Final plot

Let's add one last modification. We want clear labels for the axis and title for the plot. We can do this with  
```{r}
ggplot(subset(dat2, !is.na(Sex) & !is.na(Education)), 
       aes(x = SACSECVAL, fill = Education ))+
  geom_histogram(aes(y=..density..))+
  facet_grid(~Sex)+
  labs(x="Overall Secular Values", y="Density", 
       title = "Histogram across education and sex")
```

That is a histogram ready for a paper!!

# Scatter plot

When we have 2 continuous variables, the most common plot is a scatter plot with one variable on the *x* axis and another variable on the *y* axis. For this we are going to start by creating 2 scale variables *Corruption Perception* and *Lack Confidence in the Government*

```{r}
dat2$Corrup <- rowMeans(dat2[,paste0("Q",112:120)],na.rm = T)
dat2$GovConf <- rowMeans(dat2[,c("Q65","Q69","Q71","Q72","Q73")],na.rm=T)
```

This basic plot can be define like this: we specify which variable we want in the *x* and the *y* axis and then specify the scatter plot with the ```geom_point()``` 
```{r}
ggplot(dat2, aes(x = Corrup, y = GovConf ))+
  geom_point()
```

## Prediction line

A common next step would be to include the *regression* line between these 2 variables, we can do this with ```geom_smooth(method="lm")``` which specifies the linear regression line with the method linear model, or *lm*. 
```{r}
ggplot(dat2, aes(x = Corrup, y = GovConf ))+
  geom_point()+
  geom_smooth(method = "lm", se=T)
```

We can also choose a best fitting line with ```method="loess"``` , which is not stricted to set a straight line, but can present a curve line. In this case we are adding ```se=F```, because of a memory limit when computing, but we recommend to always use ```se=T```, as this will also present the confidence interval bands. 
```{r}
ggplot(dat2, aes(x = Corrup, y = GovConf ))+
  geom_point()+
  geom_smooth(method = "loess", se=F)
```

## Divide by color

We can also add color or grid plots like we did with histograms. For example, let's add the same color differences by *Sex*. Like before, we are excluding cases with ```NA```in the *Sex* variable. Now, by adding the color difference by sex we are also seeing the different *loess* lines. 
```{r}
ggplot(subset(dat2, !is.na(Sex)), 
       aes(x = Corrup, y = GovConf, colour = Sex))+
  geom_point()+
  geom_smooth(method = "loess", se=F)
```

## Grid of plots

Now, lets see how it looks when we add *Education* as a grid of plots. Here we are using the ```facet_wrap()``` which takes all the education groups and create a nice grid based on each level. We can see the bivariate relation at each education level, and the different *loess* line for each sex group
```{r}
ggplot(subset(dat2, !is.na(Sex) & !is.na(Education)), 
       aes(x = Corrup, y = GovConf, colour = Sex))+
  geom_point()+
  geom_smooth(method = "loess", se=F)+
  facet_wrap(vars(Education) )
```

## Final plot

Finaly, lets add some labels and a title:
```{r}
ggplot(subset(dat2, !is.na(Sex) & !is.na(Education)), 
       aes(x = Corrup, y = GovConf, colour = Sex))+
  geom_point()+
  geom_smooth(method = "loess", se=F)+
  facet_wrap(vars(Education) )+
  labs(x="Corruption Perception", 
       y="Lack of Confidence in the Government", 
       title = "This is some cool Plot!")
```




# Bar plot

A basic plot you might want to do for categorical data, the visual equivalent to a frequency table, is a *bar* plot. With this one we seek to represent the number of participants answering each option. We can make this plot with the ```geom_bar()``` addition to ```ggplot```.

```{r}
ggplot(dat2, 
       aes(Education))+
  geom_bar()
```


We see here a problem, the category labels are too long to show properly in the *x-axis*. A possible solution is to modify these labels to 2 or 3 letter codes. We can do this with the ```factor()``` function. 

```{r}
levels(dat2$Education)
dat2$Education <- factor(dat2$Education,
                         levels = levels(dat2$Education),
                         labels = c("EC","Pri","LS","US","PSnT","ScT","Bach","Mas","Doc"))
levels(dat2$Education)
```

With the ```levels()``` function we can see the existing category names, and decide what to change them to. Here we are changing them for some short hand codes. Now that we have shorter labels, we can do the bar plot, and read the *x-axis*

```{r}
ggplot(dat2, 
       aes(Education))+
  geom_bar()
```

Now we can see how to add some more complexity to our bar plots. First, we want to get rid of the missing values in our plots, we can do this with the ```subset()``` function when we call the data set. 

## Adding color

Next, we want to see the bar plot for multiple groups, for example for male and female participants. We can do this be adding ```fill = Sex``` to the ```aes()``` argument, and adding ```position=position_dodge()``` to ```geom_bar()```. The fill argument tells it to have different filled in colors for each group, and the position argument tells it to set each group next to each other.  

```{r}
ggplot(subset(dat2, !is.na(Sex) & !is.na(Education)), 
       aes(Education, fill = Sex))+
  geom_bar(position=position_dodge())
```

## Grid of plots

Another way to see the bar plots for different groups is by setting a grid of plots, with multiple separate bar plots for each group. We can do this by adding ```facet_wrap(vars(Sex) )```. This way ```ggplot2``` will set a grid of plots in function of the categories in the variable Sex. 


```{r}
ggplot(subset(dat2, !is.na(Sex) & !is.na(Education)), 
       aes(Education))+
  geom_bar()+
  facet_wrap(vars(Sex) )
```

# Box plot

Another type of plot you may be interested in is box plots. This one represents the range, median, and outlines for a continuous variable.

We will use the *Corruption Perception* variable for this example. We can do a simple box plot by adding ```geom_boxplot()```. 

```{r}
ggplot(dat2,
       aes(y=Corrup))+
  geom_boxplot()
```
This example, however, is not very informative. 

## By category

Next, we can get box plots across a categorical condition, for example Sex. We can remove the missing values, and set ```x=Sex```, this way we will see the categories of Sex on the *x-axis*

```{r}
ggplot(subset(dat2, !is.na(Sex) ),
       aes(y=Corrup, x=Sex))+
  geom_boxplot()
```

We can do this by Education level as well. We will also have different colors for the respective categories, with ```fill=Education```. 

```{r}
ggplot(subset(dat2, !is.na(Education) ),
       aes(y=Corrup, x=Education, fill=Education))+
  geom_boxplot()
```

## Add statistic

We can add other characteristics, for example add the *mean* in the boxes. We can do this by adding ```stat_summary(fun=mean, geom="point", shape=5, size=2)```, with this we are asking to estimate the mean, use a point geom, with a shape 5 (meaning a diamond), and size of 2 for this geom. You can play with shape and size to see how you want the plot to look. 

```{r}
ggplot(subset(dat2, !is.na(Education) ),
       aes(y=Corrup, x=Education, fill=Education))+
  geom_boxplot()+
  stat_summary(fun=mean, geom="point", shape=5, size=2)
```

## Grid of plots

As we have being doing, let's add a facet grid to this plot, in this case by Sex
```{r}
ggplot(subset(dat2, !is.na(Education) & !is.na(Sex) ),
       aes(y=Corrup, x=Education, fill=Education))+
  geom_boxplot()+
  stat_summary(fun=mean, geom="point", shape=5, size=2)+
  facet_wrap(vars(Sex) )
```

## Final plot

For the final model, we want to change the labels.

```{r}
ggplot(subset(dat2, !is.na(Education) & !is.na(Sex) ),
       aes(y=Corrup, x=Education, fill=Education))+
  geom_boxplot()+
  stat_summary(fun=mean, geom="point", shape=5, size=2)+
  facet_wrap(vars(Sex) )+
    labs(y="Corruption Perception", 
         x="Highest Education level", 
         title = "This is some cool Plot!")
```


# Reference

As we have seen there are a lot of plots you can do with ```ggplot2``` as well as many ways to edit them to look how you want. But this usually needs some extra work and figuring out how the different functionality works. For additional help we suggest the following book

* Wickham H (2016). ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York. ISBN 978-3-319-24277-4, https://ggplot2.tidyverse.org. 

