---
title: "Difference in Means: t-tests"
author: "Mauricio Garnier-Villarreal, Joris M. Schröder & Joseph Charles Van Matre"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 5
editor_options:
  chunk_output_type: inline
---

# Setup the R session

When we start working in R, we always need to setup our session.
For this we need to set our working directory, in this case I am doing that for the folder that holds the downloaded [World Values Survey (WVS)](https://www.worldvaluessurvey.org/) `SPSS` data set

```{r, eval=F, error=F, message=F}
setwd("~path_to_your_file")
```

The next step for setting up our session will be to load the packages that we will be using.
In this case we will need the *effectsize* package.
If you do not already have it installed, make sure to do so.
*effectsize* has quite a few dependencies, so it may take a moment to install.
Other than that we will use the *rio* package to import the data set.

```{r, eval=T, message=F}
#install.packages("effectsize", dependencies = TRUE)
library(effectsize)
library(rio)
```

# Import the data set

We will now import the `.sav` WVS data set 

```{r}
dat <- import("WVS_Cross-National_Wave_7_sav_v2_0.sav")
dim(dat)
```

Here we are calling our data set **dat** and asking to see the dimension of it, we see the full data set has 76897 rows, and 548 columns.

## Currating the data

Since this is a large data set, we will reduce the observations and variables to those we will use in this tutorial.

We will be looking at an index of perceptions of public corruption (Q112 through Q117), sex (Q260), and country (B_COUNTRY_ALPHA).

```{r}
#Keep only German respondents
dat <- dat[dat$B_COUNTRY_ALPHA == c("DEU"),]
#Keep only variables of interest
dat <- subset(dat,select=c(B_COUNTRY_ALPHA,Q260,Q113:Q117))
```

We are using the logical `==` to select only the subjects (rows) from Germany.
And the `subset` function, to select the desired columns.

We also need to create two additional variables to help us with this tutorial.
The first is a "Public Corruption Index".
The tutorial on **Reliability** has more on how to construct and evaluate indexes, but here we are just going to use the mean of 5 survey questions that ask respondents to rate the level of public corruption in different areas of society (e.g. public officials, media, etc.).
We can use the `summary` function to see the distribution of the *corrupt* index.

```{r}
#Create the Public Corruption Index
dat$corrupt <- rowMeans(subset(dat,select=Q113:Q117), na.rm=F)
summary(dat$corrupt)
```

### Transform categorical variables to binary

When we want to use categorical variables as predictors, it is recommended to transformed them as ```factor``` type in ```R```, or as binary. In this case we will use the *Sex* variable ```Q260```, which is coded as 1 and 2 right now (1 = Male, 2 = Female) 

With the ```factor()``` function we an transform it, and change the numbers to meaningful labels. 

```{r}
dat$Sex <- factor(dat$Q260, 
                       levels = 1:2,
                       labels = c("Male","Female") )
head(dat)
```

# Difference in Means

The **mean** is one of the most commonly used measures of central tendency for describing data (together with the median and mode).
When we want to know the "average" score on an exam, the "average" height of a group, or the "average" income in a country, we are interested in the **mean score**, height or income.

Much of quantitative social science comes down to determining if there is a difference in means, and a **t-test** is the most simple form of testing for a difference in means.

There are three ways that we can think about differences in means: 1.
**One sample t-test**: The mean of a group is different from a theoretical mean.
For example: is the mean income in a country equal to the poverty line, say €12,000 a year?
2.
**Two sample t-test**: The mean of one group is different from another group.
For example: is the mean income for men different than for women?
3.
**Paired sample t-test**: The mean of group A is different from group B, but each member of group A has a partner in group B.
This type of test is useful if there is longitudinal data (e.g. when there are scores from an pre- and post-test, an individual's pre-test score is matched with their post-test score) or if there is a clear pairing in the data (e.g. older siblings in group A could be matched with younger sibling in group B.)


# One Sample t-tests

As stated before, a one sample t-test asks if the mean of a group is different from some theoretical mean.
Let's say that we believe that the average score on the corruption index is 2 in these countries.
So our null hypothesis would be that **corrupt = 2**.
We use the `t.test` function to test this hypothesis.
We tell R where to find the that data that has the outcome we are interested `dat$corrupt` and what our hypothesis is `mu=2`.

```{r}
one.sample <- t.test(dat$corrupt, mu=2)
one.sample
```

**R** gives us a lot of information back.
It first give us the t-value, the degrees of freedom, and the p-value.
From this information, we can conclude that we can reject the null hypothesis that the sample mean is equal to 2, which **R** helpfully provides `true mean is not equal to 2` **R** also provides the 95% confidence interval and the mean.
The `t.test` function defaults to a two-sided t-test.
However, you can perform a one-sided t-test with an additional argument in the function

```{r}
#A one sided t-test with a null hy
t.test(dat$corrupt, mu=2, alternative="greater")
```

# Two Sample t-tests

Two-sample t-test will likely be the most frequent t-test that you see in research or use in your own work.
A two-sample t-test asks if the mean of one group is different from another group.
The null hypothesis is that the mean of both groups is equal; stated differently: the difference in means is 0.

Here, we are going to ask if men and women have the same perception of public corruption in these three countries.


Now we can run our t-test. For the ```t.test``` function, we first give the outcome variable (corruption index), follow by ```~``` which is the equation format in **R**, and the last part of the equations is the grouping variable. 
Commonly `t.test` is done assuming the 2 groups have equal variances, but is not necessary as there is a correction to run the analysis without assuming this, we specify this with the argument ```var.equal=F```.

```{r}
#two-sample t-test 
two.sample <- t.test(dat$corrupt ~ dat$Sex, var.equal=F)
two.sample
```

Again, **R** gives us a lot of the same information as it did last time.
We can immediately see, however, that `p-value = 0.04992` and know that we can reject the null hypothesis that the two groups have equal means with respect to the corruption index we created. As long as we use an $\alpha = 0.05$, but if we had been more restrictive and use an $\alpha = 0.01$ we would have failed to reject the null hypothesis. We can also see the 95\% confidence interval for the mean differences between men and women.

We can see that **R** is using the model without assuming equality of variances as it reports the *Welch* correction. If you wanted to apply the model assuming equality of variances, can do it with the argument ```var.equal=T``` 

```{r}
#two-sample t-test 
t.test(dat$corrupt ~ dat$Sex, var.equal=T)
```

The default in **R** is ```var.qual=F```, we recommned to use is as your default. 

# Paired sample t-test

The final t-test we will consider is a paired sample t-test.
This is perhaps the most complex form of t-test, but it is useful when you are dealing with data that come in pairs.
We are still looking at the difference in means between two groups, but now each observation in one group has a counterpart (or "buddy") in the other group.

We will transform our data set in order to show an example of this.
First we will create a "ID".
The following code is creating a unique number that is associated with exactly 2 respondents (i.e. we are treating respondents one and two as if they were the same person before and after an intervention, respondents three and four as the same person, etc.) The respondents are not actually the same person in this data set, but we will use this as an example.
We will also create a time variable that we can use to tell which data is the "pre" and which is the "post".

```{r}
#Create an example ID with a time 1 and time 2 for each individual.
datp <- na.omit(dat)
datp <- datp[1:1300,]
datp$id <-rep(seq(1:(nrow(datp)/2)), each = 2)
datp$time <- rep(1:2,(nrow(datp)/2))
#keep only ID, time, and and corruption index, excluding missing values 
datp <- datp[,c("id","time","corrupt")]
# Show first few rows of new data
head(datp)
```

Now we can run the paired sample t-test.
Our two groups are time, and we need to tell **R** that the sample is paired with the `paired = TRUE` statement.
Paired sample t-tests do not assume equality of variance, so we do not need to include any variance parameter.

```{r}
pair.sample <- t.test(datp$corrupt ~ datp$time, paired = TRUE)
pair.sample
```

Again, **R** provides the same kind of details as in our previous two examples.
The first row of data shows us that `p-value = 0.5999`, meaning that we fail to reject the null hypothesis that both time points have equal means.
We should expect this, since this is a constructed example and there is no reason why there would be a difference between two responses in this case.

**Important Note of Caution** The data you use needs to be ordered by individual.
Look at the data table above and note how the rows of data are in order first by individual, and then by time.
Use the code below to see what happens if your data is not properly ordered.

```{r}
dat.1<- datp[order(datp$corrupt),]
t.test(dat.1$corrupt ~ dat.1$time, paired = TRUE)
```

# Effect Sizes

Let's reconsider our two sample t-test results.

```{r}
two.sample
```


As we saw, we reject the null hypothesis between men and women (with $\alpha = 0.05$).
But, even a very small difference can be statistically significant with a large enough sample size.
Is a difference of **0.044** meaningful?
Measures of effect sizes is one way to answer this question.

**Cohen's d** is an effect size statistic that is used with t-tests.
It is often reported with the t-test outcomes to give readers an idea of how big or small of a difference was found; it reports the difference in means in terms of standard deviations.
It is a useful to aid interpretation because it is not relent on the scale you of your variable of interest.
We can calculate Chohen's d using the `cohens_d()` function in the `effectsize` package.

Notice that the use of the function, is the same as with the ```t.test``` function. Just adding the argument ```pooled_sd=T``` 

```{r}
cohens_d(corrupt~Sex, data=dat, pooled_sd = T)
```

These results show that there is an .11 SD difference between men and women on the corruption perception index.
Whether or not Cohen's d is large or small will depend on the audience, but a 0.8 SD change is generally considered very large, and less than 0.2 SD change is generally considered small (Cohen, 1992).

For more details about the interpretation of the Cohen's d, please see  [https://rpsychologist.com/cohend/](https://rpsychologist.com/cohend/l)


# Graphing your results

It is often helpful to readers to not only read the results of your analysis, but to have visual to go along with it.
The package `ggpubr` is an easy package to plot t-tests.

```{r}
#install.packages("ggpubr", dependencies = TRUE)
library(ggpubr)
```

Let's continue with the two sample t-test results.
We have them saved in `two.sample`.

Box and whisker plots are often the best way to display differences in distributions that we look at with t-tests.
`ggboxplot()` can do this for us.

```{r}
plot <- 
  ggboxplot(
    dat, x = "Sex", y = "corrupt", 
    ylab = "Corruption Index", xlab = "Gender",
    subtitle = "Difference in Perception of Corruption by Gender") 
plot
```

The difference in this case isn't very large, so it's a bit hard to see.
Luckily, `ggpubr` is based on `ggplot2` so we can use all of the same `ggplot2` functions.
Lets add some of the t-test statistics to this graph to give additional detail.

```{r}
plot <- plot +
  labs(caption = paste(
  "Two sample t-test. p-value = ",
  round(two.sample$p.value,4)
  ))
plot
```

# References

Cohen, J. (1992). A Power Primer. Psychological Bulletin, 112(1), 155–159.
