---
title: "Linear Regression"
author: "Mauricio Garnier-Villarreal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 5
    number_sections: true
editor_options:
  chunk_output_type: inline
---

# Setup the R session

When we start working in R, we always need to setup our session. For this we need to set our working directory, in this case I am doing that for the folder that holds the downloaded [World Values Survey (WVS)](https://www.worldvaluessurvey.org/) ```SPSS``` data set

```{r, eval=F, error=F}
setwd("~path_to_your_file")
```


The next step for setting up our session will be to load the packages that we will be using

```{r, eval=T, message=F}
library(rio)
library(psych)
library(effectsize)
library(marginaleffects)
library(ggplot2)
library(parameters)
library(car)
library(patchwork)
library(summarytools)
```


# Import the data set

Here we will be importing the ```.sav``` WVS data set

```{r, eval=T}
dat <- import("WVS_Cross-National_Wave_7_sav_v2_0.sav")
dim(dat)
```

Here we are calling our data set **dat** and asking to see the dimension of it. We see that the data set has 76897 subjects, and 548 columns. 

## Prepare the data set

In cases with large data sets like this we might want to select a subset of variables that we want to work with. Since it is not easy to see 548 variables. 

```{r, eval=T}
vars <- c("Q260","Q262", "Y001", "SACSECVAL", "Q112", "Q113", "Q114", "Q115", "Q116", "Q117", "Q118", "Q119", "Q120", "Q65", "Q69", "Q71", "Q72", "Q73")
dat2 <- dat[,vars]
dim(dat2)
head(dat2)
```

Here we are first creating a vector with the variable names for the ones I want to keep. You can see all variable names for the full data set as well:

```{r, eval=F}
colnames(dat)
```

After identifying which variables we will work with, we create a new data set **dat2** with only these 17 variables, and make sure we did it correctly by looking at the the dimension of the data **dim(dat2)**. We also look at the first 6 rows: **head(dat2)**. These are quick checks that we have created the new data correctly. 

The variables we will use here are:

* Q260: sex, 1 = Male, 2 = Female
* Q262: age in years
* Y001: post-materialism index
* SACSECVAL: secular values
* Q112-Q120: Corruption Perception Index
* Q65-Q73: Lack of Confidence in the government

### Create composite scores

We will be using the composite scores for *Corruption Perception Index* and *Lack of Confidence in the government* instead of their single items. So, we first need to compute them, we will use the mean across all items for each composite

```{r, eval=T}
dat2$Corrup <- rowMeans(dat2[,c("Q112", "Q113", "Q114", "Q115", "Q116", "Q117", "Q118", "Q119", "Q120")], na.rm=T)
dat2$LCGov <- rowMeans(dat2[,c("Q65", "Q69", "Q71", "Q72", "Q73")], na.rm=T)
head(dat2)
```
With the ```rowmeans()``` we compute the mean across the specified variables, for each subject. Remember to include the ```na.rm=T``` argument, so the missing values are properly ignored.   

### Transform categorical variables to ```factor()```

when we want to use categorical variables as predictors, it is recommended to transformed them as ```factor``` type in ```R```. In this case we will use the *Sex* variable ```Q260```, which is coded as 1 and 2 right now. 

With the ```factor()``` function we an transform it, and change the numbers to meaningful labels. 

```{r}
dat2$Sex <- factor(dat2$Q260, 
                       levels = 1:2,
                       labels = c("Male","Female") )
```


### Set variables for analysis

Now, we will select only the variables of interest in a separate data set. 

```{r}
dat2 <- na.omit(dat2[,c("Sex", "Q262", "Y001", "SACSECVAL", "Corrup", "LCGov")])
head(dat2)
dim(dat2)
```
The new ```dat2``` data set only include the 6 continuous variables of interest, and 1 binary variable. With the ```na.omit()``` function we are excluding all cases with some missing values.


# Simple Linear regression

With linear regression we need to define an outcome (dependent) and predictor(s) (independent) variables. This way specifying directional relations. For this example, we will use *Perception of corruption* as the outcome, and *Lack of confidence in the government* as predictor. 

```{r}
mod1 <- lm(Corrup ~ LCGov, data=dat2)
```

Here we named the result of the regression ```mod1```, we use the function ```lm()``` which stands for *linear model*. This function requires 2 arguments, first the *formula*, which is a general formula form used in a lot of ```R```packages. The formula general form is ``Òutcome ~ Predictor```. And the second argument is the data set. 

Notice that when we run the regression, we don't see any results. ```R``` will only show you what you ask. The default results are shown by the ```summary()``` function. If you ask for the ```summary()``` of the regression model, you can see the regression table, overall model test (*F-test*), and $R^2$. 

```{r}
summary(mod1)
```


We can also ask for the confidence intervals with 

```{r}
confint(mod1, level = 0.95)
```

For this we need to specify the regression model, and confidence level. This example will estimate the 95% Confidence Interval for all the parameters in the model. 

We can also use the package ```parameters``` to create a summary table with the coefficients, standard errors, and confidence intervals

```{r}
parameters(mod1, digits=3)
```


## Standardize solution

It is common to look at the standardize regression slopes, we can estimate this with the ```standardize_parameters()``` function from the ```parameters```package

```{r}
standardize_parameters(mod1)
```

The ```standardize_parameters()``` function only requires the previous regression model, and it reports the CI automatically. It estimates the fully standardize regression. It is worth to note that when you have only 1 predictor, the standardize regression slope is equivalent to the *Pearson* correlation coefficient. 

```{r}
corr.test(dat2[,c("Corrup","LCGov")], method="pearson")
```



## Assumptions

The linear model assumptions can not be evaluated on the data, but only once the model has been estimated. In ```R``` the most practical way to do this is with the default plots for ```lm()``` 

```{r, eval=T}
plot(mod1)
```

Here we see various diagnostic plots. The **first** one presents the fitted values (predicted) against the model residuals (difference between predicted and observed values). With this plot we can evaluate linearity, and homogeneity of variance. If the *loess* (red) line is flat, mostly on top of the *expected* (grey dotted) line, this indicates that the linearity assumption is not violated. The linearity assumption would be violated when the *loess* line presents a curve that meaningfully deviates from the *expected* line. We can evaluate the homogeneity of variance, by looking at the spread of the points across the *x-axis*. If the points seems to have similar spread would indicate no violation of the homogeneity of variance assumption. When the homogeneity of variance assumption is violated, it presents clear differences in the spread of the points across the *x-axis*.

The **second** plot presents the *QQ* plot for the standardized residuals. This plots the theoretical quantiles (*x-axis*) against the standardized residuals (*y-axis*). With this plot we can evaluate the assumption of normality of the residuals. If the residuals are perfectly normal, the residual points would follow the theoretical diagonal line. Deviations from the theoretical line would indicate violations of the residual normality assumption. In this case we can see deviations from the theoretical line at the low and high ends.

The **third** plot is similar to the first one, with the fitted values on the *x-axis*, but presenting the squared standardized residuals on the *y-axis*. With the similarities with the first plot, we can evaluate the same assumptions (linearity and homogeneity of variances), and follow the same guidelines to review them. In this case we see no meaningful deviations from either assumption.

Lastly, the **fourth** plot presents the leverage against the residuals. With this we can evaluate the presence of outliers. More specifically with Cook's distance, when outliers are present they will be shown out of a red-dotted line indicating that they are outliers according to the Cook's distance. 

It is worth to remember that the linear models are robust to violations of the normality, homogeneity of variance, and outliers. But it is not robust to violations of linearity.


## Effect size

From the overall model we will use the $R^2$, and $R^2_{adj}$ as measures of model predictive accuracy. And $\eta^2$, $\omega^2$ as measure of predictor relative relevance. For simple regression (with only one predictor) overall model and predictor relevance would be equal, as the single indicator is the whole model

We can see the overall model ones from the ```summary()``` output

```{r}
summary(mod1)
```

Or you can only extract the $R^2$ by calling only that object of the summary with the ```$``` operator

```{r}
summary(mod1)$r.squared
```

In this case we see that $R^2 = 0.1079$, meaning that Lack of confidence in the government explains 10.8% of the variance in Perception of Corruption. 

```{r}
eta_squared(Anova(mod1, type=2), partial=F)
omega_squared(Anova(mod1, type=2), partial=F)
```

We can see that $R^2$, $R^2_{adj}$, $\eta^2$, and $\omega^2$ presents the same estimate up to the third decimal. Later we will show the interpretation of these different metrics when multiple indicators are present.


## Slope plots

We also want to plot the regression line, as the predicted scores next to the observed data. We can do this with the ````marginaleffects``` package. 

```{r}
plot_predictions(mod1, condition="LCGov")
```

The ```plot_predictions()``` function takes the ````lm()``` model and the predictor to plot in the argument ```condition```, and will plot the regression line with the confidence intervals in grey (we can not see the CI bands because of the large sample they are very small) on top the observed points.

with a single predictor, this is equivalent to the observed scatter-plot. when we add multiple predictors this change to conditional regression lines. 

## Interpretation of the results

Here we will present how to interpret the model results. With a single indicator we get all the information we need from the ```summary()``` or ```parameters``` function. 

```{r}
summary(mod1)
parameters(mod1, digits=3)
```

* From the overall model test, we reject the null hypothesis of the model being as good as the sample mean at predicting the outcome. Also could be seen as rejecting the null hypothesis of $R^2 = 0$. $F(1, 71631) = 8665, p < .001$. 
* The model explains 10.8% of the variance in Perception of corruption ($R^2 = 0.1079$).
* When the Lack of confidence in the government is equal to 0, the expected score in Perception of corruption is 2.55, and we reject the null hypothesis of this intercept being equal to 0, $b_0 = 2.55, SE = 0.01, p < .001$.
* As Lack of confidence in the government increases by 1, the perception in corruption increases by 0.366 points, and we reject the null hypothesis of this slope being equal to 0, $b_1 = 0.369, SE = 0.004, p < .001$.
* From the standardize solution we see that as Lack of government confidence increases by 1 standard deviation, Perception of corruption increases by 0.33 standard deviations.

```{r}
standardize_parameters(mod1)
```


# Linear regression with a binary predictor

We can run a regression where the predictor is a binary/categorical variable. With a categorical variable with only 2 categories (binary), this becomes equivalent to independent sample *t-test*. We can specify this just by setting the categorical variable as the predictor in the ```R``` formula, like this

```{r}
mod2 <- lm(Corrup ~ Sex, data = dat2)
summary(mod2)
parameters(mod2, digits=3)
```

We see the way to implement is the same as before, it is important to remember to set up the predictor as categorical with the ```factor()``` function. We will present the difference in the interpretation in a following sections.

It is worth noting that since our predictor is categorical, we see that in the ```summary``` the slope name is ```SexFemale```. This indicates that it is the slope for the category ```Female``` and that the other category is the baseline group of comparison, ```Male``` in this case. 


## Post-hoc comparisons

With categorical predictors the slopes represent the mean difference between groups, we can estimate all possible pairwise comparisons with the ```avg_comparisons``` and specifying for which categorical variable we want pairwise differences. For this example it is exactly the same as the regression table because we only have 2 groups, but will look different when you have more than 2 groups

```{r}
avg_comparisons(mod2, 
                variables = list(Sex="pairwise") )
```


## Standardize solution

There are divided opinions about the worthiness of standardized slopes for categorical predictors. I particularly don't recommend it as it sets up the slope in an uninterpretable metric, and the measures of relative relevance such as $\eta^2$ are better to compare multiple predictors effects. 

```{r}
standardize_parameters(mod2)
```

This function doesnt standardize the categorical predictor, so the interpretation is ```Female``` group has an standardize mean difference of 0.03 lower than ```Male```, basically a *Cohen d*. 

As you can see, the ```standardize_parameters()``` doesnt standardize the slopes for categorical predictors. But... if you really want to get the standardize estimates, you can add the ```method="basic" ``` argument 

```{r}
standardize_parameters(mod2, method = "basic")
```

## Assumptions

```{r}
plot(mod2)
```

By evaluating the assumptions with the plots, we see that there is no sign of violating the assumptions of linearity, homogeneity of variances, or outliers. But, there are signs of violations of the assumption of normally distributed residuals, at the tails of the distributions.  



## Effect size

As we still have only one predictor, the $R^2$, $\eta^2$, and $\omega^2$ will present the same results. We find that ```Sex``` explains 0.025% of the outcome's variance. 

```{r}
eta_squared(Anova(mod2, type=2), partial=F)
omega_squared(Anova(mod2, type=2), partial=F)
```

## Slope plots

When we plot the slopes for a categorical predictor ```plot_predictions``` will present it like a box plot for each group. It will do this if the predictor is set as a ```factor``` type variable. 

```{r}
plot_predictions(mod2, condition="Sex")
```

The point is the mean for each group and the lines represent the CI.


## Interpretation of the results

We will get the interpretation from the ```summary()``` output

```{r}
summary(mod2)
parameters(mod2)
```


* From the overall model test, we reject the null hypothesis of the model being as good as the sample mean at predicting the outcome. Also could be seen as rejecting the null hypothesis of $R^2 = 0$. $F(1, 71631) = 18.19, p < .001$. 
* The model explains 0.025% of the variance in Perception of corruption ($R^2 = 0.00025$).
* When *Sex* is at its baseline category (*Male*), the expected score in Perception of corruption is 3.50, and we reject the null hypothesis of this intercept being equal to 0, $b_0 = 3.50, SE = 0.004, p < .001$. Or could also say that the mean Perception of corruption for *Male* is 3.50. 
* As *Sex* changes category to *Female*, the perception in corruption decreases by 0.026 points, and we reject the null hypothesis of this slope (mean change) being equal to 0, $b_1 = -0.026, SE = 0.006, p < .001$. Or we could say that the mean difference between *Male* and *Female* is 0.026. 

# Linear regression with a categorical predictor (more than 2 categories)

A binary predictor is most simple case of a categorical predictor. But in many cases you might have a predictor with more than 2 categories. Many part of the process looks the same, but we will highlight the differences in this example. 

For this example, we first need to categorize a continuous predictor. But in general we dont recommend doing this, we will do it for the example. Here we will categorize ```Age``` into age groups.  

With the ```table()``` function we see that ages range from 16 to 103 in our data set

```{r}

table(dat2$Q262)

```
Then we will use th ```cut()``` function to create groups in ranges of the age variable. We assign it into a new variable ```age_ran``` , and provide the continuous variable to the ```cut()``` function, lastly yoy add the ```breaks``` for the continuous variable. This will cut the variable into groups at the respective break points. 

With ```class()``` we see that this automatically creates a ```factor``` variable. And with the ```freq()``` function we can see the frequency table for each group and the age range. This we can check that the groups were created as we intended 

```{r}
dat2$age_ran <- cut(dat2$Q262, 
                    breaks = c(15, 18, 30, 50, 70, 103))

class(dat2$age_ran)

freq(dat2$age_ran)
```
Then we just add this new predictor in the ```lm()``` functions, just like before. We can ask for the regression table either with the ```summary``` or ```parameters``` functions. The regression table parameters show the mean difference between each group with the reference group, in this case the reference group is the ```15 to 18```

```{r}

mod_cat <- lm(Corrup ~ age_ran, data = dat2)
summary(mod_cat)

parameters(mod_cat)

```

This model is the same as an ANOVA, but presented with the regression parameters. You can also get the ANOVA table from 

```{r}
Anova(mod_cat, type=2)

```
And can also put that into the ```parameters``` function for a cleaner table

```{r}
parameters(Anova(mod_cat, type=2))
```


## Post-hoc comparisons

With categorical predictors the slopes represent the mean difference between groups, we can estimate all possible pairwise comparisons with the ```avg_comparisons``` and specifying for which categorical variable we want pairwise differences. For this example, as we have 5 groups, we get 10 total pair comparisons

```{r}
pw2 <- avg_comparisons(mod_cat, type= "response",
                       variables = list(age_ran="pairwise") )
pw2
```

We can do different types of post-hoc comparisons, for example if we specify ```variables = list(age_ran="sequential")``` we will get sequential comparisons, so the first group compared with the second, and then the second compare to the third, and so on

```{r}
pw3 <- avg_comparisons(mod_cat, type= "response",
                       variables = list(age_ran="sequential") )
pw3
```
One last example I will present here is the case of planned contrasts, this is the case were you want to test hypothesis "combining" some groups, for example I want to compare the 2 youngest age groups against the older ones. 

We will do these comparisons in 2 steps, first we will estimate the predicted means for each group with the ```avg_predictions()``` functions. This will give us aobject with the saved mean predictors for each group.

```{r}
avpred <- avg_predictions(mod_cat, 
                          variables = "age_ran")
avpred
```

Then we use th ```hypotheses()``` function to build customize hypotheses. We build this hypothesis with contrasts weights, in this example we are combining the first 2 groups, and the last 3 groups. The contrasts weights have some basic rules:
- You specify which groups go together with the sign of the weights, the ones with a ```-``` weight go together, and positive ones go together
- The weights should sume up to 1 in each combined group. Thats why we use fractions in this example

```{r}
hypotheses(avpred, 
           hypothesis = c(-1/2, -1/2, 1/3, 1/3, 1/3))
```

Then we can also test multiple customize contrast at the same time. Save all the desired contrasts into a matrix, where each column is a contrast, and each row is a group. Here we use the function ```cbind()``` to create this matrix, this function is column-bind, so each vector will be a column in the new matrix

In this example, the first contrast is the same as before. The second one combines the first 3 groups against the last 2. And the last one ignores the first gorup (with a 0), and combines the second and third group, against the fourth and fifth ones. 

```{r}
conts <- cbind(c(-1/2, -1/2, 1/3, 1/3, 1/3),
               c(-1/3, -1/3, -1/3, 1/2, 1/2),
               c(0, -1/2, -1/2, 1/2, 1/2))
conts

hypotheses(avpred, hypothesis = conts, multcomp = "fdr")
```

One last argument that we add here is the ```multcomp = "fdr" ``` which adjusts the ```p-value``` for multiple comparisons. In this case we use the false discovery rate adjustment. 


## Standardize solution

There are divided opinions about the worthiness of standardized slopes for categorical predictors. I particularly don't recommend it as it sets up the slope in an uninterpretable metric, and the measures of relative relevance such as $\eta^2$ are better to compare multiple predictors effects. 

```{r}
standardize_parameters(mod_cat)
```

This function doesnt standardize the categorical predictor, so the interpretation is ```age ran [>70-103] ``` group has an standardize mean difference of 0.35 lower than the reference group, basically a *Cohen d*. 

As you can see, the ```standardize_parameters()``` doesnt standardize the slopes for categorical predictors. But... if you really want to get the standardize estimates, you can add the ```method="basic" ``` argument 

```{r}
standardize_parameters(mod_cat, method = "basic")
```

## Assumptions

```{r}
plot(mod_cat)
```

By evaluating the assumptions with the plots, we see that there is no sign of violating the assumptions of linearity, homogeneity of variances, or outliers. But, there are signs of violations of the assumption of normally distributed residuals, at the tails of the distributions.  



## Effect size

As we still have only one predictor, the $R^2$, $\eta^2$, and $\omega^2$ will present the same results. We find that ```Age ranges``` explains 0.95% of the outcome's variance. 

```{r}
summary(mod_cat)$r.squared*100
eta_squared(Anova(mod_cat, type=2), partial=F)
omega_squared(Anova(mod_cat, type=2), partial=F)
```

## Slope plots

We can also plot the "slopes" for categorical predictors, in this case the model predictions will be predicted means and confidence intervals fo each group. Here we can see that *Perception of corruption* is very similar between the first age groups, and then it goes down. 

```{r}
plot_predictions(mod_cat, condition = "age_ran")
```




# Multiple Linear regression

An advantage of the General Linear Models, is the flexibility that they have to include multiple predictors, and they can be any combination of continuous and/or categorical. 

In this following example we are adding 3 continuous predictors, and 2 categorical:

* Sex: Male, Female
* age_ran: age groups
* Y001: post-materialism index
* SACSECVAL: secular values
* LCGov: Lack of Confidence in the government

We see that we can add multiple predictors by adding to the original formula, such as ```Outcome ~ Predictor1 + Predictor2 + Predictor3 + ...``` . This way you can extend it with as many predictors as your data, model, and theory allows it. 

```{r}
mod3 <- lm(Corrup ~ Sex + age_ran + Y001 + SACSECVAL + LCGov, data=dat2)
summary(mod3)
confint(mod3, level=0.95)
```

Or the simpler table with ```parameters``` 

```{r}
parameters(mod3, digits=3)
```

When having a larger number of predictors, and combination of continuous and categorical ones, it is useful to get the ```Anova()``` table. specially because the NHST for the categorical predictor in the ```parameters``` table is for the mean comparison between group2 against group1. While we are also interested in the overall NHST of adding that predictor. 

```{r}
parameters(Anova(mod3, type=2))
```
The ```Anova``` table provides the NHST test for the null hypothesis of each predictor having no relation to the outcome variable, keeping all other predictors constant. For the continuous predictors this will be exactly the same as the one from the ```parameters``` table. 


## Model comparison

Before we had done two models, the first one with Just *Lack of confidence* as predictor, and the second with just *Sex* as predictor. Now we have added a total of five predictors. 

Now we can also test the hypothesis of two models being equivalent. This requires one model to be a subset of the second one. In this case we will compare ```mod1``` with ```mod3```. In this way ```mod1``` is a subset model from ```mod3```, as all the predictors of ```mod1``` are included in ```mod3```, with additional predictors. 

We can compare the two models with the ``ànova()``` function, this is a general method in ```R``` which identifies the the type of models that are being compared, and uses the correct method for the comparison. In this case does an *F-test*

```{r}
anova(mod1, mod3)
```

Here the degrees of freedom (*df*) are the difference in number of estimated parameters. In this case, we reject the null hypothesis of that both models are equally good at predicting *Perception of corruption*. Would lead us to choose the larger model. 

## Post-hoc comparisons

For the categorical predictor we can ask for the conditional pairwise differences with the ```avg_comparisons``` function. These would be the post-hoc comparisons, keeping all other predictors constant. For this example we are only doing the pairwise comparison, but you can also do the sequential or planned contrasts as in the previous example

```{r}
pw4 <- avg_comparisons(mod3, 
                variables = list(age_ran="pairwise") )
pw4
```


## Standardize solution

Once we have decided to continue with ```mod3```, we can also look at the standardized slopes. Remember that this function *does not* standardized the slopes for categorical predictors by default

```{r}
standardize_parameters(mod3)
```

If we use these standardized slopes to rank the predictors effects, we would conclude that *lack of confidence* if the one with the larger effect, and *Sex* is the with with the smallest effect. Remember that for *age_ran* each of these coefficients are a pairwise cohen d against the reference group. Remember that with this we can only rank the effects, but cannot test if a predictor has an effect *significantly* larger than another. 

## Assumptions

We will again use the diagnostic plots to evaluate for violations of the model assumptions

```{r}
plot(mod3)
```

From the first, and third plots we see that there is a slight deviation from the red *loess* line with the *expected* grey dotted line. Indicating there might be small deviation from the linearity assumption at the tails. This could be added into the model by adding the quadratic effect of a predictor for example. But this still looks like small deviation that would not lead me to consider it a meaningful violation. 

The other plots lead to no indication of violations of the homogeneity of variance assumption, or outliers. And as the other models, we see the assumption of normally distributed residuals violated in the tails of the distribution. 


## Effect size

For the overall model effect size we can look at the $R^2$, and $R^2_{adj}$ in the summary output

```{r}
summary(mod3)
```

Here we see that the overall model explains 17.5% of the variance in *Perception of corruption*. But this only presents the overall model effect, it does not account for the predictive effect of each predictor.

We can identiy each predictor's predictive effect with the full $\eta^2$ and $\omega^2$. Notice to estimate the *full* version of these metrics we need to specify ```partial=F```, so it does not estimate the partial version. 

These can be interpreted as the **proportion of variance in the outcome that es explained uniquely by each predictor**

When we have more than 1 predictor we need to change type of sum of squares used to calculate the ffect sizes, We can do that with ```Anova()``` function from the ```car``` package. this way we change it to ```type=2``` sum of squares. 

Notice that I am putting all this call insie the ```print()``` function so I can define the number of decimals to present (```digits=4```)
```{r}
print(eta_squared(Anova(mod3,type=2), partial=F), digits=4)
print(omega_squared(Anova(mod3,type=2), partial=F), digits=4)
```


We can also estimate the partial version of these metrics $\eta^2_p$ and $\omega^2_p$. Which are interpreted as the **proportion of variance in the outcome that uniquely explained by each predictor, that is not explained by any other predictor**. 

```{r}
eta_squared(Anova(mod3,type=2), partial=T)
omega_squared(Anova(mod3,type=2), partial=T)
```

Here we recommend the use of the *full* version of these metrics as it presents a more direct interpretation, but the partial versions might be commonly found in publications as they are default in other software.

$\eta^2$ can bee seen as the equivalent to the $R^2$, and $\omega^2$ as the equivalent to $R^2_{adj}$. But for the individual predictor's effects

## Conditional Slope plots

For plotting the regression slopes, when we had a single predictor the regression slope was equivalent to the scatter-plot. Now that we have multiple predictors, the plots present the regression line when all the other predictors are held constant, the conditional slopes. We can plot these with the ```plot_predictions()``` function, and specify which predictor to plot with the ```condition``` argument

```{r}
plot_predictions(mod3, condition = "Sex")
plot_predictions(mod3, condition = "age_ran")
plot_predictions(mod3, condition = "Y001")
plot_predictions(mod3, condition = "SACSECVAL")
plot_predictions(mod3, condition = "LCGov")
```

If we want to export all the plots together we can use the functionality from the package ```patchwork```. So that when it is loaded (```library(patchwork)```) we can simple add ```+``` between plots and they will be presented together

```{r}
plot_predictions(mod3, condition = "Sex")+
  plot_predictions(mod3, condition = "age_ran")+
  plot_predictions(mod3, condition = "Y001")+
  plot_predictions(mod3, condition = "SACSECVAL")+
  plot_predictions(mod3, condition = "LCGov")
```


## Interpretation of the results

Now, there are a lot more pieces to add to the interpretation. Lets start with the overall model null hypothesis test, from the ```summary(mod3)``` 

* We reject the null hypothesis that the overall model is equivalent to the mean model at predicting *Perception of corruption*, $F(8, 71624) = 1899, p < .001$. With the overall model explaining 17.5% ($R^2 = 0.175$) of the variance in the outcome

Then we can continue to explain the effect of each predictor, by adding information from the ```parameters()```, ```standardize_parameters()```, and ```eta_squared()``` functions

* The expected average score of *Perception of corruption* for *Males*, *Age* between 15 and 18, and with 0 scores in all predictors is 2.91 ($b_0 = 2.91, SE = 0.02, p < .001$), and we reject the null hypothesis of this expected score being equal to 0.
* We find that *Female* have in average a *Perception of corruption* 0.04 points lower than *Male* ($b_1 = -0.035, SE = 0.005, p < .001$), while holding all other predictors constant. We reject the null hypothesis of this mean difference being equal to 0. We see that *Sex* uniquely explains 0.05% ($\eta^2 = 0.0005$) of the variance of the outcome. 
* Holding all other predictors constant, we reject the null hypothesis pair means being equal for all pairwise comparisons except for 2. For the ones that we fail to reject are *(18,30] - (15,18]* and *(30,50] - (15,18]*. Indicating that the mean differences are smaller at the first agre groups. And we reject the null hypothesis as age ranges increase. This trend has a slight parabolic shape (as seen in the plot). We see that *Age* uniquely explains 0.86% ($\eta^2 = 0.0086$) of the variance of the outcome. 
* As *Post-materialism* increases by 1 unit, *Perception of corruption* decreases by 0.065 points in average($b_3 = -0.065, SE = 0.002, p < .001$), while holding all other predictors constant. We reject the null hypothesis of this slope being equal to 0. We see that *Post-materialism* uniquely explains 0.85% ($\eta^2 = 0.008$) of the variance of the outcome. 
* As *Secular values* increases by 1 unit, *Perception of corruption* decreases by 1.07 points in average($b_4 = -1.07, SE = 0.017, p < .001$), while holding all other predictors constant. We reject the null hypothesis of this slope being equal to 0. We see that *Secular values* uniquely explains 4.38% ($\eta^2 = 0.04$) of the variance of the outcome. 
* As *Lack of Confidence in the Government* increases by 1 unit, *Perception of corruption* increases by 0.455 points in average($b_5 = 0.456, SE = 0.004, p < .001$), while holding all other predictors constant. We reject the null hypothesis of this slope being equal to 0. We see that *Lack of Confidence in the Government* uniquely explains 14.15% ($\eta^2 = 0.14$) of the variance of the outcome. 
* Based on the relative effect measure of $\eta^2$ we can conclude that *Lack of Confidence in the Government* presents the stronger predictive effect, while *Sex* presents the lowest predictive effect. After controlling for the other predictors.  


